---
title: "CP7_The Insight Collective"
author: "Mekdelawit Dejene"
date: "2026-04-06"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Part 1: Loading and Checking Data
# Data Cleaning Pipeline
# Credit: Lincoln Chankira
# GitHub: https://github.com/tangiraiC/ITEC621-practicum
# To reproduce cleaned data: run quarto::quarto_render("datacleaning.qmd") before this script

```{r}
# loading required packages
library(readr)
library(readxl)
library(dplyr)
library(tidyr)
library(stringr)
library(tidytext)
library(SnowballC)
library(stopwords)
```

```{r}
# loading the cleaned data
exp_clean       <- read.csv("data/processed/cleaned_500_experiences.csv",
                             stringsAsFactors = FALSE)
currexp_clean   <- read.csv("data/processed/cleaned_500_current_experiences.csv",
                             stringsAsFactors = FALSE)
skills_clean    <- read.csv("data/processed/cleaned_500_skills.csv",
                             stringsAsFactors = FALSE)
exp_5k_clean    <- read.csv("data/processed/cleaned_5k_experiences.csv",
                             stringsAsFactors = FALSE)
skills_5k_clean <- read.csv("data/processed/cleaned_5k_skills.csv",
                             stringsAsFactors = FALSE)
```

# Checking how many rows and columns each file has
```{r}
# resource: https://www.w3schools.com/r/r_data_frames.asp
file_summary <- cbind(
  File    = c("500 experiences", "500 current experiences", "500 skills",
              "5k experiences", "5k skills"),
  Rows    = c(nrow(exp_clean), nrow(currexp_clean), nrow(skills_clean),
              nrow(exp_5k_clean), nrow(skills_5k_clean)),
  Columns = c(ncol(exp_clean), ncol(currexp_clean), ncol(skills_clean),
              ncol(exp_5k_clean), ncol(skills_5k_clean))
)
print(file_summary)
```

# Checking column names and what the data looks like
```{r}
# the names of the columns
names(exp_clean)
names(exp_5k_clean)
names(skills_5k_clean)
```

```{r}
# to see the structure such as rows, data types and some values
str(exp_clean)
str(exp_5k_clean)
str(skills_5k_clean)
```

## Part 2: Data Quality and Validation

```{r}
# finding the total missing values
sum(is.na(exp_clean))
sum(is.na(exp_5k_clean))
sum(is.na(skills_5k_clean))
```

```{r}
# checking to see which column has a missing value
colSums(is.na(exp_clean))
colSums(is.na(exp_5k_clean))

# most missing values were related to job started or ended on and other columns
# like industry and company name. Our key columns such as linkedin url and
# job descriptions have no missing values.
```

```{r}
# checking for duplicated linkedin urls
sum(duplicated(exp_clean$linkedinUrl))
sum(duplicated(exp_5k_clean$linkedinUrl))

# no duplicates in linkedin
```

```{r}
# checking for any missing linkedin url links
sum(is.na(exp_clean$linkedinUrl))
sum(is.na(exp_5k_clean$linkedin_url))

# no missing urls
```

```{r}
# checking to see how many profiles have job desc in 500 data and 5k data
sum(!is.na(exp_clean$jobDescription) & exp_clean$jobDescription != "")
sum(!is.na(exp_5k_clean$current_job_description) &
      exp_5k_clean$current_job_description != "")

# finding the percentage of what is in the 5k data
sum(!is.na(exp_5k_clean$current_job_description) &
      exp_5k_clean$current_job_description != "") / nrow(exp_5k_clean) * 100

# we have 316/452 profiles with job description in 500 data (about 70%)
# and 2050/5865 in 5k data (about 35%)
# we will use the 500 dataset for text preprocessing since it has better coverage
```

## Part 3: Text Preprocessing
# Resource: Professor's Module 1 - Text Preprocessing for NLP (KSB-621)
# Following the tidytext approach from the instructional guide

```{r}
# prepare the job descriptions as a data frame
# this follows the professor's guide structure - one row per profile
job_text <- exp_clean |>
  select(linkedinUrl, jobDescription) |>
  filter(!is.na(jobDescription) & jobDescription != "") |>
  mutate(line_id = row_number())

cat("Profiles with job descriptions:", nrow(job_text), "\n")

# check what raw text looks like before cleaning
cat("\nBefore cleaning:\n")
cat(job_text$jobDescription[1], "\n")
```

```{r}
# Step 1: Normalization
# Following professor's guide Step 1 - str_to_lower() and str_remove_all()
# str_to_lower() converts all text to lowercase
# str_remove_all() removes punctuation and special characters
# str_squish() removes extra whitespace

job_text <- job_text |>
  mutate(
    # lowercase everything
    text_lower = str_to_lower(jobDescription),
    # remove anything that is not a letter or space
    # also remove LinkedIn artifacts like \n and \t
    text_clean = str_remove_all(text_lower, "[^a-zA-Z\\s]"),
    text_clean = str_remove_all(text_clean, "\\\\n|\\\\t|\\\\r"),
    # remove extra whitespace
    text_clean = str_squish(text_clean)
  )

# check what cleaned text looks like
cat("After normalization:\n")
cat(job_text$text_clean[1], "\n")
```

```{r}
# Step 2: Tokenization
# Following professor's guide Step 2 - unnest_tokens()
# unnest_tokens() splits text into one word per row
# it also automatically lowercases and removes punctuation

job_words <- job_text |>
  unnest_tokens(output = word, input = text_clean, token = "words")

cat("Total tokens:", nrow(job_words), "\n")
cat("Average tokens per profile:", round(nrow(job_words) / nrow(job_text), 1), "\n")

# see what the top words look like before stopword removal
job_words |>
  count(word, sort = TRUE) |>
  head(20)
```

```{r}
# Step 3: Stopword Removal
# Following professor's guide Step 3 - get_stopwords() and anti_join()
# Stopwords are common words like "the" "and" "for" that carry no skill meaning
# get_stopwords() from the stopwords package uses the snowball list

stopwords_en <- get_stopwords(language = "en", source = "snowball")

job_words_clean <- job_words |>
  anti_join(stopwords_en, by = "word") |>
  # also remove very short tokens - usually artifacts
  filter(str_length(word) > 2) |>
  # remove tokens that are entirely numeric
  filter(!str_detect(word, "^[0-9]+$"))

cat("Tokens after stopword removal:", nrow(job_words_clean), "\n")

# top words after stopword removal
job_words_clean |>
  count(word, sort = TRUE) |>
  head(20)
```

```{r}
# Step 4: Stemming
# Following professor's guide Step 4 - wordStem() from SnowballC
# Stemming reduces words to their root form
# e.g. "managing" "manager" "management" all become "manag"
# This helps the model recognize the same skill written different ways

job_words_stemmed <- job_words_clean |>
  mutate(word_stem = wordStem(word, language = "english"))

# compare original words vs their stems
job_words_stemmed |>
  filter(word != word_stem) |>
  distinct(word, word_stem) |>
  head(20)

cat("Unique words before stemming:", n_distinct(job_words_clean$word), "\n")
cat("Unique stems after stemming: ", n_distinct(job_words_stemmed$word_stem), "\n")
```

```{r}
# Step 5: Bigrams (two-word phrases)
# Following professor's guide Step 5 - unnest_tokens() with token = "ngrams"
# Skills are often two words e.g. "project management" "data analysis"
# Single word tokens miss these - bigrams captures two word pairs

job_bigrams <- job_text |>
  unnest_tokens(output = bigram, input = text_clean,
                token = "ngrams", n = 2) |>
  filter(!is.na(bigram))

# remove bigrams where either word is a stopword
job_bigrams_clean <- job_bigrams |>
  separate(bigram, into = c("word1", "word2"), sep = " ") |>
  anti_join(stopwords_en, by = c("word1" = "word")) |>
  anti_join(stopwords_en, by = c("word2" = "word")) |>
  unite(bigram, word1, word2, sep = " ")

# top bigrams - these are likely to contain real skills
job_bigrams_clean |>
  count(bigram, sort = TRUE) |>
  head(20)
```

```{r}
# Step 6: Complete Pipeline Summary
# Following professor's guide Step 6 - putting it all together
# This is the final clean version combining all steps

processed_text <- job_text |>
  unnest_tokens(output = word, input = text_clean, token = "words") |>
  anti_join(get_stopwords(source = "snowball"), by = "word") |>
  filter(str_length(word) > 2) |>
  filter(!str_detect(word, "^[0-9]+$")) |>
  mutate(word_stem = wordStem(word, language = "english")) |>
  count(linkedinUrl, word_stem, sort = TRUE) |>
  rename(word = word_stem)

cat("Final processed text dimensions:", nrow(processed_text), "rows\n")
glimpse(processed_text)
```

#Building a data dictionary 
```{r}
library(tibble)
library(dplyr)

data_dictionary <- tibble(
  label = c(
    "leadership_management",
    "data_analysis",
    "business_operations",
    "customer_relations",
    "communication_skills",
    "strategic_planning"
  ),
  
  description = c(
    "Leading teams, managing people, and overseeing projects",
    "Working with data, analysis, and tools like Excel",
    "General business tasks, operations, and office tools",
    "Customer interaction, support, and service roles",
    "Communication, presentations, and speaking skills",
    "Planning, strategy, and organizational direction"
  ),
  
  keywords = c(
    "leadership, management, project management",
    "data analysis, excel, microsoft excel",
    "microsoft office, operations",
    "customer service, client, customer",
    "communication, public speaking, presentation",
    "strategic planning, strategy"
  )
)

print(data_dictionary)

write.csv(data_dictionary, "data_dictionary.csv", row.names = FALSE)
```

#converting keywords into patterns for matching 
```{r}
library(stringr)

data_dictionary <- data_dictionary %>%
  mutate(pattern = paste0("\\b(", keywords, ")\\b"))
```


#Apply dictionary labels 
```{r}
library(stringr)
library(purrr)
library(dplyr)
library(tidytext)

jobcleaned_text <- job_text$text_clean

text_df <- data.frame(
  document = seq_along(jobcleaned_text),
  text = sapply(jobcleaned_text, as.character),
  stringsAsFactors = FALSE
)

tokens <- text_df %>%
  unnest_tokens(word, text)

data(stop_words)

tokens_cleaned <- tokens %>%
  anti_join(stop_words, by = "word")

label_documents <- function(text_vector, dict) {
  
  map_dfr(seq_along(text_vector), function(i) {
    txt <- text_vector[i]
    
    matched_labels <- map_lgl(dict$keywords, function(kws) {
      any(str_detect(txt, regex(paste(kws, collapse = "|"), ignore_case = TRUE)))
    })
    
    tibble(
      document = i,
      label = dict$label[matched_labels]
    )
  })
}
cleaned_docs <- tokens_cleaned %>%
  group_by(document) %>%
  summarise(text = paste(word, collapse = " "), .groups = "drop")

doc_labels <- label_documents(cleaned_docs$text, data_dictionary)

head(doc_labels)
```

#Labeling and running clean docs
```{r}
library(purrr)

label_documents <- function(text_vector, dict) {
  
  map_dfr(seq_along(text_vector), function(i) {
    txt <- text_vector[i]
    
    matches <- str_detect(txt, regex(dict$pattern, ignore_case = TRUE))
    
    tibble(
      document = i,
      label = dict$label[matches]
    )
  })
}

doc_labels <- label_documents(cleaned_docs$text, data_dictionary)

head(doc_labels)

print(data_dictionary)
```

#Ground truth dataset creation 
```{r}
ground_truth <- cleaned_docs %>%
  mutate(true_label = NA)
```

#Set dictionary 
```{r}
library(jsonlite)

write_json(data_dictionary, "data_dictionary.json", pretty = TRUE)
```
