---
title: "CP7_The Insight Collective"
author: "Mekdelawit Dejene, Hanna Weinstein, Ayman Ibenbrahim, Lincoln Chanakira"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
options(scipen = 999)
set.seed(621)
```

## Executive Summary

This document prepares the Kogod Career Development LinkedIn profile data for Question B: Skills Extraction. The workflow loads the cleaned 5k experience dataset, checks data quality, combines relevant profile text fields, preprocesses the text, creates TF-IDF features, generates weakly supervised label columns from the self-reported `Skills` field, and exports train, validation, and test files for CP8 modeling.

## Libraries

```{r libraries}
library(readr)
library(dplyr)
library(tidyr)
library(stringr)
library(tidytext)
library(SnowballC)
library(stopwords)
library(ggplot2)
library(forcats)
library(scales)
library(Matrix)
```

## Part 1: Load and Inspect the Data

```{r load-data}
# Read from the project data folder so the document renders from the cp7 directory
exp_5k_clean <- read_csv("../data/processed/cleaned_5k_experiences.csv", show_col_types = FALSE)

dim(exp_5k_clean)
names(exp_5k_clean)
glimpse(exp_5k_clean)
```

```{r file-summary}
file_summary <- tibble(
  rows = nrow(exp_5k_clean),
  columns = ncol(exp_5k_clean),
  unique_profiles = n_distinct(exp_5k_clean$linkedin_url)
)

file_summary
```

## Part 2: Data Quality and Validation

```{r quality-checks}
missing_summary <- tibble(
  column = names(exp_5k_clean),
  missing_n = sapply(exp_5k_clean, function(x) sum(is.na(x))),
  missing_pct = round(sapply(exp_5k_clean, function(x) mean(is.na(x)) * 100), 1)
)

missing_summary
```

```{r integrity-checks}
cat("Duplicate linkedin_url values:", sum(duplicated(exp_5k_clean$linkedin_url)), "\n")
cat("Missing linkedin_url values:", sum(is.na(exp_5k_clean$linkedin_url)), "\n")
cat("Rows with non-missing Skills:", sum(!is.na(exp_5k_clean$Skills) & str_trim(exp_5k_clean$Skills) != ""), "\n")
```

## Part 3: Build the Modeling Text Field

```{r build-text}
text_cols <- c(
  "headline",
  "field_of_study",
  "current_job_title",
  "current_job_description",
  "current_company_name",
  "current_job_industry",
  "previous_job_title",
  "previous_job_description",
  "previous_company_name",
  "previous_job_Industry"
)

exp_model <- exp_5k_clean %>%
  mutate(across(all_of(text_cols), ~coalesce(., ""))) %>%
  mutate(
    combined_text = paste(
      headline,
      field_of_study,
      current_job_title,
      current_job_description,
      current_company_name,
      current_job_industry,
      previous_job_title,
      previous_job_description,
      previous_company_name,
      previous_job_Industry,
      sep = " "
    ),
    combined_text = str_squish(combined_text),
    has_text = combined_text != ""
  )

exp_model %>%
  summarise(
    profiles_total = n(),
    profiles_with_text = sum(has_text),
    pct_with_text = percent(mean(has_text))
  )
```

```{r filter-usable-rows}
exp_model <- exp_model %>%
  filter(has_text) %>%
  mutate(
    total_experience_count = coalesce(total_experience_count, 0),
    text_char_len = nchar(combined_text),
    text_word_len = str_count(combined_text, boundary("word")),
    current_desc_present = if_else(str_trim(current_job_description) == "", 0, 1),
    previous_desc_present = if_else(str_trim(previous_job_description) == "", 0, 1)
  )

exp_model %>%
  select(linkedin_url, combined_text, total_experience_count, text_char_len, text_word_len) %>%
  head(3)
```

## Part 4: Text Preprocessing

```{r normalize-text}
job_text <- exp_model %>%
  transmute(
    linkedin_url,
    combined_text,
    text_lower = str_to_lower(combined_text),
    text_clean = text_lower %>%
      str_replace_all("•|\\t|\\n|\\r", " ") %>%
      str_replace_all("http[s]?://\\S+", " ") %>%
      str_replace_all("[^a-zA-Z\\s]", " ") %>%
      str_squish()
  )

job_text %>% select(linkedin_url, text_clean) %>% head(2)
```

```{r tokenize}
stopwords_en <- get_stopwords(language = "en", source = "snowball")
custom_stopwords <- tibble(word = c("eg", "etc", "na", "llc", "inc", "amp"))
all_stopwords <- bind_rows(stopwords_en, custom_stopwords) %>% distinct(word)

job_words <- job_text %>%
  select(linkedin_url, text_clean) %>%
  unnest_tokens(output = word, input = text_clean, token = "words")

job_words_clean <- job_words %>%
  anti_join(all_stopwords, by = "word") %>%
  filter(str_length(word) > 2) %>%
  filter(!str_detect(word, "^[0-9]+$")) %>%
  mutate(word_stem = wordStem(word, language = "english"))

job_words_clean %>% count(word_stem, sort = TRUE) %>% head(20)
```

```{r processed-text}
processed_text <- job_words_clean %>%
  count(linkedin_url, word_stem, sort = TRUE) %>%
  rename(word = word_stem)

cat("Processed token rows:", nrow(processed_text), "\n")
cat("Vocabulary size:", n_distinct(processed_text$word), "\n")
```

## Part 5: Feature Engineering and Variable Creation

```{r tfidf-build}
# Reduce the vocabulary slightly for stability and memory management
term_stats <- processed_text %>%
  group_by(word) %>%
  summarise(
    doc_freq = n_distinct(linkedin_url),
    total_count = sum(n),
    .groups = "drop"
  ) %>%
  filter(doc_freq >= 5)

processed_text_small <- processed_text %>%
  inner_join(term_stats %>% select(word), by = "word")

word_counts <- processed_text_small %>%
  count(linkedin_url, word, wt = n, sort = TRUE)

tfidf <- word_counts %>%
  bind_tf_idf(term = word, document = linkedin_url, n = n)

feature_matrix <- tfidf %>%
  select(linkedin_url, word, tf_idf) %>%
  pivot_wider(names_from = word, values_from = tf_idf, values_fill = 0)

cat("Feature matrix dimensions:", nrow(feature_matrix), "rows x", ncol(feature_matrix), "columns\n")
```

```{r metadata-features}
metadata_features <- exp_model %>%
  select(
    linkedin_url,
    total_experience_count,
    text_char_len,
    text_word_len,
    current_desc_present,
    previous_desc_present
  ) %>%
  rename_with(~paste0("meta_", .x), -linkedin_url)

metadata_features %>% head(3)
```

## Part 6: Labeling and Ground Truth

```{r top-skills}
# Use the self-reported Skills field as weakly supervised ground truth.
skills_long <- exp_model %>%
  select(linkedin_url, Skills) %>%
  filter(!is.na(Skills), str_trim(Skills) != "") %>%
  separate_rows(Skills, sep = ",") %>%
  mutate(
    skill = str_squish(str_to_lower(Skills))
  ) %>%
  filter(skill != "") %>%
  distinct(linkedin_url, skill)

top_skills <- skills_long %>%
  count(skill, sort = TRUE) %>%
  slice_head(n = 10)

top_skills
```

```{r label-matrix}
# Sanitize skill names so they become valid, readable column names.
sanitize_skill_name <- function(x) {
  x %>%
    str_to_lower() %>%
    str_replace_all("[^a-z0-9]+", "_") %>%
    str_replace_all("_+", "_") %>%
    str_replace_all("^_|_$", "")
}

label_matrix <- skills_long %>%
  filter(skill %in% top_skills$skill) %>%
  mutate(label_name = paste0("label_", sanitize_skill_name(skill)), value = 1L) %>%
  select(linkedin_url, label_name, value) %>%
  pivot_wider(names_from = label_name, values_from = value, values_fill = 0)

label_matrix %>% head(3)
```

```{r assemble-model-dataset}
model_dataset <- feature_matrix %>%
  left_join(metadata_features, by = "linkedin_url") %>%
  left_join(label_matrix, by = "linkedin_url") %>%
  mutate(across(starts_with("label_"), ~coalesce(., 0L)))

model_dataset %>%
  summarise(
    rows = n(),
    columns = ncol(.),
    label_columns = sum(str_detect(names(.), "^label_"))
  )
```

```{r class-balance}
label_balance <- model_dataset %>%
  summarise(across(starts_with("label_"), ~mean(. == 1))) %>%
  pivot_longer(cols = everything(), names_to = "label", values_to = "positive_rate") %>%
  mutate(positive_rate = percent(positive_rate))

label_balance
```

## Part 7: Data Partitioning

```{r partitioning}
# We split once at the profile level and carry both feature columns and label columns forward.
set.seed(621)

n_profiles <- nrow(model_dataset)
train_size <- floor(0.70 * n_profiles)
val_size   <- floor(0.15 * n_profiles)
test_size  <- n_profiles - train_size - val_size

shuffled_idx <- sample(n_profiles)
train_idx <- shuffled_idx[1:train_size]
val_idx   <- shuffled_idx[(train_size + 1):(train_size + val_size)]
test_idx  <- shuffled_idx[(train_size + val_size + 1):n_profiles]

train_data <- model_dataset[train_idx, ]
validation_data <- model_dataset[val_idx, ]
test_data <- model_dataset[test_idx, ]

tibble(
  split = c("train", "validation", "test"),
  rows = c(nrow(train_data), nrow(validation_data), nrow(test_data)),
  pct = scales::percent(c(nrow(train_data), nrow(validation_data), nrow(test_data)) / nrow(model_dataset))
)
```

```{r save-outputs}
write_csv(train_data, "train_data.csv")
write_csv(validation_data, "validation_data.csv")
write_csv(test_data, "test_data.csv")
write_csv(top_skills, "top_skill_targets.csv")
write_csv(label_balance, "label_balance_summary.csv")
```

## Part 8: Notes for CP8

The train, validation, and test CSV files created above are the modeling inputs for CP8. Each exported file contains:

- `linkedin_url` as the profile identifier
- TF-IDF feature columns
- metadata columns prefixed with `meta_`
- label columns prefixed with `label_`

In CP8, the team can select one label column at a time, such as `label_data_analysis` or `label_project_management`, and compare multiple candidate models on the same held-out split.
