---
title: "Data Preprocessing - 500 Profile Dataset"
author: "Team"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(tidyverse)
library(stringr)
library(textclean)
library(stopwords)
library(here)

project_root <- normalizePath(file.path("..", "..", ".."))
processed_dir <- file.path(project_root, "data", "processed")

dir.create(processed_dir, recursive = TRUE, showWarnings = FALSE)
```

# Introduction

In this section, we will clean and normalize the profile text from the
500-profile dataset so it can be converted into machine-learning features
consistently.

## What This File Does

- Loads `model_ready_data.rds` from the preparation step.
- Cleans `profile_text` by removing URLs, emails, HTML, punctuation, and extra spaces.
- Creates `profile_text_clean`, `token_count`, and `char_count`.
- Removes rows with very little usable text.
- Creates `profile_text_nostop` by removing common English stopwords.
- Saves `clean_text_data.rds` and `clean_text_data.csv`.

# Load Prepared Data

```{r load-prepared-data}
df <- read_rds(file.path(processed_dir, "model_ready_data.rds"))

glimpse(df)
```

# Clean Profile Text

```{r clean-profile-text}
clean_text <- function(text) {
  text |>
    replace_non_ascii() |>
    str_to_lower() |>
    replace_url(replacement = " ") |>
    replace_email(replacement = " ") |>
    replace_html(replacement = " ") |>
    str_replace_all("[^a-z0-9\\s]", " ") |>
    str_replace_all("\\b\\d+\\b", " ") |>
    str_replace_all("\\s+", " ") |>
    str_trim()
}

stop_words <- stopwords("en")

df_clean <- df |>
  mutate(
    profile_text_clean = map_chr(profile_text, clean_text),
    token_count = str_count(profile_text_clean, "\\S+"),
    char_count = nchar(profile_text_clean)
  ) |>
  filter(!is.na(profile_text_clean), profile_text_clean != "", token_count > 2)

glimpse(df_clean)
summary(df_clean$token_count)
```

# Remove Stopwords

```{r remove-stopwords}
remove_stopwords_from_string <- function(text, stop_words) {
  tokens <- unlist(str_split(text, "\\s+"))
  tokens <- tokens[!tokens %in% stop_words]
  paste(tokens, collapse = " ")
}

df_clean <- df_clean |>
  mutate(
    profile_text_nostop = map_chr(
      profile_text_clean,
      remove_stopwords_from_string,
      stop_words = stop_words
    )
  )

df_clean |>
  select(linkedin_url, profile_text_clean, profile_text_nostop, token_count, char_count) |>
  head()
```

# Save Outputs

```{r save-clean-data}
clean_rds_path <- file.path(processed_dir, "clean_text_data.rds")
clean_csv_path <- file.path(processed_dir, "clean_text_data.csv")

write_rds(df_clean, clean_rds_path)
write_csv(df_clean |> select(-skills_list), clean_csv_path)

cat("Clean text RDS saved to:", clean_rds_path, "\n")
cat("Clean text CSV saved to:", clean_csv_path, "\n")
```
