---
title: "LinkedIn Data Cleaning Pipeline"
author: "Generated from Python Script"
date: today
format:
  html:
    toc: true
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
---

# Overview

This Quarto document reproduces the LinkedIn cleaning workflow in
`cleaningScripts/datacleaning.py`. It reads the raw 500-profile and
5K-profile datasets, cleans them, writes the processed CSV files, and
shows preview tables in the rendered report.

## Expected Outputs

The document writes these files to `data/processed/`:

- `cleaned_500_experiences.csv`
- `cleaned_500_current_experiences.csv`
- `cleaned_500_skills.csv`
- `cleaned_5k_experiences.csv`
- `cleaned_5k_skills.csv`

# Setup

## Libraries

```{r setup}
library(readr)
library(readxl)
library(dplyr)
library(tidyr)
library(stringr)
library(knitr)
```

## Project Paths

```{r paths}
resolve_project_path <- function(...) {
  candidates <- c(
    normalizePath(file.path("..", ...), winslash = "/", mustWork = FALSE),
    normalizePath(file.path(".", ...), winslash = "/", mustWork = FALSE)
  )

  existing_candidate <- candidates[file.exists(candidates)][1]
  if (!is.na(existing_candidate)) {
    return(existing_candidate)
  }

  candidates[[1]]
}

data_500_experience_full_path <- resolve_project_path(
  "data", "raw", "LinkedIn_Scrape_500_Experiences_Wide_Full_List.csv"
)
data_500_experience_current_path <- resolve_project_path(
  "data", "raw", "LinkedIn_Scrape_500_Experiences_Wide_Current_List.csv"
)
data_500_skills_path <- resolve_project_path(
  "data", "raw", "LinkedIn_Scrape_500_Skills_List.csv"
)
data_5k_experience_path <- resolve_project_path(
  "data", "raw", "LinkedIn_Scrape_full_5865.xlsx"
)
data_5k_skills_path <- resolve_project_path(
  "data", "raw", "LinkedIn_Scrape_full_5865_Skills_List.csv"
)
processed_dir <- resolve_project_path("data", "processed")

dir.create(processed_dir, recursive = TRUE, showWarnings = FALSE)
```

## Input Validation

```{r validate-inputs}
required_input_files <- c(
  data_500_experience_full_path,
  data_500_experience_current_path,
  data_500_skills_path,
  data_5k_experience_path,
  data_5k_skills_path
)

missing_input_files <- required_input_files[!file.exists(required_input_files)]

if (length(missing_input_files) > 0) {
  stop(
    paste(
      "Missing required input files:",
      paste(missing_input_files, collapse = ", ")
    )
  )
}

tibble::tibble(
  input_file = basename(required_input_files),
  location = required_input_files
) |>
  knitr::kable(caption = "Validated input files")
```

# Cleaning Logic

## Helper Functions

```{r helpers}
join_columns <- function(df, columns, output_column) {
  existing_columns <- columns[columns %in% colnames(df)]
  if (length(existing_columns) == 0) {
    return(df)
  }

  df[[output_column]] <- apply(df[, existing_columns, drop = FALSE], 1, function(row) {
    paste(na.omit(row), collapse = " ")
  })

  df <- df[, !colnames(df) %in% existing_columns]
  df
}

drop_prefixed_columns <- function(df, prefix, max_index) {
  columns_to_drop <- paste0(prefix, "_", 1:max_index)
  columns_to_drop <- columns_to_drop[columns_to_drop %in% colnames(df)]

  if (length(columns_to_drop) > 0) {
    df <- df[, !colnames(df) %in% columns_to_drop]
  }

  df
}

select_500_columns <- function(df, max_index = 27) {
  allowed_columns <- c("linkedinUrl", "jobDescription", "jobTitle", "companyName")
  allowed_columns <- c(
    allowed_columns,
    paste0("jobStartedOn_", 1:max_index),
    paste0("jobEndedOn_", 1:max_index)
  )
  allowed_columns <- allowed_columns[allowed_columns %in% colnames(df)]

  df[, allowed_columns, drop = FALSE]
}
```

## Experience Cleaning Function

```{r clean-full-experience}
clean_full_experience_df <- function(df, max_index = 27) {
  job_description_columns <- paste0("jobDescription_", 1:max_index)
  title_columns <- paste0("title_", 1:max_index)
  company_name_columns <- paste0("companyName_", 1:max_index)

  df <- join_columns(df, job_description_columns, "jobDescription")
  df <- join_columns(df, title_columns, "jobTitle")
  df <- join_columns(df, company_name_columns, "companyName")

  df <- drop_prefixed_columns(df, "companyId", max_index)
  df <- drop_prefixed_columns(df, "companySize", max_index)
  df <- drop_prefixed_columns(df, "companyIndustry", max_index)
  df <- drop_prefixed_columns(df, "companyWebsite", max_index)
  df <- drop_prefixed_columns(df, "employmentType", max_index)
  df <- drop_prefixed_columns(df, "jobLocation", max_index)
  df <- drop_prefixed_columns(df, "jobLocationCountry", max_index)
  df <- drop_prefixed_columns(df, "jobStillWorking", max_index)

  select_500_columns(df, max_index)
}
```

## Dataset-Specific Cleaning Functions

```{r dataset-cleaners}
clean_500_dataset <- function() {
  df_experiences <- read_csv(data_500_experience_full_path, show_col_types = FALSE)
  df_current_experiences <- read_csv(data_500_experience_current_path, show_col_types = FALSE)
  df_skills <- read_csv(data_500_skills_path, show_col_types = FALSE)

  list(
    experiences = clean_full_experience_df(df_experiences),
    current_experiences = clean_full_experience_df(df_current_experiences, max_index = 1),
    skills = df_skills
  )
}

clean_5k_dataset <- function() {
  df_experiences <- read_excel(data_5k_experience_path, sheet = "Scraped data")
  df_skills <- read_csv(data_5k_skills_path, show_col_types = FALSE)

  columns_to_drop <- c(
    "current_company_url",
    "Company_size",
    "last_scraped_date",
    "previous_job_start_date",
    "previous_job_location",
    "previous_job_country",
    "previous_company_size",
    "employment_type",
    "is_currently_employed",
    "current_job_country",
    "current_job_location",
    "current_job_start_date",
    "email",
    "full_name",
    "education_start_year",
    "education_end_year",
    "previous_job_end_date",
    "total_experience_years",
    "current_job_duration_years"
  )

  columns_to_drop <- columns_to_drop[columns_to_drop %in% colnames(df_experiences)]
  if (length(columns_to_drop) > 0) {
    df_experiences <- df_experiences[, !colnames(df_experiences) %in% columns_to_drop]
  }

  list(
    experiences = df_experiences,
    skills = df_skills
  )
}
```

## Save Functions

```{r save-functions}
save_cleaned_data <- function(data_list) {
  write_csv(data_list$experiences, file.path(processed_dir, "cleaned_500_experiences.csv"))
  write_csv(
    data_list$current_experiences,
    file.path(processed_dir, "cleaned_500_current_experiences.csv")
  )
  write_csv(data_list$skills, file.path(processed_dir, "cleaned_500_skills.csv"))
}

save_cleaned_5k_data <- function(data_list) {
  write_csv(data_list$experiences, file.path(processed_dir, "cleaned_5k_experiences.csv"))
  write_csv(data_list$skills, file.path(processed_dir, "cleaned_5k_skills.csv"))
}
```

# Execution

## Run The Pipeline

```{r main, cache=TRUE}
data_500 <- clean_500_dataset()
save_cleaned_data(data_500)

data_5k <- clean_5k_dataset()
save_cleaned_5k_data(data_5k)
```

## Output Summary

```{r output-summary}
output_summary <- tibble::tibble(
  file_name = c(
    "cleaned_500_experiences.csv",
    "cleaned_500_current_experiences.csv",
    "cleaned_500_skills.csv",
    "cleaned_5k_experiences.csv",
    "cleaned_5k_skills.csv"
  ),
  rows = c(
    nrow(data_500$experiences),
    nrow(data_500$current_experiences),
    nrow(data_500$skills),
    nrow(data_5k$experiences),
    nrow(data_5k$skills)
  ),
  columns = c(
    ncol(data_500$experiences),
    ncol(data_500$current_experiences),
    ncol(data_500$skills),
    ncol(data_5k$experiences),
    ncol(data_5k$skills)
  ),
  output_path = file.path(processed_dir, file_name)
)

knitr::kable(output_summary, caption = "Generated output files")
```

<!--
# Data Previews

## 500 Dataset

### Full Experience Preview

```{r preview-500-experiences}
knitr::kable(head(data_500$experiences), caption = "500 full experiences preview")
```

### Current Experience Preview

```{r preview-500-current}
knitr::kable(
  head(data_500$current_experiences),
  caption = "500 current experiences preview"
)
```

### Skills Preview

```{r preview-500-skills}
knitr::kable(head(data_500$skills), caption = "500 skills preview")
```

## 5K Dataset

### Experience Preview

```{r preview-5k-experiences}
knitr::kable(head(data_5k$experiences), caption = "5K experiences preview")
```

### Skills Preview

```{r preview-5k-skills}
knitr::kable(head(data_5k$skills), caption = "5K skills preview")
```
-->

# Summary

The Quarto document now runs the full cleaning workflow, validates the required
input files before processing, writes all five output CSVs to `data/processed/`,
and includes a structured summary plus preview tables in the rendered report.
