---
title: "LinkedIn Data Cleaning Pipeline"
author: "Generated from Python Script"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(readxl)
library(dplyr)
library(tidyr)
library(stringr)
```

## Overview

This RMarkdown document performs the same data cleaning as the Python script `datacleaning.py`. It processes LinkedIn scrape datasets for 500 and 5K profiles.

## File Paths

```{r paths}
# Define file paths
data_500_experience_full_path <- "./data/raw/LinkedIn_Scrape_500_Experiences_Wide_Full_List.csv"
data_500_experience_current_path <- "./data/raw/LinkedIn_Scrape_500_Experiences_Wide_Current_List.csv"
data_500_skills_path <- "./data/raw/LinkedIn_Scrape_500_Skills_List.csv"

data_5k_experience_path <- "./data/raw/LinkedIn_Scrape_full_5865.xlsx"
data_5k_skills_path <- "./data/raw/LinkedIn_Scrape_full_5865_Skills_List.csv"

processed_dir <- "./data/processed"

# Create processed directory if it doesn't exist
dir.create(processed_dir, recursive = TRUE, showWarnings = FALSE)
```

## Helper Functions

```{r helpers}
# Function to join columns into a single column
join_columns <- function(df, columns, output_column) {
  existing_columns <- columns[columns %in% colnames(df)]
  if (length(existing_columns) == 0) {
    return(df)
  }

  # Combine columns, removing NA values
  df[[output_column]] <- apply(df[, existing_columns, drop = FALSE], 1, function(row) {
    paste(na.omit(row), collapse = " ")
  })

  # Remove original columns
  df <- df[, !colnames(df) %in% existing_columns]
  return(df)
}

# Function to drop prefixed columns
drop_prefixed_columns <- function(df, prefix, max_index) {
  columns_to_drop <- paste0(prefix, "_", 1:max_index)
  columns_to_drop <- columns_to_drop[columns_to_drop %in% colnames(df)]
  if (length(columns_to_drop) > 0) {
    df <- df[, !colnames(df) %in% columns_to_drop]
  }
  return(df)
}

# Function to select 500 columns
select_500_columns <- function(df, max_index = 27) {
  allowed_columns <- c("linkedinUrl", "jobDescription", "jobTitle", "companyName")
  allowed_columns <- c(allowed_columns,
                       paste0("jobStartedOn_", 1:max_index),
                       paste0("jobEndedOn_", 1:max_index))
  allowed_columns <- allowed_columns[allowed_columns %in% colnames(df)]
  return(df[, allowed_columns, drop = FALSE])
}
```

## Clean Full Experience DataFrame

```{r clean_full_experience}
clean_full_experience_df <- function(df, max_index = 27) {
  df <- df  # Copy

  # Define column groups to consolidate
  job_description_columns <- paste0("jobDescription_", 1:max_index)
  title_columns <- paste0("title_", 1:max_index)
  company_name_columns <- paste0("companyName_", 1:max_index)

  # Consolidate wide columns
  df <- join_columns(df, job_description_columns, "jobDescription")
  df <- join_columns(df, title_columns, "jobTitle")
  df <- join_columns(df, company_name_columns, "companyName")

  # Remove unwanted prefixed columns
  df <- drop_prefixed_columns(df, "companyId", max_index)
  df <- drop_prefixed_columns(df, "companySize", max_index)
  df <- drop_prefixed_columns(df, "companyIndustry", max_index)
  df <- drop_prefixed_columns(df, "companyWebsite", max_index)
  df <- drop_prefixed_columns(df, "employmentType", max_index)
  df <- drop_prefixed_columns(df, "jobLocation", max_index)
  df <- drop_prefixed_columns(df, "jobLocationCountry", max_index)
  df <- drop_prefixed_columns(df, "jobStillWorking", max_index)

  # Select final columns
  df <- select_500_columns(df, max_index)
  return(df)
}
```

## Clean 500 Dataset

```{r clean_500}
clean_500_dataset <- function() {
  df_experiences <- read_csv(data_500_experience_full_path, show_col_types = FALSE)
  df_current_experiences <- read_csv(data_500_experience_current_path, show_col_types = FALSE)
  df_skills <- read_csv(data_500_skills_path, show_col_types = FALSE)

  df_experiences <- clean_full_experience_df(df_experiences)
  df_current_experiences <- clean_full_experience_df(df_current_experiences, max_index = 1)

  return(list(experiences = df_experiences,
              current_experiences = df_current_experiences,
              skills = df_skills))
}
```

## Clean 5K Dataset

```{r clean_5k}
clean_5k_dataset <- function() {
  df_experiences <- read_excel(data_5k_experience_path, sheet = "Scraped data")
  df_skills <- read_csv(data_5k_skills_path, show_col_types = FALSE)

  # Columns to remove
  columns_to_drop <- c(
    "current_company_url",
    "Company_size",
    "last_scraped_date",
    "previous_job_start_date",
    "previous_job_location",
    "previous_job_country",
    "previous_company_size",
    "employment_type",
    "is_currently_employed",
    "current_job_country",
    "current_job_location",
    "current_job_start_date",
    "email",
    "full_name",
    "education_start_year",
    "education_end_year",
    "previous_job_end_date",
    "total_experience_years",
    "current_job_duration_years"
  )

  # Drop columns that exist
  columns_to_drop <- columns_to_drop[columns_to_drop %in% colnames(df_experiences)]
  if (length(columns_to_drop) > 0) {
    df_experiences <- df_experiences[, !colnames(df_experiences) %in% columns_to_drop]
  }

  return(list(experiences = df_experiences, skills = df_skills))
}
```

## Save Functions

```{r save_functions}
save_cleaned_data <- function(data_list) {
  write_csv(data_list$experiences, file.path(processed_dir, "cleaned_500_experiences.csv"))
  write_csv(data_list$current_experiences, file.path(processed_dir, "cleaned_500_current_experiences.csv"))
  write_csv(data_list$skills, file.path(processed_dir, "cleaned_500_skills.csv"))
}

save_cleaned_5k_data <- function(data_list) {
  write_csv(data_list$experiences, file.path(processed_dir, "cleaned_5k_experiences.csv"))
  write_csv(data_list$skills, file.path(processed_dir, "cleaned_5k_skills.csv"))
}
```

## Display Functions

```{r display}
showheads <- function(data_list) {
  cat("Experiences DataFrame:\n")
  print(head(data_list$experiences))
  cat("\nCurrent Experiences DataFrame:\n")
  print(head(data_list$current_experiences))
  cat("\nSkills DataFrame:\n")
  print(head(data_list$skills))
}

showheads_5k <- function(data_list) {
  cat("5K Experiences DataFrame:\n")
  print(head(data_list$experiences))
  cat("\n5K Skills DataFrame:\n")
  print(head(data_list$skills))
}
```

## Main Execution

```{r main}
# Process 500 dataset
data_500 <- clean_500_dataset()
save_cleaned_data(data_500)
showheads(data_500)

# Process 5K dataset
data_5k <- clean_5k_dataset()
save_cleaned_5k_data(data_5k)
showheads_5k(data_5k)
```

## Summary

The cleaning pipeline has processed both datasets and saved the results to the `data/processed/` directory.