---
title: "Model Input Encoding - 500 Profile Dataset"
author: "Team"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(tidyverse)
library(rsample)
library(quanteda)
library(Matrix)
library(here)

project_root <- normalizePath(file.path("..", "..", ".."))
processed_dir <- file.path(project_root, "data", "processed")

dir.create(processed_dir, recursive = TRUE, showWarnings = FALSE)
```

# Introduction

In this section, we will split the 500-profile dataset into training,
validation, and test sets, build TF-IDF text features, and create multi-label
target matrices for the most frequent skills.

## What This File Does

- Loads `clean_text_data.rds` from preprocessing.
- Splits the data into training, validation, and test sets.
- Selects the top 30 skills from the training data only.
- Creates binary target matrices for those top skills.
- Converts cleaned profile text into TF-IDF feature matrices.
- Saves `X_train`, `X_valid`, `X_test`, `y_train`, `y_valid`, `y_test`, and `top_skills`.

# Load Clean Text Data

```{r load-clean-text-data}
df <- read_rds(file.path(processed_dir, "clean_text_data.rds"))

glimpse(df)
```

# Create Train, Validation, and Test Splits

```{r create-data-splits}
set.seed(621)

split_1 <- initial_split(df, prop = 0.70, strata = has_any_skill)
train_data <- training(split_1)
temp_data <- testing(split_1)

split_2 <- initial_split(temp_data, prop = 0.50, strata = has_any_skill)
valid_data <- training(split_2)
test_data <- testing(split_2)

nrow(train_data)
nrow(valid_data)
nrow(test_data)
```

# Choose Top Training Skills

```{r choose-top-skills}
TOP_N_SKILLS <- 30

train_skill_freq <- train_data |>
  select(linkedin_url, skills_list) |>
  unnest(skills_list) |>
  count(skills_list, sort = TRUE)

top_skills <- train_skill_freq |>
  slice_head(n = TOP_N_SKILLS) |>
  pull(skills_list)

top_skills
```

# Keep Top Skills in Each Split

```{r keep-top-skills}
keep_top_skills <- function(skill_vec, top_skills) {
  skill_vec[skill_vec %in% top_skills]
}

train_data <- train_data |>
  mutate(skills_top = map(skills_list, keep_top_skills, top_skills = top_skills))

valid_data <- valid_data |>
  mutate(skills_top = map(skills_list, keep_top_skills, top_skills = top_skills))

test_data <- test_data |>
  mutate(skills_top = map(skills_list, keep_top_skills, top_skills = top_skills))
```

# Build Binary Label Matrices

```{r build-label-matrices}
make_label_matrix <- function(df, top_skills) {
  y <- matrix(0L, nrow = nrow(df), ncol = length(top_skills))
  colnames(y) <- top_skills

  for (i in seq_len(nrow(df))) {
    present_skills <- df$skills_top[[i]]

    if (length(present_skills) > 0) {
      y[i, colnames(y) %in% present_skills] <- 1L
    }
  }

  y
}

y_train <- make_label_matrix(train_data, top_skills)
y_valid <- make_label_matrix(valid_data, top_skills)
y_test <- make_label_matrix(test_data, top_skills)
```

# Build TF-IDF Features

```{r build-tfidf-features}
corp_train <- corpus(train_data$profile_text_nostop)
corp_valid <- corpus(valid_data$profile_text_nostop)
corp_test <- corpus(test_data$profile_text_nostop)

dfm_train <- dfm(tokens(corp_train))
dfm_valid <- dfm(tokens(corp_valid))
dfm_test <- dfm(tokens(corp_test))

dfm_train <- dfm_trim(dfm_train, min_termfreq = 3)
feat_names <- featnames(dfm_train)

dfm_valid <- dfm_match(dfm_valid, features = feat_names)
dfm_test <- dfm_match(dfm_test, features = feat_names)

tfidf_train <- dfm_tfidf(dfm_train)
tfidf_valid <- dfm_tfidf(dfm_valid, force = TRUE)
tfidf_test <- dfm_tfidf(dfm_test, force = TRUE)

X_train <- as(tfidf_train, "dgCMatrix")
X_valid <- as(tfidf_valid, "dgCMatrix")
X_test <- as(tfidf_test, "dgCMatrix")

dim(X_train)
dim(X_valid)
dim(X_test)

dim(y_train)
dim(y_valid)
dim(y_test)
```

# Save Encoded Objects

```{r save-encoded-objects}
write_rds(train_data, file.path(processed_dir, "train_data.rds"))
write_rds(valid_data, file.path(processed_dir, "valid_data.rds"))
write_rds(test_data, file.path(processed_dir, "test_data.rds"))

write_rds(top_skills, file.path(processed_dir, "top_skills.rds"))

write_rds(X_train, file.path(processed_dir, "X_train.rds"))
write_rds(X_valid, file.path(processed_dir, "X_valid.rds"))
write_rds(X_test, file.path(processed_dir, "X_test.rds"))

write_rds(y_train, file.path(processed_dir, "y_train.rds"))
write_rds(y_valid, file.path(processed_dir, "y_valid.rds"))
write_rds(y_test, file.path(processed_dir, "y_test.rds"))

cat("Saved train, validation, test, TF-IDF, label, and top-skill objects to:", processed_dir, "\n")
```
