---
title: "Modeling - 500 Profile Dataset"
author: "Team"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(tidyverse)
library(glmnet)
library(xgboost)
library(Matrix)
library(here)

project_root <- normalizePath(file.path("..", "..", ".."))
models_dir <- file.path(project_root, "models")
results_dir <- file.path(project_root, "results")
processed_dir <- file.path(project_root, "data", "processed")

dir.create(models_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(results_dir, recursive = TRUE, showWarnings = FALSE)
```

# Introduction

In this section, we will train candidate multi-label models using a one-vs-rest
strategy for each target skill in the 500-profile dataset.

## What This File Does

- Loads TF-IDF feature matrices and binary skill target matrices.
- Trains one GLMNET model per skill using a one-vs-rest setup.
- Trains one XGBoost model per skill using the same target labels.
- Creates validation predictions for model comparison.
- Creates test predictions for final evaluation.
- Saves trained models and prediction objects in `models/` and `results/`.

# Load Encoded Inputs

```{r load-encoded-inputs}
X_train <- read_rds(file.path(processed_dir, "X_train.rds"))
X_valid <- read_rds(file.path(processed_dir, "X_valid.rds"))
X_test <- read_rds(file.path(processed_dir, "X_test.rds"))

y_train <- read_rds(file.path(processed_dir, "y_train.rds"))
y_valid <- read_rds(file.path(processed_dir, "y_valid.rds"))
y_test <- read_rds(file.path(processed_dir, "y_test.rds"))

top_skills <- read_rds(file.path(processed_dir, "top_skills.rds"))

dim(X_train)
dim(X_valid)
dim(X_test)

dim(y_train)
dim(y_valid)
dim(y_test)

top_skills
```

# GLMNET Helper Functions

```{r glmnet-helper-functions}
train_glmnet_multilabel <- function(X_train, y_train, top_skills) {
  models <- vector("list", length(top_skills))
  names(models) <- top_skills

  for (j in seq_along(top_skills)) {
    models[[j]] <- cv.glmnet(
      x = X_train,
      y = y_train[, j],
      family = "binomial",
      alpha = 0,
      type.measure = "class",
      nfolds = 5
    )
  }

  models
}

predict_glmnet_multilabel <- function(models, X, threshold = 0.50) {
  pred_probs <- sapply(models, function(m) {
    as.numeric(predict(m, newx = X, s = "lambda.min", type = "response"))
  })

  if (is.vector(pred_probs)) {
    pred_probs <- matrix(pred_probs, ncol = 1)
  }

  colnames(pred_probs) <- names(models)

  pred_class <- ifelse(pred_probs >= threshold, 1L, 0L)

  list(prob = pred_probs, class = pred_class)
}
```

# XGBoost Helper Functions

```{r xgboost-helper-functions}
train_xgb_multilabel <- function(X_train, y_train, top_skills) {
  models <- vector("list", length(top_skills))
  names(models) <- top_skills

  for (j in seq_along(top_skills)) {
    dtrain <- xgb.DMatrix(data = X_train, label = y_train[, j])

    models[[j]] <- xgb.train(
      params = list(
        objective = "binary:logistic",
        eval_metric = "logloss",
        eta = 0.1,
        max_depth = 6,
        subsample = 0.8,
        colsample_bytree = 0.8
      ),
      data = dtrain,
      nrounds = 100,
      verbose = 0
    )
  }

  models
}

predict_xgb_multilabel <- function(models, X, threshold = 0.50) {
  pred_probs <- sapply(models, function(m) {
    as.numeric(predict(m, newdata = X))
  })

  if (is.vector(pred_probs)) {
    pred_probs <- matrix(pred_probs, ncol = 1)
  }

  colnames(pred_probs) <- names(models)

  pred_class <- ifelse(pred_probs >= threshold, 1L, 0L)

  list(prob = pred_probs, class = pred_class)
}
```

# Train Candidate Models

```{r train-candidate-models}
set.seed(621)

glmnet_models <- train_glmnet_multilabel(X_train, y_train, top_skills)
xgb_models <- train_xgb_multilabel(X_train, y_train, top_skills)

saveRDS(glmnet_models, file.path(models_dir, "glmnet_multilabel_models.rds"))
saveRDS(xgb_models, file.path(models_dir, "xgb_multilabel_models.rds"))
```

# Validation Predictions

```{r validation-predictions}
glmnet_valid_pred <- predict_glmnet_multilabel(glmnet_models, X_valid, threshold = 0.50)
xgb_valid_pred <- predict_xgb_multilabel(xgb_models, X_valid, threshold = 0.50)

saveRDS(glmnet_valid_pred, file.path(results_dir, "glmnet_valid_pred.rds"))
saveRDS(xgb_valid_pred, file.path(results_dir, "xgb_valid_pred.rds"))
```

# Test Predictions

```{r test-predictions}
glmnet_test_pred <- predict_glmnet_multilabel(glmnet_models, X_test, threshold = 0.50)
xgb_test_pred <- predict_xgb_multilabel(xgb_models, X_test, threshold = 0.50)

saveRDS(glmnet_test_pred, file.path(results_dir, "glmnet_test_pred.rds"))
saveRDS(xgb_test_pred, file.path(results_dir, "xgb_test_pred.rds"))
```
