---
title: "500 Model: One-vs-Rest GLMNET"
author: "Team"
output: html_document
---

This notebook trains and evaluates the one-vs-rest GLMNET model for the
500-profile LinkedIn skills extraction dataset.

# Inputs and Skill Outputs

Input variables used by this model:

- `X_train`, `X_valid`, `X_test`: TF-IDF feature matrices created from `profile_text_nostop`.
- `y_train`, `y_valid`, `y_test`: binary skill-label matrices.
- `top_skills`: the 30 most frequent skills selected from the training set.

The model output is predicted skills for each profile. Each skill in `top_skills`
is predicted as:

- `1`: the model predicts the profile has the skill
- `0`: the model predicts the profile does not have the skill

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(tidyverse)
library(glmnet)
library(Matrix)
```

# Load Encoded 500 Inputs

```{r load-encoded-inputs}
project_root <- normalizePath(file.path("..", "..", "..", ".."))
processed_dir <- file.path(project_root, "data", "processed")
model_output_dir <- file.path(processed_dir, "500_model_outputs", "glmnet")

dir.create(model_output_dir, recursive = TRUE, showWarnings = FALSE)

X_train <- read_rds(file.path(processed_dir, "X_train.rds"))
X_valid <- read_rds(file.path(processed_dir, "X_valid.rds"))
X_test <- read_rds(file.path(processed_dir, "X_test.rds"))

y_train <- read_rds(file.path(processed_dir, "y_train.rds"))
y_valid <- read_rds(file.path(processed_dir, "y_valid.rds"))
y_test <- read_rds(file.path(processed_dir, "y_test.rds"))

top_skills <- read_rds(file.path(processed_dir, "top_skills.rds"))

dim(X_train)
dim(X_valid)
dim(X_test)

dim(y_train)
dim(y_valid)
dim(y_test)

top_skills
```

# Helper Functions

```{r helper-functions}
train_glmnet_multilabel <- function(X_train, y_train, top_skills) {
  models <- vector("list", length(top_skills))
  names(models) <- top_skills

  for (j in seq_along(top_skills)) {
    models[[j]] <- cv.glmnet(
      x = X_train,
      y = y_train[, j],
      family = "binomial",
      alpha = 0,
      type.measure = "class",
      nfolds = 5
    )
  }

  models
}

predict_glmnet_multilabel <- function(models, X, threshold = 0.50) {
  pred_probs <- sapply(models, function(m) {
    as.numeric(predict(m, newx = X, s = "lambda.min", type = "response"))
  })

  if (is.vector(pred_probs)) {
    pred_probs <- matrix(pred_probs, ncol = 1)
  }

  colnames(pred_probs) <- names(models)
  pred_class <- ifelse(pred_probs >= threshold, 1L, 0L)

  list(prob = pred_probs, class = pred_class)
}

calc_micro_metrics <- function(y_true, y_pred) {
  y_true_vec <- as.vector(y_true)
  y_pred_vec <- as.vector(y_pred)

  tp <- sum(y_true_vec == 1 & y_pred_vec == 1)
  fp <- sum(y_true_vec == 0 & y_pred_vec == 1)
  fn <- sum(y_true_vec == 1 & y_pred_vec == 0)

  precision <- ifelse(tp + fp == 0, 0, tp / (tp + fp))
  recall <- ifelse(tp + fn == 0, 0, tp / (tp + fn))
  f1 <- ifelse(precision + recall == 0, 0, 2 * precision * recall / (precision + recall))

  tibble(
    precision_micro = precision,
    recall_micro = recall,
    f1_micro = f1
  )
}

calc_macro_f1 <- function(y_true, y_pred, skill_names) {
  skill_f1 <- map_dbl(seq_along(skill_names), function(j) {
    yt <- y_true[, j]
    yp <- y_pred[, j]

    tp <- sum(yt == 1 & yp == 1)
    fp <- sum(yt == 0 & yp == 1)
    fn <- sum(yt == 1 & yp == 0)

    precision <- ifelse(tp + fp == 0, 0, tp / (tp + fp))
    recall <- ifelse(tp + fn == 0, 0, tp / (tp + fn))

    ifelse(precision + recall == 0, 0, 2 * precision * recall / (precision + recall))
  })

  tibble(macro_f1 = mean(skill_f1))
}

calc_subset_accuracy <- function(y_true, y_pred) {
  tibble(subset_accuracy = mean(apply(y_true == y_pred, 1, all)))
}

evaluate_multilabel <- function(y_true, y_pred, model_name, split_name, skill_names) {
  bind_cols(
    tibble(model = model_name, split = split_name),
    calc_micro_metrics(y_true, y_pred),
    calc_macro_f1(y_true, y_pred, skill_names),
    calc_subset_accuracy(y_true, y_pred)
  )
}
```

# Train GLMNET Model

```{r train-glmnet}
set.seed(621)

glmnet_models <- train_glmnet_multilabel(X_train, y_train, top_skills)

saveRDS(glmnet_models, file.path(model_output_dir, "glmnet_500_models.rds"))
```

# Validation and Test Predictions

```{r predictions}
glmnet_valid_pred <- predict_glmnet_multilabel(glmnet_models, X_valid, threshold = 0.50)
glmnet_test_pred <- predict_glmnet_multilabel(glmnet_models, X_test, threshold = 0.50)

validation_metrics <- evaluate_multilabel(
  y_valid,
  glmnet_valid_pred$class,
  "GLMNET",
  "Validation",
  top_skills
)

test_metrics <- evaluate_multilabel(
  y_test,
  glmnet_test_pred$class,
  "GLMNET",
  "Test",
  top_skills
)

validation_metrics
test_metrics
```

# Save Outputs

```{r save-outputs}
glmnet_results <- list(
  model = "GLMNET",
  validation_metrics = validation_metrics,
  test_metrics = test_metrics,
  models = glmnet_models,
  validation_predictions = glmnet_valid_pred,
  test_predictions = glmnet_test_pred,
  top_skills = top_skills
)

saveRDS(glmnet_results, file.path(model_output_dir, "glmnet_500_results.rds"))
write_csv(validation_metrics, file.path(model_output_dir, "validation_metrics.csv"))
write_csv(test_metrics, file.path(model_output_dir, "test_metrics.csv"))

cat("GLMNET outputs saved to:", model_output_dir, "\n")
```
