---
title: "CP8 Modeling Notebook Template for Question B"
author: "Lincoln Chanakira"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Overview

This notebook provides an R Markdown template for evaluating four candidate models for **Question B: Skills Extraction** using the team's prepared **training, validation, and test** splits. This aligns with the CP7 expectation that data be partitioned into held-out sets for modeling and evaluation, and with the CP8 requirement to compare multiple candidate models on validation or test data. fileciteturn1file0 fileciteturn1file1

The models included are:

1. **Regularized Logistic Regression**
2. **XGBoost**
3. **Feature-Enhanced Model (TF-IDF + derived document features)**
4. **kNN** *(optional)*

**Important note:** the uploaded split files appear to contain a wide feature matrix plus an identifier column, but they do not expose an obvious target label in the version I inspected. Before running this notebook, set `target_col` to the actual binary skill label you want to predict.

# Required packages

```{r packages}
required_packages <- c(
  "readr", "dplyr", "Matrix", "glmnet", "xgboost", "class", "pROC"
)

missing_pkgs <- required_packages[!sapply(required_packages, requireNamespace, quietly = TRUE)]
if (length(missing_pkgs) > 0) {
  install.packages(missing_pkgs, repos='https://cran.rstudio.com/')
}

library(readr)
library(dplyr)
library(Matrix)
library(glmnet)
library(xgboost)
library(class)
library(pROC)
```

# Load split data

Place this `.Rmd` file in the same folder as:
- `train_data.csv`
- `validation_data.csv`
- `test_data.csv`

```{r load-data}
train_path <- "../data/Train, Test, and Validation /train_data.csv"
valid_path <- "../data/Train, Test, and Validation /validation_data.csv"
test_path  <- "../data/Train, Test, and Validation /test_data.csv"

train_df <- read_csv(train_path, show_col_types = FALSE)
valid_df <- read_csv(valid_path, show_col_types = FALSE)
test_df  <- read_csv(test_path, show_col_types = FALSE)

cat("Train shape:", dim(train_df), "\n")
cat("Validation shape:", dim(valid_df), "\n")
cat("Test shape:", dim(test_df), "\n")
```

# User settings

Replace `YOUR_LABEL_HERE` with the actual skill label column you want to model.

Examples:
- `python`
- `sql`
- `machine_learning`

```{r user-settings}
target_col <- "data_analysis"
id_cols <- c("linkedinUrl")

if (!(target_col %in% names(train_df))) {
  stop(
    paste0(
      "target_col='", target_col, "' was not found in train_df. ",
      "Set target_col to an actual binary label column before running this notebook."
    )
  )
}
```

# Basic validation and alignment

```{r validation}
common_cols <- Reduce(intersect, list(names(train_df), names(valid_df), names(test_df)))
train_df <- train_df[, common_cols]
valid_df <- valid_df[, common_cols]
test_df  <- test_df[, common_cols]

train_df <- train_df %>% filter(!is.na(.data[[target_col]]))
valid_df <- valid_df %>% filter(!is.na(.data[[target_col]]))
test_df  <- test_df %>% filter(!is.na(.data[[target_col]]))

train_df[[target_col]] <- as.integer(train_df[[target_col]])
valid_df[[target_col]] <- as.integer(valid_df[[target_col]])
test_df[[target_col]]  <- as.integer(test_df[[target_col]])

predictor_cols <- setdiff(common_cols, c(id_cols, target_col))

cat("Number of predictors:", length(predictor_cols), "\n")
cat("Positive rate (train):", round(mean(train_df[[target_col]]), 4), "\n")
cat("Positive rate (validation):", round(mean(valid_df[[target_col]]), 4), "\n")
cat("Positive rate (test):", round(mean(test_df[[target_col]]), 4), "\n")
```

# Helper functions

```{r helpers}
calc_metrics <- function(actual, prob, threshold = 0.5) {
  pred <- ifelse(prob >= threshold, 1, 0)
  tp <- sum(pred == 1 & actual == 1)
  tn <- sum(pred == 0 & actual == 0)
  fp <- sum(pred == 1 & actual == 0)
  fn <- sum(pred == 0 & actual == 1)

  precision <- ifelse(tp + fp == 0, NA, tp / (tp + fp))
  recall    <- ifelse(tp + fn == 0, NA, tp / (tp + fn))
  f1        <- ifelse(is.na(precision) || is.na(recall) || (precision + recall) == 0,
                      NA, 2 * precision * recall / (precision + recall))
  accuracy  <- mean(pred == actual)
  auc       <- tryCatch(as.numeric(roc(actual, prob, quiet = TRUE)$auc), error = function(e) NA)

  data.frame(
    threshold = threshold,
    accuracy = accuracy,
    precision = precision,
    recall = recall,
    f1 = f1,
    auc = auc,
    tp = tp,
    tn = tn,
    fp = fp,
    fn = fn
  )
}

find_best_threshold <- function(actual, prob, grid = seq(0.10, 0.90, by = 0.05)) {
  scores <- bind_rows(lapply(grid, function(th) calc_metrics(actual, prob, threshold = th)))
  scores %>% arrange(desc(f1), desc(recall), desc(precision)) %>% slice(1)
}

make_sparse_matrix <- function(df, cols) {
  mat <- as.matrix(df[, cols, drop = FALSE])
  Matrix(mat, sparse = TRUE)
}
```

# Build sparse matrices

```{r sparse-matrices}
X_train <- make_sparse_matrix(train_df, predictor_cols)
X_valid <- make_sparse_matrix(valid_df, predictor_cols)
X_test  <- make_sparse_matrix(test_df, predictor_cols)

y_train <- train_df[[target_col]]
y_valid <- valid_df[[target_col]]
y_test  <- test_df[[target_col]]
```

# Model 1: Regularized Logistic Regression

This is a strong baseline for sparse high-dimensional text features.

```{r logistic-model}
set.seed(621)
logit_cv <- cv.glmnet(
  x = X_train,
  y = y_train,
  family = "binomial",
  alpha = 1,
  type.measure = "class",
  nfolds = 5,
  standardize = TRUE
)

logit_valid_prob <- as.numeric(predict(logit_cv, newx = X_valid, s = "lambda.min", type = "response"))
logit_best_th <- find_best_threshold(y_valid, logit_valid_prob)
logit_test_prob <- as.numeric(predict(logit_cv, newx = X_test, s = "lambda.min", type = "response"))
logit_test_metrics <- calc_metrics(y_test, logit_test_prob, threshold = logit_best_th$threshold)

logit_test_metrics
```

# Model 2: XGBoost

This is a strong non-linear model and is often competitive when the feature space is rich.

```{r xgboost-model}
dtrain <- xgb.DMatrix(data = X_train, label = y_train)
dvalid <- xgb.DMatrix(data = X_valid, label = y_valid)
dtest  <- xgb.DMatrix(data = X_test, label = y_test)

xgb_params <- list(
  objective = "binary:logistic",
  eval_metric = "auc",
  eta = 0.05,
  max_depth = 6,
  min_child_weight = 1,
  subsample = 0.8,
  colsample_bytree = 0.8
)

xgb_fit <- xgb.train(
  params = xgb_params,
  data = dtrain,
  nrounds = 300,
  watchlist = list(train = dtrain, valid = dvalid),
  early_stopping_rounds = 20,
  verbose = 0
)

xgb_valid_prob <- predict(xgb_fit, newdata = dvalid)
xgb_best_th <- find_best_threshold(y_valid, xgb_valid_prob)
xgb_test_prob <- predict(xgb_fit, newdata = dtest)
xgb_test_metrics <- calc_metrics(y_test, xgb_test_prob, threshold = xgb_best_th$threshold)

xgb_test_metrics
```

# Model 3: Feature-Enhanced Model

Because the uploaded split files appear to be feature matrices rather than raw profile records, this notebook derives additional document-level features from the existing predictors.

```{r enhanced-model}
add_doc_features <- function(df, cols) {
  x <- as.matrix(df[, cols, drop = FALSE])
  df %>%
    mutate(
      nnz_terms = rowSums(x != 0),
      tfidf_sum = rowSums(x),
      tfidf_mean = rowMeans(x),
      tfidf_max = apply(x, 1, max),
      tfidf_sd = apply(x, 1, sd)
    )
}

train_feat <- add_doc_features(train_df, predictor_cols)
valid_feat <- add_doc_features(valid_df, predictor_cols)
test_feat  <- add_doc_features(test_df, predictor_cols)

feature_cols_enhanced <- c(predictor_cols, "nnz_terms", "tfidf_sum", "tfidf_mean", "tfidf_max", "tfidf_sd")

X_train_enh <- make_sparse_matrix(train_feat, feature_cols_enhanced)
X_valid_enh <- make_sparse_matrix(valid_feat, feature_cols_enhanced)
X_test_enh  <- make_sparse_matrix(test_feat, feature_cols_enhanced)

enh_cv <- cv.glmnet(
  x = X_train_enh,
  y = y_train,
  family = "binomial",
  alpha = 0.5,
  type.measure = "class",
  nfolds = 5,
  standardize = TRUE
)

enh_valid_prob <- as.numeric(predict(enh_cv, newx = X_valid_enh, s = "lambda.min", type = "response"))
enh_best_th <- find_best_threshold(y_valid, enh_valid_prob)
enh_test_prob <- as.numeric(predict(enh_cv, newx = X_test_enh, s = "lambda.min", type = "response"))
enh_test_metrics <- calc_metrics(y_test, enh_test_prob, threshold = enh_best_th$threshold)

enh_test_metrics
```

# Model 4: kNN (optional)

kNN can be computationally expensive with wide sparse matrices, so this version uses a reduced set of high-variance predictors.

```{r knn-model}
variance_vec <- apply(train_df[, predictor_cols, drop = FALSE], 2, var)
top_knn_cols <- names(sort(variance_vec, decreasing = TRUE))[1:min(100, length(variance_vec))]

scale_train <- scale(train_df[, top_knn_cols, drop = FALSE])
train_center <- attr(scale_train, "scaled:center")
train_scale  <- attr(scale_train, "scaled:scale")
train_scale[train_scale == 0] <- 1

scale_valid <- scale(valid_df[, top_knn_cols, drop = FALSE], center = train_center, scale = train_scale)
scale_test  <- scale(test_df[, top_knn_cols, drop = FALSE], center = train_center, scale = train_scale)

knn_valid_pred <- knn(
  train = scale_train,
  test = scale_valid,
  cl = factor(y_train),
  k = 5,
  prob = TRUE
)

knn_valid_prob <- ifelse(knn_valid_pred == 1,
                         attr(knn_valid_pred, "prob"),
                         1 - attr(knn_valid_pred, "prob"))
knn_best_th <- find_best_threshold(y_valid, knn_valid_prob)

knn_test_pred <- knn(
  train = scale_train,
  test = scale_test,
  cl = factor(y_train),
  k = 5,
  prob = TRUE
)
knn_test_prob <- ifelse(knn_test_pred == 1,
                        attr(knn_test_pred, "prob"),
                        1 - attr(knn_test_pred, "prob"))
knn_test_metrics <- calc_metrics(y_test, knn_test_prob, threshold = knn_best_th$threshold)

knn_test_metrics
```

# Compare candidate models

```{r compare-results}
results <- bind_rows(
  cbind(model = "Regularized Logistic Regression", logit_test_metrics),
  cbind(model = "XGBoost", xgb_test_metrics),
  cbind(model = "Feature-Enhanced Elastic Net", enh_test_metrics),
  cbind(model = "kNN", knn_test_metrics)
) %>%
  arrange(desc(f1), desc(auc), desc(recall))

results
```

# Export results

```{r export-results}
write_csv(results, paste0("model_results_", target_col, ".csv"))
```

# Optional extension: one-vs-rest loop for multiple skills

If your dataset contains multiple binary skill label columns, you can adapt the notebook into a one-vs-rest workflow. A simple template is below.

```{r one-vs-rest-template, eval=FALSE}
skill_labels <- c("python", "sql", "machine_learning")
all_skill_results <- list()

for (skill in skill_labels) {
  message("Running models for label: ", skill)
  # Wrap the modeling blocks above into a function and call it here.
}
```

# Interpretation notes for CP8

When you build your CP8 table, report:
- the **model name/configuration**
- the **input feature set**
- the **primary metric**
- the **test or validation performance score**
- whether the model is a **top candidate**
- any notes on interpretability, computational cost, and limitations

For Question B, **F1-score** is usually the strongest primary metric because it balances precision and recall in a skills extraction setting, which matches the evaluation rationale already described in your CP6 plan. fileciteturn1file2
