---
title: "5k vs 500 Dataset Comparison"
author: "Team"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(tidyverse)
library(tidytext)
library(here)

processed_dir <- file.path("..", "data", "processed")
results_dir <- file.path("..", "results")
comparison_dir <- file.path("..", "results", "dataset_comparison")

dir.create(comparison_dir, recursive = TRUE, showWarnings = FALSE)
```

# Introduction

This notebook compares the original 5k LinkedIn dataset and the 500-profile
dataset at the processed-data level. The goal is to show differences in dataset
size, text coverage, skill label coverage, and modeling readiness, then recommend
which dataset is stronger for the CP8 skills extraction pipeline.

# Load Processed Datasets

```{r load-processed-datasets}
data_5k_path <- file.path(processed_dir, "linkedin_full_5865_preprocessed.csv")
data_500_path <- file.path(processed_dir, "clean_text_data.rds")

data_5k_raw <- read_csv(data_5k_path, show_col_types = FALSE)
data_500_raw <- read_rds(data_500_path)

cat("5k processed data shape:", dim(data_5k_raw), "\n")
cat("500 processed data shape:", dim(data_500_raw), "\n")

glimpse(data_5k_raw)
glimpse(data_500_raw)
```

# Harmonize Common Fields

```{r harmonize-common-fields}
split_skill_string <- function(x) {
  x <- ifelse(is.na(x), "", as.character(x))
  x <- str_to_lower(x)
  x <- str_replace_all(x, "\\s*[,;|]\\s*", ",")
  str_split(x, ",") |>
    map(~ {
      skills <- str_squish(.x)
      skills <- skills[skills != ""]
      unique(skills)
    })
}

count_words <- function(x) {
  str_count(ifelse(is.na(x), "", x), "\\S+")
}

data_5k <- data_5k_raw |>
  mutate(
    dataset = "5k dataset",
    profile_text_compare = profile_text,
    token_count_compare = count_words(profile_text_compare),
    char_count_compare = nchar(profile_text_compare),
    skills_compare = split_skill_string(skills),
    skill_count_compare = lengths(skills_compare),
    has_any_skill_compare = if_else(skill_count_compare > 0, 1L, 0L)
  ) |>
  select(
    dataset,
    linkedin_url,
    profile_text_compare,
    token_count_compare,
    char_count_compare,
    skills_compare,
    skill_count_compare,
    has_any_skill_compare,
    has_current_job_description,
    has_previous_job_description,
    total_experience_count
  )

data_500 <- data_500_raw |>
  mutate(
    dataset = "500 dataset",
    profile_text_compare = profile_text_clean,
    token_count_compare = token_count,
    char_count_compare = char_count,
    skills_compare = skills_list,
    skill_count_compare = skill_count,
    has_any_skill_compare = has_any_skill
  ) |>
  select(
    dataset,
    linkedin_url,
    profile_text_compare,
    token_count_compare,
    char_count_compare,
    skills_compare,
    skill_count_compare,
    has_any_skill_compare,
    has_current_job_description,
    has_previous_job_description,
    total_experience_count
  )

combined_processed <- bind_rows(data_5k, data_500)

combined_processed |>
  count(dataset)
```

# Summary Comparison

```{r summary-comparison}
summary_comparison <- combined_processed |>
  group_by(dataset) |>
  summarise(
    profiles = n(),
    profiles_with_skills = sum(has_any_skill_compare),
    pct_profiles_with_skills = mean(has_any_skill_compare) * 100,
    avg_skill_count = mean(skill_count_compare),
    median_skill_count = median(skill_count_compare),
    avg_token_count = mean(token_count_compare),
    median_token_count = median(token_count_compare),
    avg_char_count = mean(char_count_compare),
    pct_current_description = mean(has_current_job_description) * 100,
    pct_previous_description = mean(has_previous_job_description) * 100,
    avg_experience_count = mean(total_experience_count),
    .groups = "drop"
  )

summary_comparison

write_csv(summary_comparison, file.path(comparison_dir, "dataset_summary_comparison.csv"))
```

# Unique Skill Coverage

```{r unique-skill-coverage}
skill_coverage <- combined_processed |>
  select(dataset, skills_compare) |>
  unnest(skills_compare) |>
  filter(!is.na(skills_compare), skills_compare != "") |>
  group_by(dataset) |>
  summarise(
    unique_skills = n_distinct(skills_compare),
    total_skill_assignments = n(),
    .groups = "drop"
  )

skill_coverage

write_csv(skill_coverage, file.path(comparison_dir, "skill_coverage_comparison.csv"))
```

# Top Skills by Dataset

```{r top-skills}
top_skills_by_dataset <- combined_processed |>
  select(dataset, skills_compare) |>
  unnest(skills_compare) |>
  filter(!is.na(skills_compare), skills_compare != "") |>
  count(dataset, skills_compare, sort = TRUE) |>
  group_by(dataset) |>
  slice_max(order_by = n, n = 15, with_ties = FALSE) |>
  ungroup()

top_skills_by_dataset

write_csv(top_skills_by_dataset, file.path(comparison_dir, "top_skills_by_dataset.csv"))
```

# Graphical Comparison

```{r graphical-comparison, fig.width=10, fig.height=6}
profile_count_plot <- summary_comparison |>
  ggplot(aes(x = dataset, y = profiles, fill = dataset)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = scales::comma(profiles)), vjust = -0.3) +
  labs(
    title = "Profile Count by Dataset",
    x = NULL,
    y = "Profiles"
  ) +
  theme_minimal()

profile_count_plot

ggsave(
  filename = file.path(comparison_dir, "profile_count_comparison.png"),
  plot = profile_count_plot,
  width = 10,
  height = 6
)
```

```{r token-distribution-plot, fig.width=10, fig.height=6}
token_distribution_plot <- combined_processed |>
  ggplot(aes(x = token_count_compare, fill = dataset)) +
  geom_histogram(bins = 40, alpha = 0.65, position = "identity") +
  facet_wrap(~ dataset, scales = "free_y") +
  labs(
    title = "Profile Text Token Count Distribution",
    x = "Token count",
    y = "Profiles"
  ) +
  theme_minimal()

token_distribution_plot

ggsave(
  filename = file.path(comparison_dir, "token_count_distribution.png"),
  plot = token_distribution_plot,
  width = 10,
  height = 6
)
```

```{r skill-count-plot, fig.width=10, fig.height=6}
skill_count_plot <- combined_processed |>
  ggplot(aes(x = skill_count_compare, fill = dataset)) +
  geom_histogram(bins = 35, alpha = 0.65, position = "identity") +
  facet_wrap(~ dataset, scales = "free_y") +
  labs(
    title = "Skill Count Distribution",
    x = "Skills per profile",
    y = "Profiles"
  ) +
  theme_minimal()

skill_count_plot

ggsave(
  filename = file.path(comparison_dir, "skill_count_distribution.png"),
  plot = skill_count_plot,
  width = 10,
  height = 6
)
```

```{r top-skills-plot, fig.width=12, fig.height=8}
top_skills_plot <- top_skills_by_dataset |>
  mutate(skills_compare = reorder_within(skills_compare, n, dataset)) |>
  ggplot(aes(x = skills_compare, y = n, fill = dataset)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  facet_wrap(~ dataset, scales = "free_y") +
  scale_x_reordered() +
  labs(
    title = "Top 15 Skills by Dataset",
    x = NULL,
    y = "Profiles with skill"
  ) +
  theme_minimal()

top_skills_plot

ggsave(
  filename = file.path(comparison_dir, "top_skills_comparison.png"),
  plot = top_skills_plot,
  width = 12,
  height = 8
)
```

# Modeling Result Comparison

```{r modeling-result-comparison}
model_results_5k_path <- file.path(processed_dir, "cp8_model_results.rds")
model_results_500_path <- file.path(results_dir, "test_results.csv")

modeling_comparison <- tibble()

if (file.exists(model_results_5k_path)) {
  results_5k <- read_rds(model_results_5k_path)$test_results |>
    mutate(dataset = "5k dataset") |>
    select(dataset, model, precision_micro, recall_micro, f1_micro, f1_macro)

  modeling_comparison <- bind_rows(modeling_comparison, results_5k)
}

if (file.exists(model_results_500_path)) {
  results_500 <- read_csv(model_results_500_path, show_col_types = FALSE) |>
    mutate(dataset = "500 dataset") |>
    select(dataset, model, precision_micro, recall_micro, f1_micro, f1_macro = macro_f1)

  modeling_comparison <- bind_rows(modeling_comparison, results_500)
}

modeling_comparison

write_csv(modeling_comparison, file.path(comparison_dir, "modeling_test_result_comparison.csv"))
```

```{r model-performance-plot, fig.width=10, fig.height=6}
if (nrow(modeling_comparison) > 0) {
  model_performance_plot <- modeling_comparison |>
    ggplot(aes(x = model, y = f1_micro, fill = dataset)) +
    geom_col(position = "dodge") +
    labs(
      title = "Test Micro F1 by Dataset and Model",
      x = "Model",
      y = "Test micro F1"
    ) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 25, hjust = 1))

  model_performance_plot

  ggsave(
    filename = file.path(comparison_dir, "test_micro_f1_comparison.png"),
    plot = model_performance_plot,
    width = 10,
    height = 6
  )
}
```

# Dataset Recommendation

```{r dataset-recommendation}
best_processed_dataset <- summary_comparison |>
  mutate(
    modeling_readiness_score =
      profiles +
      profiles_with_skills +
      unique(skill_coverage$unique_skills[match(dataset, skill_coverage$dataset)])
  ) |>
  arrange(desc(modeling_readiness_score))

best_processed_dataset

best_modeling_result <- modeling_comparison |>
  arrange(desc(f1_micro)) |>
  slice_head(n = 1)

best_modeling_result
```

## Recommendation

Based on the processed-data comparison, the **5k dataset is the better dataset
to use for the final modeling pipeline**. It has many more profiles, more
training examples with skill labels, broader label coverage, and stronger held
out model performance in the current runs.

The **500 dataset is still useful** for quick testing, debugging, and classroom
demonstration because it runs much faster. However, for the final CP8 modeling
result, the 5k dataset provides the stronger foundation.
