---
title: "CP8 Data Preparation"
author: "Lincoln Chanakira"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Load Packages

```{r packages}
library(readr)
library(readxl)
library(dplyr)
```

# Import Dataset

```{r import-5k-data}
linkedin_full_path <- "../data/raw/LinkedIn_Scrape_full_5865.xlsx"

linkedin_full <- read_excel(linkedin_full_path, sheet = "Scraped data")

cat("LinkedIn full shape:", dim(linkedin_full), "\n")
```

# Inspect Imported Data

```{r inspect-5k-data}
glimpse(linkedin_full)

head(linkedin_full)
```

# Remove Columns and Save Prepared Dataset

```{r remove-columns-save}
columns_to_remove <- c(
  "full_name",
  "email",
  "education_start_year",
  "education_end_year",
  "current_company_url",
  "Company_size",
  "current_job_start_date",
  "last_scraped_date",
  "previous_job_start_date",
  "previous_job_end_date",
  "previous_company_size",
  "previous_company_name",
  "employment_type",
  "is_currently_employed",
  "current_job_duration_years"
)

linkedin_prepared <- linkedin_full |>
  select(-any_of(columns_to_remove))

prepared_output_path <- "../data/processed/linkedin_full_5865_prepared.csv"
write_csv(linkedin_prepared, prepared_output_path)

cat("Prepared dataset shape:", dim(linkedin_prepared), "\n")
cat("Prepared dataset saved to:", prepared_output_path, "\n")
glimpse(linkedin_prepared)
```

# External Validation Data Preparation

This section prepares the joined KSB alumni + LinkedIn external validation file
for the same downstream preprocessing workflow. It does not change the main 5k
workflow above; it writes a separate prepared CSV for external validation.

```{r prepare-external-validation-data}
# Import the joined external validation dataset created in cleaningScripts.
external_validation_input_path <- "../data/externalvalidation/ksb_alumni_linkedin_joined.csv"

ksb_external_validation <- read_csv(
  external_validation_input_path,
  show_col_types = FALSE
)

external_validation_columns <- c(
  "City",
  "State",
  "Country",
  "Job_Title",
  "Employer",
  "Grad_Year",
  "Degree",
  "Degree_Program",
  "Degree_Major",
  "Index",
  "field_of_study",
  "education_start_year",
  "education_end_year",
  "current_job_title",
  "current_job_description",
  "current_company_name",
  "current_company_url",
  "current_job_industry",
  "current_job_country",
  "current_job_location",
  "current_job_start_date",
  "previous_job_title",
  "previous_job_description",
  "previous_company_name",
  "total_experience_years",
  "Skills",
  "total_experience_count"
)

# Create any expected external-validation columns that are absent so the rest of
# the CP8 pipeline always receives the same schema.
missing_external_validation_columns <- setdiff(
  external_validation_columns,
  names(ksb_external_validation)
)

for (column_name in missing_external_validation_columns) {
  ksb_external_validation[[column_name]] <- NA
}

# The external validation file does not include linkedin_url/headline in the same
# way as the raw 5k scrape, so create stable workflow-compatible fields.
if (!"linkedin_url" %in% names(ksb_external_validation)) {
  ksb_external_validation <- ksb_external_validation |>
    mutate(linkedin_url = paste0("external_validation_index_", Index))
}

if (!"headline" %in% names(ksb_external_validation)) {
  ksb_external_validation <- ksb_external_validation |>
    mutate(headline = coalesce(as.character(current_job_title), as.character(Job_Title), ""))
}

ksb_external_validation_prepared <- ksb_external_validation |>
  mutate(
    linkedin_url = as.character(linkedin_url),
    headline = as.character(headline),
    Skills = as.character(Skills)
  ) |>
  # External validation labels come from Skills, so remove rows without labels.
  filter(!is.na(Skills), trimws(Skills) != "") |>
  select(
    linkedin_url,
    headline,
    all_of(external_validation_columns)
  )

external_prepared_output_path <- "../data/processed/ksb_alumni_linkedin_prepared.csv"
write_csv(ksb_external_validation_prepared, external_prepared_output_path)

cat("External validation prepared shape:", dim(ksb_external_validation_prepared), "\n")
cat("External validation prepared dataset saved to:", external_prepared_output_path, "\n")
cat("Missing expected external-validation columns created as NA:",
    length(missing_external_validation_columns), "\n")
glimpse(ksb_external_validation_prepared)
```
