---
title: "External Validation Modeling"
author: "Lincoln Chanakira"
output: html_document
---

This notebook runs the KSB alumni + LinkedIn external validation dataset through
all saved 5k models. It does not retrain the models and does not change the
existing CP8 modeling workflow. It loads the already encoded external validation
TF-IDF matrix and evaluates each saved model against the external validation
skill labels.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Load Packages

```{r packages}
library(readr)
library(dplyr)
library(tibble)
library(purrr)
library(Matrix)
library(glmnet)
library(xgboost)
library(class)
```

# File Paths

```{r paths}
processed_dir <- "../data/processed"

split_input_path <- file.path(processed_dir, "train_validation_test_split.rds")
external_tfidf_path <- file.path(processed_dir, "ksb_alumni_linkedin_tfidf.rds")
external_targets_path <- file.path(processed_dir, "ksb_alumni_linkedin_multilabel_targets.csv")
external_profiles_path <- file.path(processed_dir, "ksb_alumni_linkedin_encoded_profiles.csv")

model_output_root <- file.path(processed_dir, "5k_model_outputs")
external_output_root <- file.path(processed_dir, "external_validation_model_outputs")

dir.create(external_output_root, recursive = TRUE, showWarnings = FALSE)
```

# Load Training and External Validation Data

```{r load-data}
# Load training split objects because KNN needs the original training feature
# matrix and labels to classify new external validation profiles.
split_objects <- readRDS(split_input_path)

X_train <- split_objects$X_train
y_train_full <- as.matrix(split_objects$y_train)

# Load external validation model inputs created by model_input_encoding.rmd.
X_external <- readRDS(external_tfidf_path)
external_targets <- read_csv(external_targets_path, show_col_types = FALSE)
external_profiles <- read_csv(external_profiles_path, show_col_types = FALSE)

cat("Training feature matrix:", dim(X_train), "\n")
cat("External validation feature matrix:", dim(X_external), "\n")
cat("External validation targets:", dim(external_targets), "\n")
cat("External validation profiles:", dim(external_profiles), "\n")
```

# Helper Functions

```{r helper-functions}
# Divide safely so metrics return zero instead of NaN when a model predicts no
# positives or a label has no positives.
safe_divide <- function(num, den) {
  ifelse(den == 0, 0, num / den)
}

evaluate_multilabel <- function(y_true, y_pred_binary) {
  y_true <- as.matrix(y_true)
  y_pred_binary <- as.matrix(y_pred_binary)

  tp_micro <- sum((y_true == 1) & (y_pred_binary == 1))
  fp_micro <- sum((y_true == 0) & (y_pred_binary == 1))
  fn_micro <- sum((y_true == 1) & (y_pred_binary == 0))

  precision_micro <- safe_divide(tp_micro, tp_micro + fp_micro)
  recall_micro <- safe_divide(tp_micro, tp_micro + fn_micro)
  f1_micro <- safe_divide(
    2 * precision_micro * recall_micro,
    precision_micro + recall_micro
  )

  precision_by_label <- numeric(ncol(y_true))
  recall_by_label <- numeric(ncol(y_true))
  f1_by_label <- numeric(ncol(y_true))

  for (j in seq_len(ncol(y_true))) {
    tp <- sum((y_true[, j] == 1) & (y_pred_binary[, j] == 1))
    fp <- sum((y_true[, j] == 0) & (y_pred_binary[, j] == 1))
    fn <- sum((y_true[, j] == 1) & (y_pred_binary[, j] == 0))

    p <- safe_divide(tp, tp + fp)
    r <- safe_divide(tp, tp + fn)
    f <- safe_divide(2 * p * r, p + r)

    precision_by_label[j] <- p
    recall_by_label[j] <- r
    f1_by_label[j] <- f
  }

  tibble(
    precision_micro = precision_micro,
    recall_micro = recall_micro,
    f1_micro = f1_micro,
    precision_macro = mean(precision_by_label, na.rm = TRUE),
    recall_macro = mean(recall_by_label, na.rm = TRUE),
    f1_macro = mean(f1_by_label, na.rm = TRUE)
  )
}

predict_binary_from_probs <- function(prob_matrix, threshold = 0.50) {
  prob_matrix <- as.matrix(prob_matrix)

  if (length(threshold) == 1) {
    return(ifelse(prob_matrix >= threshold, 1L, 0L))
  }

  threshold_matrix <- matrix(
    rep(threshold, each = nrow(prob_matrix)),
    nrow = nrow(prob_matrix),
    byrow = FALSE
  )

  ifelse(prob_matrix >= threshold_matrix, 1L, 0L)
}

get_thresholds <- function(model_results, retained_skills) {
  thresholds <- model_results$thresholds

  if (is.null(names(thresholds))) {
    names(thresholds) <- model_results$retained_skills
  }

  thresholds <- thresholds[retained_skills]
  thresholds[is.na(thresholds)] <- 0.50
  thresholds
}

make_prediction_summary <- function(y_true, pred_binary, prob_matrix) {
  tibble(
    skill = colnames(y_true),
    actual_count = as.numeric(colSums(y_true)),
    predicted_count = as.numeric(colSums(pred_binary)),
    mean_probability = as.numeric(colMeans(prob_matrix))
  ) |>
    arrange(desc(actual_count), desc(predicted_count))
}

save_external_model_outputs <- function(model_slug, model_name, probs, preds, metrics, summary) {
  model_dir <- file.path(external_output_root, model_slug)
  dir.create(model_dir, recursive = TRUE, showWarnings = FALSE)

  saveRDS(
    list(
      model = model_name,
      external_metrics = metrics,
      external_probabilities = probs,
      external_predictions = preds,
      prediction_summary = summary,
      retained_skills = colnames(preds)
    ),
    file.path(model_dir, paste0(model_slug, "_external_validation_results.rds"))
  )

  write_csv(metrics, file.path(model_dir, "external_validation_metrics.csv"))
  write_csv(summary, file.path(model_dir, "external_prediction_summary.csv"))
}
```

# Align External Targets

```{r align-external-targets}
# Use the same 25 modeled labels saved by the model notebooks. All models use
# the same retained skill order, so the external target matrix is aligned once.
baseline_results <- readRDS(file.path(
  model_output_root,
  "baseline_frequency_prior",
  "baseline_frequency_prior_5k_results.rds"
))

retained_skills <- baseline_results$retained_skills

missing_external_labels <- setdiff(retained_skills, names(external_targets))
if (length(missing_external_labels) > 0) {
  stop(
    "External targets are missing modeled skill labels: ",
    paste(missing_external_labels, collapse = ", ")
  )
}

y_external <- external_targets |>
  select(all_of(retained_skills)) |>
  as.matrix()

cat("External labels evaluated:", ncol(y_external), "\n")
print(retained_skills)
```

# Baseline Frequency Prior

```{r run-baseline-external-validation}
# The baseline model predicts each skill using the training-set prevalence.
baseline_thresholds <- get_thresholds(baseline_results, retained_skills)
baseline_prevalence <- colMeans(y_train_full[, retained_skills, drop = FALSE])

baseline_external_probs <- matrix(
  rep(baseline_prevalence, each = nrow(y_external)),
  nrow = nrow(y_external),
  byrow = FALSE
)
colnames(baseline_external_probs) <- retained_skills

baseline_external_pred <- predict_binary_from_probs(
  baseline_external_probs,
  threshold = baseline_thresholds
)
colnames(baseline_external_pred) <- retained_skills

baseline_external_metrics <- evaluate_multilabel(
  y_external,
  baseline_external_pred
) |>
  mutate(model = "Baseline Frequency Prior", .before = 1)

baseline_external_summary <- make_prediction_summary(
  y_external,
  baseline_external_pred,
  baseline_external_probs
)

save_external_model_outputs(
  "baseline_frequency_prior",
  "Baseline Frequency Prior",
  baseline_external_probs,
  baseline_external_pred,
  baseline_external_metrics,
  baseline_external_summary
)

baseline_external_metrics
```

# Logistic Regression

```{r run-logistic-external-validation}
logit_results <- readRDS(file.path(
  model_output_root,
  "logistic_regression",
  "logistic_regression_5k_results.rds"
))

logit_thresholds <- get_thresholds(logit_results, retained_skills)

logit_external_probs <- matrix(
  0,
  nrow = nrow(X_external),
  ncol = length(retained_skills)
)
colnames(logit_external_probs) <- retained_skills

for (j in seq_along(retained_skills)) {
  model_j <- logit_results$models[[j]]

  logit_external_probs[, j] <- as.numeric(
    predict(model_j, newx = X_external, s = "lambda.min", type = "response")
  )
}

logit_external_pred <- predict_binary_from_probs(
  logit_external_probs,
  threshold = logit_thresholds
)
colnames(logit_external_pred) <- retained_skills

logit_external_metrics <- evaluate_multilabel(
  y_external,
  logit_external_pred
) |>
  mutate(model = "Logistic Regression", .before = 1)

logit_external_summary <- make_prediction_summary(
  y_external,
  logit_external_pred,
  logit_external_probs
)

save_external_model_outputs(
  "logistic_regression",
  "Logistic Regression",
  logit_external_probs,
  logit_external_pred,
  logit_external_metrics,
  logit_external_summary
)

logit_external_metrics
```

# KNN

```{r run-knn-external-validation}
knn_results <- readRDS(file.path(
  model_output_root,
  "knn",
  "knn_5k_results.rds"
))

knn_thresholds <- get_thresholds(knn_results, retained_skills)

# Reuse the KNN feature selection and scaling parameters saved from training.
top_feature_indices <- knn_results$top_feature_indices
X_train_dense_raw <- as.matrix(X_train[, top_feature_indices])
X_external_dense_raw <- as.matrix(X_external[, top_feature_indices])

X_train_dense <- scale(
  X_train_dense_raw,
  center = knn_results$center,
  scale = knn_results$scale
)
X_external_dense <- scale(
  X_external_dense_raw,
  center = knn_results$center,
  scale = knn_results$scale
)

knn_external_probs <- matrix(
  0,
  nrow = nrow(X_external_dense),
  ncol = length(retained_skills)
)
colnames(knn_external_probs) <- retained_skills

for (j in seq_along(retained_skills)) {
  y_j <- y_train_full[, retained_skills[j]]

  if (length(unique(y_j)) == 2) {
    train_labels_factor <- factor(y_j, levels = c(0, 1))

    pred_external <- knn(
      train = X_train_dense,
      test = X_external_dense,
      cl = train_labels_factor,
      k = knn_results$k_value,
      prob = TRUE
    )

    pred_external_char <- as.character(pred_external)
    pred_external_prob <- ifelse(
      pred_external_char == "1",
      attr(pred_external, "prob"),
      1 - attr(pred_external, "prob")
    )

    knn_external_probs[, j] <- pred_external_prob
  } else {
    knn_external_probs[, j] <- mean(y_j)
  }
}

knn_external_pred <- predict_binary_from_probs(
  knn_external_probs,
  threshold = knn_thresholds
)
colnames(knn_external_pred) <- retained_skills

knn_external_metrics <- evaluate_multilabel(
  y_external,
  knn_external_pred
) |>
  mutate(model = "KNN", .before = 1)

knn_external_summary <- make_prediction_summary(
  y_external,
  knn_external_pred,
  knn_external_probs
)

save_external_model_outputs(
  "knn",
  "KNN",
  knn_external_probs,
  knn_external_pred,
  knn_external_metrics,
  knn_external_summary
)

knn_external_metrics
```

# XGBoost

```{r run-xgboost-external-validation}
xgb_results <- readRDS(file.path(
  model_output_root,
  "xgboost",
  "xgboost_5k_results.rds"
))

xgb_thresholds <- get_thresholds(xgb_results, retained_skills)
xgb_external_dmatrix <- xgb.DMatrix(data = X_external)

xgb_external_probs <- matrix(
  0,
  nrow = nrow(X_external),
  ncol = length(retained_skills)
)
colnames(xgb_external_probs) <- retained_skills

for (j in seq_along(retained_skills)) {
  model_j <- xgb_results$models[[j]]
  xgb_external_probs[, j] <- predict(model_j, newdata = xgb_external_dmatrix)
}

xgb_external_pred <- predict_binary_from_probs(
  xgb_external_probs,
  threshold = xgb_thresholds
)
colnames(xgb_external_pred) <- retained_skills

xgb_external_metrics <- evaluate_multilabel(
  y_external,
  xgb_external_pred
) |>
  mutate(model = "XGBoost", .before = 1)

xgb_external_summary <- make_prediction_summary(
  y_external,
  xgb_external_pred,
  xgb_external_probs
)

save_external_model_outputs(
  "xgboost",
  "XGBoost",
  xgb_external_probs,
  xgb_external_pred,
  xgb_external_metrics,
  xgb_external_summary
)

xgb_external_metrics
```

# Compare External Validation Results

```{r compare-external-validation-results}
external_validation_results <- bind_rows(
  baseline_external_metrics,
  logit_external_metrics,
  knn_external_metrics,
  xgb_external_metrics
) |>
  arrange(desc(f1_micro))

external_prediction_summary <- bind_rows(
  baseline_external_summary |> mutate(model = "Baseline Frequency Prior", .before = 1),
  logit_external_summary |> mutate(model = "Logistic Regression", .before = 1),
  knn_external_summary |> mutate(model = "KNN", .before = 1),
  xgb_external_summary |> mutate(model = "XGBoost", .before = 1)
)

write_csv(
  external_validation_results,
  file.path(external_output_root, "external_validation_model_metrics.csv")
)
write_csv(
  external_prediction_summary,
  file.path(external_output_root, "external_validation_prediction_summary.csv")
)
saveRDS(
  list(
    external_validation_results = external_validation_results,
    external_prediction_summary = external_prediction_summary,
    retained_skills = retained_skills,
    external_profiles = external_profiles
  ),
  file.path(external_output_root, "external_validation_all_model_results.rds")
)

external_validation_results
```
