---
title: "CP8 Modeling"
author: "Lincoln Chanakira"
output: html_document
---

This notebook trains and compares several multi-label skill prediction models using the encoded TF-IDF inputs created in the previous step. The goal is to evaluate candidate models on the validation and test sets, then save the final model comparison results.

The setup chunk controls how code and messages appear in the rendered HTML output.

```{r setup, include=FALSE}
# Set global knitting options for cleaner rendered output.
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Load Packages

This block loads the packages needed for reading data, manipulating tables, handling sparse matrices, training models, and evaluating predictions.

```{r modeling-packages}
# Load packages used for data handling, sparse features, and modeling.
library(readr)
library(dplyr)
library(tibble)
library(purrr)
library(Matrix)
library(glmnet)
library(xgboost)
library(caret)
library(class)
```

# Load Encoded Inputs

This block imports the saved train, validation, and test split objects from the model input encoding step. These objects contain the TF-IDF feature matrices, multi-label target matrices, profile metadata, and retained skill names.

```{r load-encoded-data}
# Load encoded feature matrices, target matrices, profiles, and skill labels.
split_input_path <- "../data/processed/train_validation_test_split.rds"

split_objects <- readRDS(split_input_path)

X_train <- split_objects$X_train
X_validation <- split_objects$X_validation
X_test <- split_objects$X_test

y_train <- as.matrix(split_objects$y_train)
y_validation <- as.matrix(split_objects$y_validation)
y_test <- as.matrix(split_objects$y_test)

train_profiles <- split_objects$train_profiles
validation_profiles <- split_objects$validation_profiles
test_profiles <- split_objects$test_profiles
retained_skills <- split_objects$retained_skills

cat("X_train dimensions:", dim(X_train), "\n")
cat("X_validation dimensions:", dim(X_validation), "\n")
cat("X_test dimensions:", dim(X_test), "\n")
cat("y_train dimensions:", dim(y_train), "\n")
cat("y_validation dimensions:", dim(y_validation), "\n")
cat("y_test dimensions:", dim(y_test), "\n")
```

# Modeling Scope

This block limits the modeling run to the most common skill labels. The full encoded target matrix contains many labels, so this cap keeps the notebook computationally practical while still comparing the requested model types.

```{r modeling-scope}
# Select the most frequent labels so the full modeling workflow runs reliably.
max_model_labels <- 25

label_prevalence <- colMeans(y_train)
label_order <- order(label_prevalence, decreasing = TRUE)
selected_label_indices <- label_order[seq_len(min(max_model_labels, length(label_order)))]

y_train <- y_train[, selected_label_indices, drop = FALSE]
y_validation <- y_validation[, selected_label_indices, drop = FALSE]
y_test <- y_test[, selected_label_indices, drop = FALSE]
retained_skills <- colnames(y_train)

cat("Modeled labels:", ncol(y_train), "\n")
cat("Top modeled skills:\n")
print(retained_skills)
```

# Helper Functions

This block defines reusable evaluation utilities. The metrics are computed in both micro and macro form so the results capture both overall prediction quality and label-by-label performance. The helper functions also tune one threshold per skill label on the validation set instead of forcing every model to use a fixed 0.50 cutoff.

```{r evaluation-functions}
# Define metric helpers used consistently across all candidate models.
safe_divide <- function(num, den) {
  ifelse(den == 0, 0, num / den)
}

evaluate_multilabel <- function(y_true, y_pred_binary) {
  y_true <- as.matrix(y_true)
  y_pred_binary <- as.matrix(y_pred_binary)

  tp_micro <- sum((y_true == 1) & (y_pred_binary == 1))
  fp_micro <- sum((y_true == 0) & (y_pred_binary == 1))
  fn_micro <- sum((y_true == 1) & (y_pred_binary == 0))

  precision_micro <- safe_divide(tp_micro, tp_micro + fp_micro)
  recall_micro <- safe_divide(tp_micro, tp_micro + fn_micro)
  f1_micro <- safe_divide(
    2 * precision_micro * recall_micro,
    precision_micro + recall_micro
  )

  precision_by_label <- numeric(ncol(y_true))
  recall_by_label <- numeric(ncol(y_true))
  f1_by_label <- numeric(ncol(y_true))

  for (j in seq_len(ncol(y_true))) {
    tp <- sum((y_true[, j] == 1) & (y_pred_binary[, j] == 1))
    fp <- sum((y_true[, j] == 0) & (y_pred_binary[, j] == 1))
    fn <- sum((y_true[, j] == 1) & (y_pred_binary[, j] == 0))

    p <- safe_divide(tp, tp + fp)
    r <- safe_divide(tp, tp + fn)
    f <- safe_divide(2 * p * r, p + r)

    precision_by_label[j] <- p
    recall_by_label[j] <- r
    f1_by_label[j] <- f
  }

  tibble(
    precision_micro = precision_micro,
    recall_micro = recall_micro,
    f1_micro = f1_micro,
    precision_macro = mean(precision_by_label, na.rm = TRUE),
    recall_macro = mean(recall_by_label, na.rm = TRUE),
    f1_macro = mean(f1_by_label, na.rm = TRUE)
  )
}

predict_binary_from_probs <- function(prob_matrix, threshold = 0.50) {
  prob_matrix <- as.matrix(prob_matrix)
  if (length(threshold) == 1) {
    return(ifelse(prob_matrix >= threshold, 1L, 0L))
  }

  threshold_matrix <- matrix(
    rep(threshold, each = nrow(prob_matrix)),
    nrow = nrow(prob_matrix),
    byrow = FALSE
  )

  ifelse(prob_matrix >= threshold_matrix, 1L, 0L)
}

has_two_classes <- function(y) {
  length(unique(y)) == 2
}

tune_label_thresholds <- function(y_true, prob_matrix, grid = seq(0.05, 0.95, by = 0.05)) {
  y_true <- as.matrix(y_true)
  prob_matrix <- as.matrix(prob_matrix)
  thresholds <- rep(0.50, ncol(y_true))
  names(thresholds) <- colnames(y_true)

  for (j in seq_len(ncol(y_true))) {
    if (sum(y_true[, j] == 1) == 0) {
      next
    }

    scores <- sapply(grid, function(threshold) {
      pred <- ifelse(prob_matrix[, j] >= threshold, 1L, 0L)
      tp <- sum((y_true[, j] == 1) & (pred == 1))
      fp <- sum((y_true[, j] == 0) & (pred == 1))
      fn <- sum((y_true[, j] == 1) & (pred == 0))
      precision <- safe_divide(tp, tp + fp)
      recall <- safe_divide(tp, tp + fn)
      safe_divide(2 * precision * recall, precision + recall)
    })

    best_index <- which.max(scores)
    thresholds[j] <- grid[best_index]
  }

  thresholds
}

make_constant_probs <- function(n_rows, values, col_names) {
  probs <- matrix(
    rep(values, each = n_rows),
    nrow = n_rows,
    byrow = FALSE
  )
  colnames(probs) <- col_names
  probs
}
```

# Baseline Model: Frequency Prior

This baseline predicts each skill using its training-set prevalence. It provides a simple reference point so the trained models can be judged against a non-learning approach.

```{r baseline-model}
# Use training-set label prevalence as a simple baseline probability.
skill_prevalence <- colMeans(y_train)

baseline_probs_validation <- matrix(
  rep(skill_prevalence, each = nrow(y_validation)),
  nrow = nrow(y_validation),
  byrow = FALSE
)

baseline_probs_test <- matrix(
  rep(skill_prevalence, each = nrow(y_test)),
  nrow = nrow(y_test),
  byrow = FALSE
)

colnames(baseline_probs_validation) <- colnames(y_train)
colnames(baseline_probs_test) <- colnames(y_train)

baseline_thresholds <- tune_label_thresholds(y_validation, baseline_probs_validation)

baseline_pred_validation <- predict_binary_from_probs(
  baseline_probs_validation,
  threshold = baseline_thresholds
)
baseline_pred_test <- predict_binary_from_probs(
  baseline_probs_test,
  threshold = baseline_thresholds
)

baseline_validation_metrics <- evaluate_multilabel(y_validation, baseline_pred_validation)
baseline_test_metrics <- evaluate_multilabel(y_test, baseline_pred_test)

baseline_validation_metrics
baseline_test_metrics
```

# Model 1: One-vs-Rest Logistic Regression

This model trains one binary regularized logistic regression per skill label. Logistic regression is a strong baseline for TF-IDF features because it handles sparse text matrices well and is relatively interpretable. Labels with too few positive or negative training examples fall back to the training-set prevalence so the one-vs-rest loop does not fail on sparse labels.

```{r logistic-regression-model}
# Train one ridge logistic regression model for each selected skill label.
logit_models <- vector("list", ncol(y_train))
logit_validation_probs <- matrix(0, nrow = nrow(X_validation), ncol = ncol(y_train))
logit_test_probs <- matrix(0, nrow = nrow(X_test), ncol = ncol(y_train))

for (j in seq_len(ncol(y_train))) {
  y_j <- y_train[, j]
  class_counts <- table(factor(y_j, levels = c(0, 1)))
  min_class_count <- min(class_counts)

  if (has_two_classes(y_j) && min_class_count >= 2) {
    if (min_class_count >= 3) {
      model_j <- cv.glmnet(
        x = X_train,
        y = y_j,
        family = "binomial",
        alpha = 0,
        type.measure = "class",
        nfolds = min(5, min_class_count)
      )

      logit_validation_probs[, j] <- as.numeric(
        predict(model_j, newx = X_validation, s = "lambda.min", type = "response")
      )

      logit_test_probs[, j] <- as.numeric(
        predict(model_j, newx = X_test, s = "lambda.min", type = "response")
      )
    } else {
      model_j <- glmnet(
        x = X_train,
        y = y_j,
        family = "binomial",
        alpha = 0,
        lambda = 0.01
      )

      logit_validation_probs[, j] <- as.numeric(
        predict(model_j, newx = X_validation, s = 0.01, type = "response")
      )

      logit_test_probs[, j] <- as.numeric(
        predict(model_j, newx = X_test, s = 0.01, type = "response")
      )
    }

    logit_models[[j]] <- model_j
  } else {
    logit_validation_probs[, j] <- mean(y_j)
    logit_test_probs[, j] <- mean(y_j)
  }
}

colnames(logit_validation_probs) <- colnames(y_train)
colnames(logit_test_probs) <- colnames(y_train)

logit_thresholds <- tune_label_thresholds(y_validation, logit_validation_probs)

logit_pred_validation <- predict_binary_from_probs(
  logit_validation_probs,
  threshold = logit_thresholds
)
logit_pred_test <- predict_binary_from_probs(
  logit_test_probs,
  threshold = logit_thresholds
)

logit_validation_metrics <- evaluate_multilabel(y_validation, logit_pred_validation)
logit_test_metrics <- evaluate_multilabel(y_test, logit_pred_test)

logit_validation_metrics
logit_test_metrics
```

# Model 2: One-vs-Rest KNN

This model predicts skills based on nearby profiles in TF-IDF space. The feature matrix is reduced before converting to dense format because KNN is memory-intensive on high-dimensional sparse text features. The dense matrices are scaled using the training-set center and scale values so validation and test data are transformed consistently.

```{r knn-model}
# Reduce TF-IDF features before running KNN to keep runtime and memory manageable.
knn_feature_count <- min(500, ncol(X_train))
feature_variance <- Matrix::colMeans(X_train != 0)
top_feature_indices <- order(feature_variance, decreasing = TRUE)[seq_len(knn_feature_count)]

X_train_dense_raw <- as.matrix(X_train[, top_feature_indices])
X_validation_dense_raw <- as.matrix(X_validation[, top_feature_indices])
X_test_dense_raw <- as.matrix(X_test[, top_feature_indices])

knn_center <- colMeans(X_train_dense_raw)
knn_scale <- apply(X_train_dense_raw, 2, sd)
knn_scale[is.na(knn_scale) | knn_scale == 0] <- 1

X_train_dense <- scale(X_train_dense_raw, center = knn_center, scale = knn_scale)
X_validation_dense <- scale(X_validation_dense_raw, center = knn_center, scale = knn_scale)
X_test_dense <- scale(X_test_dense_raw, center = knn_center, scale = knn_scale)

k_value <- 5

knn_validation_probs <- matrix(0, nrow = nrow(X_validation_dense), ncol = ncol(y_train))
knn_test_probs <- matrix(0, nrow = nrow(X_test_dense), ncol = ncol(y_train))

for (j in seq_len(ncol(y_train))) {
  y_j <- y_train[, j]

  if (has_two_classes(y_j)) {
    train_labels_factor <- factor(y_j, levels = c(0, 1))

    pred_validation <- knn(
      train = X_train_dense,
      test = X_validation_dense,
      cl = train_labels_factor,
      k = k_value,
      prob = TRUE
    )

    pred_validation_char <- as.character(pred_validation)
    pred_validation_prob <- ifelse(
      pred_validation_char == "1",
      attr(pred_validation, "prob"),
      1 - attr(pred_validation, "prob")
    )

    knn_validation_probs[, j] <- pred_validation_prob

    pred_test <- knn(
      train = X_train_dense,
      test = X_test_dense,
      cl = train_labels_factor,
      k = k_value,
      prob = TRUE
    )

    pred_test_char <- as.character(pred_test)
    pred_test_prob <- ifelse(
      pred_test_char == "1",
      attr(pred_test, "prob"),
      1 - attr(pred_test, "prob")
    )

    knn_test_probs[, j] <- pred_test_prob
  } else {
    knn_validation_probs[, j] <- mean(y_j)
    knn_test_probs[, j] <- mean(y_j)
  }
}

colnames(knn_validation_probs) <- colnames(y_train)
colnames(knn_test_probs) <- colnames(y_train)

knn_thresholds <- tune_label_thresholds(y_validation, knn_validation_probs)

knn_pred_validation <- predict_binary_from_probs(
  knn_validation_probs,
  threshold = knn_thresholds
)
knn_pred_test <- predict_binary_from_probs(
  knn_test_probs,
  threshold = knn_thresholds
)

knn_validation_metrics <- evaluate_multilabel(y_validation, knn_pred_validation)
knn_test_metrics <- evaluate_multilabel(y_test, knn_pred_test)

knn_validation_metrics
knn_test_metrics
```

# Model 3: One-vs-Rest XGBoost

This model trains one gradient-boosted tree classifier per skill label. XGBoost can capture non-linear relationships between terms and skills, which gives it a different modeling strategy from logistic regression and KNN. Each label gets its own `scale_pos_weight` value so rare positive labels receive more weight during training.

```{r xgboost-model}
# Train one XGBoost binary classifier for each selected skill label.
xgb_models <- vector("list", ncol(y_train))
xgb_validation_probs <- matrix(0, nrow = nrow(X_validation), ncol = ncol(y_train))
xgb_test_probs <- matrix(0, nrow = nrow(X_test), ncol = ncol(y_train))

for (j in seq_len(ncol(y_train))) {
  y_j <- y_train[, j]
  positive_count <- sum(y_j == 1)
  negative_count <- sum(y_j == 0)

  if (has_two_classes(y_j) && positive_count > 0 && negative_count > 0) {
    label_scale_pos_weight <- negative_count / positive_count

    dtrain <- xgb.DMatrix(data = X_train, label = y_j)
    dvalidation <- xgb.DMatrix(data = X_validation, label = y_validation[, j])
    dtest <- xgb.DMatrix(data = X_test, label = y_test[, j])

    watchlist <- list(train = dtrain, eval = dvalidation)

    model_j <- xgb.train(
      params = list(
        objective = "binary:logistic",
        eval_metric = "logloss",
        max_depth = 4,
        eta = 0.10,
        subsample = 0.80,
        colsample_bytree = 0.80,
        scale_pos_weight = label_scale_pos_weight
      ),
      data = dtrain,
      nrounds = 50,
      watchlist = watchlist,
      early_stopping_rounds = 10,
      verbose = 0
    )

    xgb_models[[j]] <- model_j

    xgb_validation_probs[, j] <- predict(model_j, newdata = dvalidation)
    xgb_test_probs[, j] <- predict(model_j, newdata = dtest)
  } else {
    xgb_validation_probs[, j] <- mean(y_j)
    xgb_test_probs[, j] <- mean(y_j)
  }
}

colnames(xgb_validation_probs) <- colnames(y_train)
colnames(xgb_test_probs) <- colnames(y_train)

xgb_thresholds <- tune_label_thresholds(y_validation, xgb_validation_probs)

xgb_pred_validation <- predict_binary_from_probs(
  xgb_validation_probs,
  threshold = xgb_thresholds
)
xgb_pred_test <- predict_binary_from_probs(
  xgb_test_probs,
  threshold = xgb_thresholds
)

xgb_validation_metrics <- evaluate_multilabel(y_validation, xgb_pred_validation)
xgb_test_metrics <- evaluate_multilabel(y_test, xgb_pred_test)

xgb_validation_metrics
xgb_test_metrics
```

# Compare Model Performance

This block combines validation and test metrics into comparison tables. The validation set is used for model selection, while the test set provides a final held-out performance check.

```{r compare-models}
# Combine metrics from all candidate models into validation and test tables.
validation_results <- bind_rows(
  baseline_validation_metrics |> mutate(model = "Baseline Frequency Prior"),
  logit_validation_metrics |> mutate(model = "Logistic Regression"),
  knn_validation_metrics |> mutate(model = "KNN"),
  xgb_validation_metrics |> mutate(model = "XGBoost")
) |>
  select(model, everything())

test_results <- bind_rows(
  baseline_test_metrics |> mutate(model = "Baseline Frequency Prior"),
  logit_test_metrics |> mutate(model = "Logistic Regression"),
  knn_test_metrics |> mutate(model = "KNN"),
  xgb_test_metrics |> mutate(model = "XGBoost")
) |>
  select(model, everything())

validation_results
test_results
```

# Select Final Model

This block selects the model with the strongest validation F1-micro score. F1-micro is useful for this multi-label task because it balances precision and recall across all skill predictions.

```{r select-final-model}
# Choose the final candidate using validation F1-micro.
best_model_name <- validation_results |>
  arrange(desc(f1_micro)) |>
  slice(1) |>
  pull(model)

cat("Selected final model based on validation F1-micro:", best_model_name, "\n")
```

# Inspect Final Predictions

This block summarizes how often the selected model predicts each skill in the test set. Comparing predicted counts with actual counts helps reveal whether the selected model underpredicts or overpredicts common skills.

```{r inspect-predictions}
# Pull predictions from the selected model and compare predicted versus actual skill counts.
if (best_model_name == "Logistic Regression") {
  final_test_probs <- logit_test_probs
  final_test_pred <- logit_pred_test
} else if (best_model_name == "KNN") {
  final_test_probs <- knn_test_probs
  final_test_pred <- knn_pred_test
} else if (best_model_name == "XGBoost") {
  final_test_probs <- xgb_test_probs
  final_test_pred <- xgb_pred_test
} else {
  final_test_probs <- baseline_probs_test
  final_test_pred <- baseline_pred_test
}

predicted_skill_counts <- colSums(final_test_pred)
actual_skill_counts <- colSums(y_test)

prediction_summary <- tibble(
  skill = colnames(y_test),
  actual_count = actual_skill_counts,
  predicted_count = predicted_skill_counts
) |>
  arrange(desc(actual_count))

head(prediction_summary, 20)
```

# Save Modeling Results

This block saves the model comparison tables, selected model name, prediction summary, and modeled skill labels. Saving these objects makes it easier to reuse the modeling results in the final CP8 write-up.

```{r save-model-results}
# Save the final modeling outputs for reporting and downstream analysis.
model_results_output_path <- "../data/processed/cp8_model_results.rds"

model_results <- list(
  validation_results = validation_results,
  test_results = test_results,
  baseline_validation_metrics = baseline_validation_metrics,
  baseline_test_metrics = baseline_test_metrics,
  logit_validation_metrics = logit_validation_metrics,
  logit_test_metrics = logit_test_metrics,
  knn_validation_metrics = knn_validation_metrics,
  knn_test_metrics = knn_test_metrics,
  xgb_validation_metrics = xgb_validation_metrics,
  xgb_test_metrics = xgb_test_metrics,
  baseline_thresholds = baseline_thresholds,
  logit_thresholds = logit_thresholds,
  knn_thresholds = knn_thresholds,
  xgb_thresholds = xgb_thresholds,
  best_model_name = best_model_name,
  prediction_summary = prediction_summary,
  retained_skills = retained_skills
)

saveRDS(model_results, model_results_output_path)

cat("Model results saved to:", model_results_output_path, "\n")
```
