---
title: "EDA for Skill Extraction from Job Descriptions: 5k Training vs External Validation"
author: "Lincoln Chanakira"
output: html_document
---

This exploratory analysis supports the research question:

> How well can skills be extracted or predicted from LinkedIn job descriptions,
> job titles, and related profile text?

The analysis compares the two datasets used in the final CP8 modeling and
external validation workflow:

- The 5k LinkedIn dataset used for training and internal testing.
- The KSB alumni + LinkedIn external validation dataset used for out-of-sample
  validation.

The main EDA focus is:

- Dataset dimensions.
- Total number of skill assignments.
- Number of unique skills.
- Skills per profile.
- Job-description text availability.
- Skill overlap between the 5k training data and external validation data.
- Label imbalance and top-skill concentration.
- Coverage of the 100 modeled skills in the external validation dataset.
- TF-IDF vocabulary and sparsity diagnostics.
- Text-to-skill signal checks using token-skill association.
- Skill co-occurrence patterns for multi-label modeling.
- Train-vs-external validation drift diagnostics.
- Model-readiness risks and practical recommendations.

These measures show whether the 5k training data has enough label coverage for
modeling and whether the external validation dataset is comparable enough for a
fair out-of-sample check.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Load Packages and Paths

```{r packages}
library(readr)
library(dplyr)
library(tidyr)
library(stringr)
library(purrr)
library(ggplot2)
library(tidytext)
library(scales)
library(tibble)
library(forcats)
```

```{r paths}
processed_dir <- file.path("..", "data", "processed")
eda_dir <- file.path("..", "results", "skills_extraction_eda_5k_external")

dir.create(eda_dir, recursive = TRUE, showWarnings = FALSE)
```

# Helper Functions

```{r helper-functions}
# Normalize text so token and skill comparisons are consistent across datasets.
clean_text_for_eda <- function(x) {
  x <- ifelse(is.na(x), "", as.character(x))
  x <- str_to_lower(x)
  x <- str_replace_all(x, "[\r\n\t]+", " ")
  x <- str_replace_all(x, "[[:punct:]]+", " ")
  x <- str_replace_all(x, "\\s+", " ")
  str_squish(x)
}

# Convert comma/semicolon/pipe-separated skill strings into unique skill vectors.
split_skill_string <- function(x) {
  x <- ifelse(is.na(x), "", as.character(x))
  x <- str_to_lower(x)
  x <- str_replace_all(x, "\\s*[,;|]\\s*", ",")
  str_split(x, ",") |>
    map(~ {
      skills <- str_squish(.x)
      skills <- skills[skills != ""]
      unique(skills)
    })
}

count_words <- function(x) {
  str_count(ifelse(is.na(x), "", as.character(x)), "\\S+")
}

collapse_row_text <- function(df, columns) {
  if (length(columns) == 0) {
    return(rep("", nrow(df)))
  }

  df |>
    select(all_of(columns)) |>
    mutate(across(everything(), as.character)) |>
    pmap_chr(function(...) {
      values <- c(...)
      values <- values[!is.na(values) & values != ""]
      paste(values, collapse = " ")
    })
}

make_dataset_summary <- function(df) {
  df |>
    group_by(dataset) |>
    summarise(
      profiles = n(),
      profiles_with_skills = sum(skill_count > 0),
      pct_profiles_with_skills = mean(skill_count > 0) * 100,
      avg_skills_per_profile = mean(skill_count),
      median_skills_per_profile = median(skill_count),
      avg_profile_tokens = mean(profile_token_count),
      median_profile_tokens = median(profile_token_count),
      pct_profiles_with_current_description = mean(has_current_job_description) * 100,
      pct_profiles_with_previous_description = mean(has_previous_job_description) * 100,
      avg_experience_count = mean(total_experience_count, na.rm = TRUE),
      .groups = "drop"
    )
}
```

# Load and Harmonize Datasets

```{r load-5k-dataset}
# The 5k dataset is the primary training dataset used by the CP8 modeling
# workflow. It already contains cleaned profile_text and parsed skill counts.
data_5k_path <- file.path(processed_dir, "linkedin_full_5865_preprocessed.csv")
data_5k_raw <- read_csv(data_5k_path, show_col_types = FALSE)

data_5k <- data_5k_raw |>
  transmute(
    dataset = "5k training dataset",
    profile_id = linkedin_url,
    profile_text = clean_text_for_eda(profile_text),
    current_job_title = current_job_title,
    current_job_description = current_job_description,
    previous_job_title = previous_job_title,
    previous_job_description = previous_job_description,
    skills = skills,
    skill_list = split_skill_string(skills),
    skill_count = lengths(skill_list),
    has_current_job_description = has_current_job_description,
    has_previous_job_description = has_previous_job_description,
    total_experience_count = total_experience_count,
    source_role = "model training and internal validation"
  ) |>
  mutate(profile_token_count = count_words(profile_text))

cat("5k dataset shape:", dim(data_5k), "\n")
glimpse(data_5k)
```

```{r load-external-validation-dataset}
# The external validation dataset is the KSB alumni + LinkedIn file after CP8
# preparation and preprocessing. Rows with empty Skills were removed upstream.
external_path <- file.path(processed_dir, "ksb_alumni_linkedin_preprocessed.csv")
external_raw <- read_csv(external_path, show_col_types = FALSE)

external_validation <- external_raw |>
  transmute(
    dataset = "KSB alumni external validation",
    profile_id = linkedin_url,
    profile_text = clean_text_for_eda(profile_text),
    current_job_title = current_job_title,
    current_job_description = current_job_description,
    previous_job_title = previous_job_title,
    previous_job_description = previous_job_description,
    skills = skills,
    skill_list = split_skill_string(skills),
    skill_count = lengths(skill_list),
    has_current_job_description = has_current_job_description,
    has_previous_job_description = has_previous_job_description,
    total_experience_count = total_experience_count,
    source_role = "external validation"
  ) |>
  mutate(profile_token_count = count_words(profile_text))

cat("External validation shape:", dim(external_validation), "\n")
glimpse(external_validation)
```

```{r combine-datasets}
combined_profiles <- bind_rows(data_5k, external_validation)

combined_profiles |>
  count(dataset, source_role)
```

# Core Dataset and Skill Inventory

This section is the main dataset-level EDA. It compares the size of the 5k
training dataset and the external validation dataset, then summarizes the skill
labels that the model is trying to predict.

```{r dataset-dimensions}
dataset_dimensions <- tibble(
  dataset = c("5k training dataset", "KSB alumni external validation"),
  raw_rows = c(nrow(data_5k_raw), nrow(external_raw)),
  raw_columns = c(ncol(data_5k_raw), ncol(external_raw)),
  analysis_rows = c(nrow(data_5k), nrow(external_validation)),
  analysis_columns = c(ncol(data_5k), ncol(external_validation))
)

dataset_dimensions

write_csv(dataset_dimensions, file.path(eda_dir, "dataset_dimensions.csv"))
```

```{r skill-long-table}
skills_long <- combined_profiles |>
  select(dataset, profile_id, skill_list) |>
  unnest(skill_list) |>
  rename(skill = skill_list) |>
  filter(!is.na(skill), skill != "")
```

```{r skill-inventory}
skill_inventory <- skills_long |>
  group_by(dataset) |>
  summarise(
    total_number_of_skills = n(),
    unique_skills = n_distinct(skill),
    profiles_with_skills = n_distinct(profile_id),
    avg_profiles_per_skill = total_number_of_skills / unique_skills,
    .groups = "drop"
  )

skill_inventory

write_csv(skill_inventory, file.path(eda_dir, "skill_inventory_summary.csv"))
```

```{r skills-per-profile-summary}
skills_per_profile_summary <- combined_profiles |>
  group_by(dataset) |>
  summarise(
    profiles = n(),
    profiles_with_zero_skills = sum(skill_count == 0),
    min_skills_per_profile = min(skill_count),
    p25_skills_per_profile = quantile(skill_count, 0.25),
    median_skills_per_profile = median(skill_count),
    mean_skills_per_profile = mean(skill_count),
    p75_skills_per_profile = quantile(skill_count, 0.75),
    max_skills_per_profile = max(skill_count),
    .groups = "drop"
  )

skills_per_profile_summary

write_csv(
  skills_per_profile_summary,
  file.path(eda_dir, "skills_per_profile_summary.csv")
)
```

```{r skill-inventory-plot, fig.width=10, fig.height=6}
skill_inventory_plot <- skill_inventory |>
  select(dataset, total_number_of_skills, unique_skills) |>
  pivot_longer(
    cols = c(total_number_of_skills, unique_skills),
    names_to = "metric",
    values_to = "count"
  ) |>
  mutate(
    metric = recode(
      metric,
      total_number_of_skills = "Total skill assignments",
      unique_skills = "Unique skills"
    )
  ) |>
  ggplot(aes(x = dataset, y = count, fill = dataset)) +
  geom_col(show.legend = FALSE, width = 0.65) +
  geom_text(aes(label = comma(count)), vjust = -0.25, size = 3.5) +
  facet_wrap(~ metric, scales = "free_y") +
  labs(
    title = "Total and Unique Skill Labels by Dataset",
    subtitle = "Total skills count all profile-skill assignments; unique skills count distinct labels",
    x = NULL,
    y = "Count"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))

skill_inventory_plot

ggsave(
  file.path(eda_dir, "skill_inventory_plot.png"),
  skill_inventory_plot,
  width = 10,
  height = 6
)
```

```{r skills-per-profile-plot, fig.width=11, fig.height=6}
skills_per_profile_plot <- combined_profiles |>
  ggplot(aes(x = skill_count, fill = dataset)) +
  geom_histogram(binwidth = 1, alpha = 0.8, boundary = 0, show.legend = FALSE) +
  facet_wrap(~ dataset, scales = "free_y") +
  scale_x_continuous(breaks = pretty_breaks()) +
  labs(
    title = "Skills per Profile",
    subtitle = "Each row can have multiple skill labels, so this is a multi-label prediction problem",
    x = "Number of skills listed on a profile",
    y = "Profiles"
  ) +
  theme_minimal()

skills_per_profile_plot

ggsave(
  file.path(eda_dir, "skills_per_profile_plot.png"),
  skills_per_profile_plot,
  width = 11,
  height = 6
)
```

# Other Important EDA Metrics

These additional metrics connect the dataset summaries back to the modeling
problem. A skill extraction model needs enough text signal, enough repeated
labels to learn from, and enough overlap between the training label space and
the external validation label space.

```{r text-signal-summary}
text_signal_summary <- combined_profiles |>
  group_by(dataset) |>
  summarise(
    profiles = n(),
    profiles_with_profile_text = sum(profile_token_count > 0),
    pct_profiles_with_profile_text = mean(profile_token_count > 0) * 100,
    profiles_with_current_job_description = sum(has_current_job_description == 1),
    pct_profiles_with_current_job_description = mean(has_current_job_description == 1) * 100,
    profiles_with_previous_job_description = sum(has_previous_job_description == 1),
    pct_profiles_with_previous_job_description = mean(has_previous_job_description == 1) * 100,
    total_profile_text_tokens = sum(profile_token_count),
    mean_profile_text_tokens = mean(profile_token_count),
    median_profile_text_tokens = median(profile_token_count),
    .groups = "drop"
  )

text_signal_summary

write_csv(text_signal_summary, file.path(eda_dir, "text_signal_summary.csv"))
```

```{r label-density-summary}
top_20_skill_names <- skills_long |>
  group_by(dataset) |>
  count(skill, sort = TRUE) |>
  slice_max(order_by = n, n = 20, with_ties = FALSE) |>
  ungroup()

top_20_skill_assignments <- skills_long |>
  semi_join(top_20_skill_names, by = c("dataset", "skill")) |>
  count(dataset, name = "assignments_in_top_20_skills")

label_density_summary <- skill_inventory |>
  left_join(top_20_skill_assignments, by = "dataset") |>
  mutate(
    possible_profile_skill_pairs = profiles_with_skills * unique_skills,
    label_density_pct = total_number_of_skills / possible_profile_skill_pairs * 100,
    pct_assignments_in_top_20_skills = assignments_in_top_20_skills / total_number_of_skills * 100
  ) |>
  select(
    dataset,
    total_number_of_skills,
    unique_skills,
    possible_profile_skill_pairs,
    label_density_pct,
    assignments_in_top_20_skills,
    pct_assignments_in_top_20_skills
  )

label_density_summary

write_csv(label_density_summary, file.path(eda_dir, "label_density_summary.csv"))
```

```{r skill-overlap-summary}
training_skills <- skills_long |>
  filter(dataset == "5k training dataset") |>
  distinct(skill) |>
  pull(skill)

external_skills <- skills_long |>
  filter(dataset == "KSB alumni external validation") |>
  distinct(skill) |>
  pull(skill)

skill_overlap_summary <- tibble(
  training_unique_skills = length(training_skills),
  external_unique_skills = length(external_skills),
  shared_unique_skills = length(intersect(training_skills, external_skills)),
  external_unique_skills_seen_in_training = sum(external_skills %in% training_skills),
  external_unique_skills_unseen_in_training = sum(!external_skills %in% training_skills),
  pct_external_unique_skills_seen_in_training = mean(external_skills %in% training_skills) * 100
)

skill_overlap_summary

write_csv(skill_overlap_summary, file.path(eda_dir, "skill_overlap_summary.csv"))
```

```{r skill-overlap-plot, fig.width=9, fig.height=5}
skill_overlap_plot_data <- tibble(
  overlap_group = c("External skills seen in 5k training", "External skills not seen in 5k training"),
  unique_skills = c(
    skill_overlap_summary$external_unique_skills_seen_in_training,
    skill_overlap_summary$external_unique_skills_unseen_in_training
  )
)

skill_overlap_plot <- skill_overlap_plot_data |>
  ggplot(aes(x = overlap_group, y = unique_skills, fill = overlap_group)) +
  geom_col(show.legend = FALSE, width = 0.65) +
  geom_text(aes(label = comma(unique_skills)), vjust = -0.25, size = 3.5) +
  labs(
    title = "External Skill Overlap with 5k Training Labels",
    subtitle = "External skills outside the training label space cannot be predicted directly",
    x = NULL,
    y = "Unique external skills"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 10, hjust = 1))

skill_overlap_plot

ggsave(
  file.path(eda_dir, "skill_overlap_plot.png"),
  skill_overlap_plot,
  width = 9,
  height = 5
)
```

```{r modeled-skill-readiness-summary}
external_model_metrics_path <- file.path(
  processed_dir,
  "external_validation_model_outputs",
  "external_validation_all_model_results.rds"
)

if (file.exists(external_model_metrics_path)) {
  external_model_results <- readRDS(external_model_metrics_path)
  modeled_skills <- external_model_results$retained_skills
} else {
  split_objects <- readRDS(file.path(processed_dir, "train_validation_test_split.rds"))
  y_train <- as.matrix(split_objects$y_train)
  label_order <- order(colMeans(y_train), decreasing = TRUE)
  modeled_skills <- colnames(y_train)[label_order[seq_len(min(100, ncol(y_train)))]]
}

modeled_skill_readiness_summary <- skills_long |>
  filter(dataset == "KSB alumni external validation") |>
  mutate(in_modeled_skill_space = skill %in% modeled_skills) |>
  summarise(
    modeled_skill_count = length(modeled_skills),
    external_skill_assignments = n(),
    external_assignments_in_modeled_skill_space = sum(in_modeled_skill_space),
    pct_external_assignments_in_modeled_skill_space = mean(in_modeled_skill_space) * 100,
    external_unique_skills = n_distinct(skill),
    external_unique_skills_in_modeled_skill_space = n_distinct(skill[in_modeled_skill_space]),
    pct_external_unique_skills_in_modeled_skill_space =
      external_unique_skills_in_modeled_skill_space / external_unique_skills * 100
  )

modeled_skill_readiness_summary

write_csv(
  modeled_skill_readiness_summary,
  file.path(eda_dir, "modeled_skill_readiness_summary.csv")
)
```

# Dataset Readiness for Skill Extraction

```{r dataset-summary}
dataset_summary <- make_dataset_summary(combined_profiles)
dataset_summary

write_csv(dataset_summary, file.path(eda_dir, "dataset_readiness_summary.csv"))
```

```{r dataset-summary-plot, fig.width=11, fig.height=6}
readiness_long <- dataset_summary |>
  select(
    dataset,
    pct_profiles_with_skills,
    pct_profiles_with_current_description,
    pct_profiles_with_previous_description
  ) |>
  pivot_longer(
    cols = -dataset,
    names_to = "metric",
    values_to = "percent"
  ) |>
  mutate(
    metric = recode(
      metric,
      pct_profiles_with_skills = "Has skill labels",
      pct_profiles_with_current_description = "Has current job description",
      pct_profiles_with_previous_description = "Has previous job description"
    )
  )

coverage_plot <- readiness_long |>
  ggplot(aes(x = dataset, y = percent, fill = metric)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.7) +
  geom_text(
    aes(label = paste0(round(percent, 1), "%")),
    position = position_dodge(width = 0.75),
    vjust = -0.25,
    size = 3
  ) +
  scale_y_continuous(labels = label_percent(scale = 1), limits = c(0, 105)) +
  labs(
    title = "Readiness for Skill Extraction",
    subtitle = "Skill labels define targets; job descriptions provide the text signal",
    x = NULL,
    y = "Profile coverage",
    fill = NULL
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))

coverage_plot

ggsave(
  file.path(eda_dir, "readiness_coverage_plot.png"),
  coverage_plot,
  width = 11,
  height = 6
)
```

# Text Signal: How Much Job Description Text Exists?

```{r token-summary}
token_summary <- combined_profiles |>
  group_by(dataset) |>
  summarise(
    min_tokens = min(profile_token_count),
    p25_tokens = quantile(profile_token_count, 0.25),
    median_tokens = median(profile_token_count),
    mean_tokens = mean(profile_token_count),
    p75_tokens = quantile(profile_token_count, 0.75),
    max_tokens = max(profile_token_count),
    .groups = "drop"
  )

token_summary

write_csv(token_summary, file.path(eda_dir, "profile_text_token_summary.csv"))
```

```{r token-distribution, fig.width=11, fig.height=6}
token_plot <- combined_profiles |>
  ggplot(aes(x = profile_token_count, fill = dataset)) +
  geom_histogram(bins = 45, alpha = 0.75, show.legend = FALSE) +
  facet_wrap(~ dataset, scales = "free_y") +
  labs(
    title = "Profile Text Length Distribution",
    subtitle = "Longer text gives TF-IDF models more evidence for skill prediction",
    x = "Tokens in combined profile text",
    y = "Profiles"
  ) +
  theme_minimal()

token_plot

ggsave(
  file.path(eda_dir, "profile_text_token_distribution.png"),
  token_plot,
  width = 11,
  height = 6
)
```

# Skill Label Coverage and Imbalance

```{r skill-coverage}
skill_coverage <- skill_inventory |>
  rename(total_skill_assignments = total_number_of_skills)

skill_coverage

write_csv(skill_coverage, file.path(eda_dir, "skill_coverage_summary.csv"))
```

```{r top-skills, fig.width=12, fig.height=8}
top_skills <- skills_long |>
  count(dataset, skill, sort = TRUE) |>
  group_by(dataset) |>
  slice_max(order_by = n, n = 20, with_ties = FALSE) |>
  ungroup()

top_skills_plot <- top_skills |>
  mutate(skill = reorder_within(skill, n, dataset)) |>
  ggplot(aes(x = skill, y = n, fill = dataset)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  facet_wrap(~ dataset, scales = "free_y") +
  scale_x_reordered() +
  labs(
    title = "Top 20 Skill Labels by Dataset",
    subtitle = "The most common labels dominate multi-label model performance",
    x = NULL,
    y = "Profiles with skill"
  ) +
  theme_minimal()

top_skills_plot

write_csv(top_skills, file.path(eda_dir, "top_20_skills_by_dataset.csv"))
ggsave(
  file.path(eda_dir, "top_20_skills_by_dataset.png"),
  top_skills_plot,
  width = 12,
  height = 8
)
```

```{r long-tail-skills, fig.width=11, fig.height=6}
skill_frequency_rank <- skills_long |>
  count(dataset, skill, sort = TRUE) |>
  group_by(dataset) |>
  arrange(desc(n), .by_group = TRUE) |>
  mutate(skill_rank = row_number()) |>
  ungroup()

long_tail_plot <- skill_frequency_rank |>
  ggplot(aes(x = skill_rank, y = n, color = dataset)) +
  geom_line(linewidth = 0.9) +
  scale_x_log10(labels = comma) +
  scale_y_log10(labels = comma) +
  labs(
    title = "Skill Label Long Tail",
    subtitle = "A few skills are common; many skills appear rarely and are harder to learn",
    x = "Skill rank, log scale",
    y = "Profiles with skill, log scale",
    color = NULL
  ) +
  theme_minimal()

long_tail_plot

ggsave(
  file.path(eda_dir, "skill_label_long_tail.png"),
  long_tail_plot,
  width = 11,
  height = 6
)
```

# Text Terms Most Available for Modeling

```{r top-profile-terms}
top_terms <- combined_profiles |>
  select(dataset, profile_id, profile_text) |>
  unnest_tokens(term, profile_text) |>
  anti_join(stop_words, by = c("term" = "word")) |>
  filter(str_detect(term, "[a-z]"), nchar(term) > 2) |>
  count(dataset, term, sort = TRUE) |>
  group_by(dataset) |>
  slice_max(order_by = n, n = 20, with_ties = FALSE) |>
  ungroup()

top_terms

write_csv(top_terms, file.path(eda_dir, "top_profile_text_terms.csv"))
```

```{r top-profile-terms-plot, fig.width=12, fig.height=8}
top_terms_plot <- top_terms |>
  mutate(term = reorder_within(term, n, dataset)) |>
  ggplot(aes(x = term, y = n, fill = dataset)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  facet_wrap(~ dataset, scales = "free_y") +
  scale_x_reordered() +
  labs(
    title = "Most Frequent Profile Text Terms",
    subtitle = "These terms are the strongest raw text signals available before TF-IDF weighting",
    x = NULL,
    y = "Term count"
  ) +
  theme_minimal()

top_terms_plot

ggsave(
  file.path(eda_dir, "top_profile_text_terms.png"),
  top_terms_plot,
  width = 12,
  height = 8
)
```

# Do Skill Labels Appear in Job Description Text?

```{r explicit-skill-mentions}
# This checks whether the skill label itself appears in the profile text. The
# model can still learn implicit signals, but explicit mentions are easier.
skill_mentions <- skills_long |>
  left_join(
    combined_profiles |> select(dataset, profile_id, profile_text),
    by = c("dataset", "profile_id")
  ) |>
  mutate(skill_explicitly_mentioned = str_detect(profile_text, fixed(skill))) |>
  group_by(dataset) |>
  summarise(
    skill_assignments = n(),
    explicit_skill_mentions = sum(skill_explicitly_mentioned, na.rm = TRUE),
    pct_explicit_mentions = explicit_skill_mentions / skill_assignments * 100,
    .groups = "drop"
  )

skill_mentions

write_csv(skill_mentions, file.path(eda_dir, "explicit_skill_mentions_summary.csv"))
```

```{r explicit-skill-mentions-plot, fig.width=10, fig.height=5}
skill_mentions_plot <- skill_mentions |>
  ggplot(aes(x = dataset, y = pct_explicit_mentions, fill = dataset)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = paste0(round(pct_explicit_mentions, 1), "%")), vjust = -0.25) +
  scale_y_continuous(labels = label_percent(scale = 1), limits = c(0, 105)) +
  labs(
    title = "How Often Are Skill Labels Explicitly Mentioned in Profile Text?",
    subtitle = "Lower percentages mean models must infer skills from surrounding job language",
    x = NULL,
    y = "Skill assignments explicitly mentioned"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))

skill_mentions_plot

ggsave(
  file.path(eda_dir, "explicit_skill_mentions_plot.png"),
  skill_mentions_plot,
  width = 10,
  height = 5
)
```

# External Validation Drift

```{r external-drift}
training_skills <- skills_long |>
  filter(dataset == "5k training dataset") |>
  distinct(skill) |>
  pull(skill)

external_skills <- skills_long |>
  filter(dataset == "KSB alumni external validation") |>
  distinct(skill) |>
  pull(skill)

external_overlap <- tibble(
  external_unique_skills = length(external_skills),
  external_skills_seen_in_training = sum(external_skills %in% training_skills),
  external_skills_unseen_in_training = sum(!external_skills %in% training_skills),
  pct_external_skills_seen_in_training = mean(external_skills %in% training_skills) * 100
)

external_overlap

write_csv(external_overlap, file.path(eda_dir, "external_skill_overlap_with_training.csv"))
```

```{r modeled-skill-coverage}
# The current external validation modeling file evaluates the 100 modeled labels.
# This table shows how many of the external validation skill assignments fall
# inside that model label space.
external_model_metrics_path <- file.path(
  processed_dir,
  "external_validation_model_outputs",
  "external_validation_all_model_results.rds"
)

if (file.exists(external_model_metrics_path)) {
  external_model_results <- readRDS(external_model_metrics_path)
  modeled_skills <- external_model_results$retained_skills
} else {
  split_objects <- readRDS(file.path(processed_dir, "train_validation_test_split.rds"))
  y_train <- as.matrix(split_objects$y_train)
  label_order <- order(colMeans(y_train), decreasing = TRUE)
  modeled_skills <- colnames(y_train)[label_order[seq_len(min(100, ncol(y_train)))]]
}

modeled_skill_coverage <- skills_long |>
  filter(dataset == "KSB alumni external validation") |>
  mutate(in_modeled_skill_space = skill %in% modeled_skills) |>
  summarise(
    external_skill_assignments = n(),
    assignments_in_100_skill_model_space = sum(in_modeled_skill_space),
    pct_assignments_in_100_skill_model_space = mean(in_modeled_skill_space) * 100,
    unique_external_skills = n_distinct(skill),
    unique_external_skills_in_100_skill_model_space = n_distinct(skill[in_modeled_skill_space])
  )

modeled_skill_coverage

write_csv(modeled_skill_coverage, file.path(eda_dir, "external_modeled_skill_coverage.csv"))
```

# Relationship Between Text Length and Skill Labels

```{r text-length-vs-skills, fig.width=11, fig.height=6}
text_skill_plot <- combined_profiles |>
  ggplot(aes(x = profile_token_count, y = skill_count, color = dataset)) +
  geom_point(alpha = 0.35) +
  geom_smooth(method = "loess", se = FALSE) +
  facet_wrap(~ dataset, scales = "free") +
  labs(
    title = "Do Longer Job Profiles Have More Skill Labels?",
    subtitle = "More profile text can expose more skills, but also introduces noise",
    x = "Profile text tokens",
    y = "Skill labels per profile",
    color = NULL
  ) +
  theme_minimal()

text_skill_plot

ggsave(
  file.path(eda_dir, "text_length_vs_skill_count.png"),
  text_skill_plot,
  width = 11,
  height = 6
)
```

# Model-Aware EDA: TF-IDF Vocabulary and Sparsity

**Non-technical explanation.** This section checks whether the profile text has
enough repeated language for a model to learn from. A model cannot learn well if
nearly every word appears only once, or if most profiles share very few terms.

**Technical explanation.** Logistic Regression, KNN, and XGBoost use the TF-IDF
feature space created from `profile_text`. Text models usually create a very
sparse document-term matrix: most profiles contain only a small fraction of all
vocabulary terms. This is normal, but extreme sparsity can weaken KNN distance
calculations and make rare-term patterns harder for tree-based models to learn.

```{r tfidf-vocabulary-sparsity}
profile_terms <- combined_profiles |>
  select(dataset, profile_id, profile_text) |>
  unnest_tokens(term, profile_text) |>
  anti_join(stop_words, by = c("term" = "word")) |>
  filter(str_detect(term, "[a-z]"), nchar(term) > 2)

profile_term_counts <- profile_terms |>
  count(dataset, profile_id, term, name = "term_count")

vocabulary_summary <- profile_term_counts |>
  group_by(dataset) |>
  summarise(
    profiles = n_distinct(profile_id),
    vocabulary_size = n_distinct(term),
    observed_profile_term_pairs = n(),
    possible_profile_term_pairs = profiles * vocabulary_size,
    tfidf_space_sparsity_pct =
      (1 - observed_profile_term_pairs / possible_profile_term_pairs) * 100,
    avg_unique_terms_per_profile = observed_profile_term_pairs / profiles,
    .groups = "drop"
  )

vocabulary_summary

write_csv(
  vocabulary_summary,
  file.path(eda_dir, "tfidf_vocabulary_sparsity_summary.csv")
)
```

```{r term-document-frequency, fig.width=11, fig.height=6}
term_document_frequency <- profile_term_counts |>
  count(dataset, term, name = "document_frequency") |>
  group_by(dataset) |>
  mutate(term_rank = row_number(desc(document_frequency))) |>
  ungroup()

term_document_frequency_plot <- term_document_frequency |>
  ggplot(aes(x = term_rank, y = document_frequency, color = dataset)) +
  geom_line(linewidth = 0.8) +
  scale_x_log10(labels = comma) +
  scale_y_log10(labels = comma) +
  labs(
    title = "Document Frequency of Profile Text Terms",
    subtitle = "A steep long tail means many terms appear in very few profiles",
    x = "Term rank, log scale",
    y = "Profiles containing term, log scale",
    color = NULL
  ) +
  theme_minimal()

term_document_frequency_plot

ggsave(
  file.path(eda_dir, "term_document_frequency_long_tail.png"),
  term_document_frequency_plot,
  width = 11,
  height = 6
)
```

**Interpretation for modeling.** High sparsity is expected in NLP, but it means
KNN may struggle because most profiles are far apart in the feature space.
Logistic Regression is usually more stable for sparse TF-IDF features. XGBoost
may still perform well, but it needs careful threshold tuning because it can
learn common-skill patterns more easily than rare-skill patterns.

# Profiles with No Skill Labels

**Non-technical explanation.** Profiles with no skill labels are not useless.
They teach the model what a profile with no observable modeled skills looks like.
Removing them can make the model too aggressive and cause it to predict skills
too often.

**Technical explanation.** In a multi-label target matrix, profiles with zero
skills become all-zero target rows. These rows are important negative examples.
Removing them changes the base rate of positive labels, inflates label density,
and can bias precision/recall estimates.

```{r zero-skill-profile-diagnostics}
zero_skill_profile_summary <- combined_profiles |>
  mutate(has_any_skill = skill_count > 0) |>
  group_by(dataset, has_any_skill) |>
  summarise(
    profiles = n(),
    mean_profile_tokens = mean(profile_token_count),
    median_profile_tokens = median(profile_token_count),
    pct_with_current_job_description = mean(has_current_job_description == 1) * 100,
    pct_with_previous_job_description = mean(has_previous_job_description == 1) * 100,
    .groups = "drop"
  ) |>
  mutate(skill_label_status = if_else(has_any_skill, "Has at least one skill", "Has zero skills")) |>
  select(dataset, skill_label_status, everything(), -has_any_skill)

zero_skill_profile_summary

write_csv(
  zero_skill_profile_summary,
  file.path(eda_dir, "zero_skill_profile_diagnostics.csv")
)
```

```{r zero-skill-profile-plot, fig.width=10, fig.height=5}
zero_skill_profile_plot <- combined_profiles |>
  mutate(skill_label_status = if_else(skill_count > 0, "Has at least one skill", "Has zero skills")) |>
  count(dataset, skill_label_status) |>
  group_by(dataset) |>
  mutate(percent = n / sum(n) * 100) |>
  ungroup() |>
  ggplot(aes(x = dataset, y = percent, fill = skill_label_status)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.7) +
  geom_text(
    aes(label = paste0(round(percent, 1), "%")),
    position = position_dodge(width = 0.75),
    vjust = -0.25,
    size = 3
  ) +
  scale_y_continuous(labels = label_percent(scale = 1), limits = c(0, 105)) +
  labs(
    title = "Profiles With and Without Skill Labels",
    subtitle = "Zero-skill profiles provide important negative examples for the model",
    x = NULL,
    y = "Percent of profiles",
    fill = NULL
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))

zero_skill_profile_plot

ggsave(
  file.path(eda_dir, "zero_skill_profile_plot.png"),
  zero_skill_profile_plot,
  width = 10,
  height = 5
)
```

# Text-to-Skill Signal Strength

**Non-technical explanation.** This section checks whether certain words in the
profile text are meaningfully connected to specific skills. For example, if
profiles labeled with `python` also contain words like `data`, `analysis`,
`automation`, or `machine`, that gives the model a real signal.

**Technical explanation.** The table below estimates token-skill association
using a simple lift ratio. Lift compares how often a term appears among profiles
with a given skill versus how often it appears overall. A lift greater than 1
means the term is more common among profiles with that skill than in the full
dataset.

```{r token-skill-association}
# Focus on modeled skills so the signal check matches the downstream model label space.
modeled_skills_available <- modeled_skills[modeled_skills %in% training_skills]
selected_signal_skills <- modeled_skills_available[seq_len(min(15, length(modeled_skills_available)))]

training_profiles <- combined_profiles |>
  filter(dataset == "5k training dataset") |>
  select(dataset, profile_id, profile_text)

training_profile_terms <- training_profiles |>
  unnest_tokens(term, profile_text) |>
  anti_join(stop_words, by = c("term" = "word")) |>
  filter(str_detect(term, "[a-z]"), nchar(term) > 2) |>
  distinct(profile_id, term)

training_skill_profile_map <- skills_long |>
  filter(dataset == "5k training dataset", skill %in% selected_signal_skills) |>
  distinct(profile_id, skill)

training_profile_count <- n_distinct(training_profiles$profile_id)

term_base_rates <- training_profile_terms |>
  count(term, name = "profiles_with_term") |>
  mutate(base_rate = profiles_with_term / training_profile_count)

skill_profile_counts <- training_skill_profile_map |>
  count(skill, name = "profiles_with_skill")

token_skill_association <- training_skill_profile_map |>
  inner_join(training_profile_terms, by = "profile_id") |>
  count(skill, term, name = "profiles_with_skill_and_term") |>
  left_join(skill_profile_counts, by = "skill") |>
  left_join(term_base_rates, by = "term") |>
  mutate(
    term_rate_within_skill = profiles_with_skill_and_term / profiles_with_skill,
    lift = term_rate_within_skill / base_rate
  ) |>
  filter(profiles_with_skill_and_term >= 3, profiles_with_term >= 5) |>
  group_by(skill) |>
  slice_max(order_by = lift, n = 8, with_ties = FALSE) |>
  arrange(skill, desc(lift)) |>
  ungroup()

token_skill_association

write_csv(
  token_skill_association,
  file.path(eda_dir, "token_skill_association_lift.csv")
)
```

```{r token-skill-association-plot, fig.width=13, fig.height=9}
if (nrow(token_skill_association) > 0) {
  token_skill_association_plot <- token_skill_association |>
    mutate(term = reorder_within(term, lift, skill)) |>
    ggplot(aes(x = term, y = lift, fill = skill)) +
    geom_col(show.legend = FALSE) +
    coord_flip() +
    facet_wrap(~ skill, scales = "free_y") +
    scale_x_reordered() +
    labs(
      title = "Strongest Text Terms Associated with Selected Skills",
      subtitle = "Lift greater than 1 means the term is more common among profiles with that skill",
      x = NULL,
      y = "Association lift"
    ) +
    theme_minimal()

  token_skill_association_plot

  ggsave(
    file.path(eda_dir, "token_skill_association_lift.png"),
    token_skill_association_plot,
    width = 13,
    height = 9
  )
}
```

**Interpretation for modeling.** If the lift table contains intuitive terms for
major skills, that supports the use of TF-IDF models. If the terms look generic
or unrelated, then the model may be learning broad profile style rather than
actual skill evidence.

# Multi-Label Skill Co-Occurrence

**Non-technical explanation.** Skills often appear together. A profile with
`data analysis` may also list `excel`, `sql`, or `python`. Understanding these
combinations helps explain why the task is multi-label rather than single-label.

**Technical explanation.** This section identifies frequent skill pairs within
the same profile. Co-occurrence can support interpretation and model diagnostics,
although the current one-vs-rest model setup predicts each skill independently.

```{r skill-cooccurrence}
cooccurrence_skills <- skills_long |>
  filter(dataset == "5k training dataset", skill %in% modeled_skills) |>
  distinct(profile_id, skill)

skill_pairs <- cooccurrence_skills |>
  inner_join(cooccurrence_skills, by = "profile_id", suffix = c("_1", "_2")) |>
  filter(skill_1 < skill_2) |>
  count(skill_1, skill_2, name = "profiles_with_both") |>
  arrange(desc(profiles_with_both))

top_skill_pairs <- skill_pairs |>
  slice_max(order_by = profiles_with_both, n = 25, with_ties = FALSE)

top_skill_pairs

write_csv(top_skill_pairs, file.path(eda_dir, "top_skill_cooccurrence_pairs.csv"))
```

```{r skill-cooccurrence-plot, fig.width=12, fig.height=8}
if (nrow(top_skill_pairs) > 0) {
  skill_pair_plot <- top_skill_pairs |>
    mutate(skill_pair = paste(skill_1, "+", skill_2)) |>
    mutate(skill_pair = fct_reorder(skill_pair, profiles_with_both)) |>
    ggplot(aes(x = skill_pair, y = profiles_with_both)) +
    geom_col() +
    coord_flip() +
    labs(
      title = "Most Common Skill Pairs in the 5k Training Data",
      subtitle = "Frequently paired skills show the multi-label structure of the problem",
      x = NULL,
      y = "Profiles containing both skills"
    ) +
    theme_minimal()

  skill_pair_plot

  ggsave(
    file.path(eda_dir, "top_skill_cooccurrence_pairs.png"),
    skill_pair_plot,
    width = 12,
    height = 8
  )
}
```

# Train vs External Validation Distribution Shift

**Non-technical explanation.** External validation is only fair if the external
dataset is reasonably comparable to the training data. If the alumni dataset uses
different language, has shorter or longer text, or contains very different
skills, model performance may drop even if the model was trained correctly.

**Technical explanation.** This section compares profile-text length and
skill-label frequency between the 5k training data and the external validation
set. The Jensen-Shannon divergence score below is a bounded
distribution-distance measure. Values closer to 0 indicate more similar
distributions; larger values indicate stronger drift.

```{r text-length-drift}
text_length_drift <- combined_profiles |>
  mutate(
    token_bucket = case_when(
      profile_token_count == 0 ~ "0 tokens",
      profile_token_count <= 25 ~ "1-25 tokens",
      profile_token_count <= 75 ~ "26-75 tokens",
      profile_token_count <= 150 ~ "76-150 tokens",
      profile_token_count <= 300 ~ "151-300 tokens",
      TRUE ~ "301+ tokens"
    ),
    token_bucket = factor(
      token_bucket,
      levels = c("0 tokens", "1-25 tokens", "26-75 tokens", "76-150 tokens", "151-300 tokens", "301+ tokens")
    )
  ) |>
  count(dataset, token_bucket) |>
  group_by(dataset) |>
  mutate(percent = n / sum(n) * 100) |>
  ungroup()

text_length_drift

write_csv(text_length_drift, file.path(eda_dir, "text_length_distribution_drift.csv"))
```

```{r text-length-drift-plot, fig.width=11, fig.height=6}
text_length_drift_plot <- text_length_drift |>
  ggplot(aes(x = token_bucket, y = percent, fill = dataset)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.7) +
  scale_y_continuous(labels = label_percent(scale = 1)) +
  labs(
    title = "Training vs External Profile Text Length",
    subtitle = "Large differences may indicate external validation drift",
    x = "Profile text length bucket",
    y = "Percent of profiles",
    fill = NULL
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))

text_length_drift_plot

ggsave(
  file.path(eda_dir, "text_length_distribution_drift.png"),
  text_length_drift_plot,
  width = 11,
  height = 6
)
```

```{r skill-frequency-drift-js}
js_divergence <- function(p, q) {
  eps <- 1e-12
  p <- p + eps
  q <- q + eps
  p <- p / sum(p)
  q <- q / sum(q)
  m <- 0.5 * (p + q)
  0.5 * sum(p * log(p / m)) + 0.5 * sum(q * log(q / m))
}

skill_distribution_wide <- skills_long |>
  filter(skill %in% modeled_skills) |>
  count(dataset, skill, name = "skill_frequency") |>
  complete(dataset, skill = modeled_skills, fill = list(skill_frequency = 0)) |>
  group_by(dataset) |>
  mutate(skill_probability = skill_frequency / sum(skill_frequency)) |>
  ungroup() |>
  select(dataset, skill, skill_probability) |>
  pivot_wider(names_from = dataset, values_from = skill_probability, values_fill = 0)

training_col <- "5k training dataset"
external_col <- "KSB alumni external validation"

skill_distribution_shift_summary <- tibble(
  compared_label_space = "Modeled skills only",
  jensen_shannon_divergence = js_divergence(
    skill_distribution_wide[[training_col]],
    skill_distribution_wide[[external_col]]
  )
)

skill_distribution_shift_summary

write_csv(
  skill_distribution_shift_summary,
  file.path(eda_dir, "skill_distribution_shift_js_divergence.csv")
)
```

```{r skill-frequency-comparison, fig.width=10, fig.height=7}
skill_frequency_comparison <- skills_long |>
  filter(skill %in% modeled_skills) |>
  count(dataset, skill, name = "skill_frequency") |>
  group_by(dataset) |>
  mutate(skill_rate = skill_frequency / sum(skill_frequency)) |>
  ungroup() |>
  select(dataset, skill, skill_rate) |>
  pivot_wider(names_from = dataset, values_from = skill_rate, values_fill = 0) |>
  mutate(abs_rate_difference = abs(.data[[training_col]] - .data[[external_col]])) |>
  arrange(desc(abs_rate_difference))

skill_frequency_comparison |>
  slice_head(n = 20)

write_csv(
  skill_frequency_comparison,
  file.path(eda_dir, "skill_frequency_train_external_comparison.csv")
)
```

```{r skill-frequency-comparison-plot, fig.width=12, fig.height=7}
top_shifted_skills <- skill_frequency_comparison |>
  slice_head(n = 15) |>
  select(skill, all_of(training_col), all_of(external_col)) |>
  pivot_longer(cols = -skill, names_to = "dataset", values_to = "skill_rate")

skill_frequency_comparison_plot <- top_shifted_skills |>
  mutate(skill = fct_reorder(skill, skill_rate, .fun = max)) |>
  ggplot(aes(x = skill, y = skill_rate, fill = dataset)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.7) +
  coord_flip() +
  scale_y_continuous(labels = label_percent()) +
  labs(
    title = "Modeled Skills with Largest Training vs External Frequency Differences",
    subtitle = "Large differences can explain weaker external validation performance",
    x = NULL,
    y = "Share of modeled-skill assignments",
    fill = NULL
  ) +
  theme_minimal()

skill_frequency_comparison_plot

ggsave(
  file.path(eda_dir, "skill_frequency_train_external_comparison.png"),
  skill_frequency_comparison_plot,
  width = 12,
  height = 7
)
```

# Model Readiness and Risk Diagnostics

**Non-technical explanation.** This section translates the EDA into modeling
decisions. It explains what the dataset makes easy, what it makes difficult, and
what must be handled carefully before reporting final model results.

**Technical explanation.** The main risks are label imbalance, sparse TF-IDF
features, distribution shift, and incomplete external label-space coverage.
These risks affect model selection, threshold tuning, and evaluation
interpretation.

```{r model-readiness-risk-table}
model_readiness_risks <- tibble::tribble(
  ~risk_area, ~plain_language_meaning, ~technical_meaning, ~recommended_action,
  "Sparse text features",
  "Each profile uses only a small portion of the full vocabulary.",
  "The TF-IDF matrix is high-dimensional and mostly zeros.",
  "Use regularized Logistic Regression as a strong sparse-text baseline; interpret KNN cautiously.",
  "Skill-label imbalance",
  "Some skills appear often while many appear rarely.",
  "Macro metrics will be lower than micro metrics because rare labels are harder to learn.",
  "Report micro-F1, macro-F1, precision, recall, and per-skill performance for top modeled skills.",
  "Zero-skill profiles",
  "Some profiles may not contain any modeled skills.",
  "All-zero target rows provide negative examples and protect against overprediction.",
  "Keep zero-skill profiles unless they are confirmed data-quality errors.",
  "External validation drift",
  "The alumni data may not look exactly like the 5k training data.",
  "Different text length or skill distributions can lower out-of-sample performance.",
  "Explain external validation results separately from internal test results.",
  "Limited modeled label space",
  "The model can only predict skills included in the retained skill list.",
  "External labels outside the modeled skill space are out-of-vocabulary targets.",
  "Report coverage of external labels inside the 100-skill model space before judging performance."
)

model_readiness_risks

write_csv(model_readiness_risks, file.path(eda_dir, "model_readiness_risk_table.csv"))
```

**Modeling recommendation.** Based on this EDA, the final modeling workflow
should keep Logistic Regression, KNN, and XGBoost, but the evaluation should not
rely on accuracy. The correct evaluation frame is multi-label performance: micro
precision, micro recall, micro F1, macro precision, macro recall, macro F1, and
per-skill diagnostics. Threshold tuning is necessary because the default
probability cutoff may not balance precision and recall well for rare skills.

# Key Takeaways

```{r key-takeaways}
takeaways <- tibble::tribble(
  ~finding, ~why_it_matters_for_skill_extraction,
  "Skill labels are multi-label and highly imbalanced.",
  "F1-micro is useful, but common skills can dominate model performance. Per-skill review is important.",
  "Job descriptions and titles provide the model's main text signal.",
  "Profiles without meaningful descriptions are harder to classify because the model has less evidence.",
  "The external alumni dataset is smaller but directly tests out-of-sample generalization.",
  "This is the best check of whether the training workflow transfers to KSB alumni profiles.",
  "Explicit skill mentions are only part of the problem.",
  "When skill names are not directly in the text, models must infer skills from job duties, titles, industries, and context.",
  "The 100-skill model space should be checked against external labels.",
  "If many external skills are outside the modeled label space, the model cannot predict them regardless of text quality."
)

takeaways

write_csv(takeaways, file.path(eda_dir, "eda_key_takeaways.csv"))
```
