---
title: "CP8 Data Preparation"
author: "Lincoln Chanakira"
output: html_document
---

## File Overview

This notebook starts the 5k LinkedIn modeling workflow. It imports the raw Excel scrape, checks that the data loaded correctly, removes fields that are not needed for modeling, and writes a prepared CSV for the cleaning step.

This notebook imports the raw 5k LinkedIn scrape, removes fields that are not needed for modeling, and saves a smaller prepared CSV for the next preprocessing step.

The setup chunk controls how code, warnings, and messages appear in the rendered HTML output.

**Summary:** This chunk sets global R Markdown options so rendered reports show code while suppressing warnings and messages. It also loads any packages needed immediately by this notebook.

```{r setup, include=FALSE}
# Set global knitting options for cleaner rendered output.
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Load Packages

This block loads the packages used to read the Excel source file, write CSV output, and select or inspect columns.

**Summary:** This chunk loads the R packages needed in this file. Loading them at the top makes the dependencies explicit before data processing begins.

```{r packages}
# Load packages for importing Excel data and basic data manipulation.
library(readr)
library(readxl)
library(dplyr)
```

# Import Dataset

This block reads the raw LinkedIn scrape from the Excel workbook. It imports the `Scraped data` sheet and prints the dataset dimensions so the row and column counts can be checked immediately.

**Summary:** This chunk imports the raw 5k LinkedIn scrape from Excel. The dimension printout confirms that the expected source data was loaded before any cleaning happens.

```{r import-5k-data}
# Read the raw 5k LinkedIn scrape from the source workbook.
linkedin_full_path <- "../../../data/raw/LinkedIn_Scrape_full_5865.xlsx"

linkedin_full <- read_excel(linkedin_full_path, sheet = "Scraped data")

cat("LinkedIn full shape:", dim(linkedin_full), "\n")
```

# Inspect Imported Data

This block previews the imported data structure. `glimpse()` shows column names and data types, while `head()` shows the first few rows.

**Summary:** This chunk previews the raw dataset structure and first rows. It is used to verify column names, data types, and obvious data issues before dropping fields.

```{r inspect-5k-data}
# Inspect the raw data before removing unused columns.
glimpse(linkedin_full)

head(linkedin_full)
```

# Remove Columns and Save Prepared Dataset

This block drops identifying, date, company-detail, and redundant employment columns that are not used by the modeling workflow. It then saves the prepared version as a CSV for preprocessing.

**Summary:** This chunk removes columns that are not used in the modeling workflow and saves the reduced dataset. This creates a cleaner prepared file for the preprocessing notebook.

```{r remove-columns-save}
# Remove columns that are not needed downstream and write the prepared dataset.
columns_to_remove <- c(
  "full_name",
  "email",
  "education_start_year",
  "education_end_year",
  "current_company_url",
  "Company_size",
  "current_job_start_date",
  "last_scraped_date",
  "previous_job_start_date",
  "previous_job_end_date",
  "previous_company_size",
  "previous_company_name",
  "employment_type",
  "is_currently_employed",
  "current_job_duration_years"
)

linkedin_prepared <- linkedin_full |>
  select(-any_of(columns_to_remove))

prepared_output_path <- "../../../data/processed/linkedin_full_5865_prepared.csv"
write_csv(linkedin_prepared, prepared_output_path)

cat("Prepared dataset shape:", dim(linkedin_prepared), "\n")
cat("Prepared dataset saved to:", prepared_output_path, "\n")
glimpse(linkedin_prepared)
```
