---
title: "CP8 Data Preparation - 500 Profile Dataset"
author: "Lincoln Chanakira"
output: html_document
---

This notebook loads the raw 500-profile LinkedIn experience and skills datasets,
creates one model-ready row per profile, and saves the prepared data for
downstream preprocessing and modeling.

# What This File Does

- Loads the raw 500 experience and skills CSV files.
- Cleans column names so they are easier to use in R.
- Combines repeated job columns into profile-level variables.
- Builds `profile_text` from job titles, descriptions, companies, and industries.
- Groups each profile's skills into `skills_list`.
- Joins profile text with skill labels to create `model_ready`.
- Saves `model_ready_data.rds` and `model_ready_data.csv`.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Load Packages

```{r packages}
library(readr)
library(dplyr)
library(stringr)
library(purrr)
library(tidyr)
```

# Set File Paths

```{r file-paths}
project_root <- normalizePath(file.path("..", "..", ".."))
raw_dir <- file.path(project_root, "data", "raw")
processed_dir <- file.path(project_root, "data", "processed")

dir.create(processed_dir, recursive = TRUE, showWarnings = FALSE)

exp_path <- file.path(raw_dir, "LinkedIn_Scrape_500_Experiences_Wide_Full_List.csv")
skills_path <- file.path(raw_dir, "LinkedIn_Scrape_500_Skills_List.csv")

prepared_csv_path <- file.path(processed_dir, "linkedin_500_prepared.csv")
prepared_rds_path <- file.path(processed_dir, "linkedin_500_prepared.rds")
model_ready_csv_path <- file.path(processed_dir, "model_ready_data.csv")
model_ready_rds_path <- file.path(processed_dir, "model_ready_data.rds")
```

# Import Raw Data

```{r import-raw-data}
clean_column_names <- function(data) {
  names(data) <- names(data) |>
    str_replace_all("([a-z0-9])([A-Z])", "\\1_\\2") |>
    str_replace_all("[^A-Za-z0-9]+", "_") |>
    str_replace_all("_+", "_") |>
    str_replace_all("^_|_$", "") |>
    str_to_lower()

  data
}

exp_raw <- read_csv(exp_path, show_col_types = FALSE) |>
  clean_column_names()

skills_raw <- read_csv(skills_path, show_col_types = FALSE) |>
  clean_column_names()

cat("Raw experience rows and columns:", dim(exp_raw), "\n")
cat("Raw skills rows and columns:", dim(skills_raw), "\n")

glimpse(exp_raw)
glimpse(skills_raw)
```

# Helper Functions

```{r helper-functions}
paste_nonempty <- function(...) {
  vals <- c(...)
  vals <- vals[!is.na(vals)]
  vals <- str_squish(as.character(vals))
  vals <- vals[vals != ""]

  if (length(vals) == 0) {
    return(NA_character_)
  }

  paste(vals, collapse = " ")
}

paste_columns <- function(data, columns) {
  if (length(columns) == 0) {
    return(rep(NA_character_, nrow(data)))
  }

  pmap_chr(select(data, all_of(columns)), paste_nonempty)
}
```

# Identify Wide Experience Columns

```{r identify-wide-columns}
title_cols <- names(exp_raw)[str_detect(names(exp_raw), "^title_\\d+$")]
desc_cols <- names(exp_raw)[str_detect(names(exp_raw), "^job_description_\\d+$")]
comp_cols <- names(exp_raw)[str_detect(names(exp_raw), "^company_name_\\d+$")]
ind_cols <- names(exp_raw)[str_detect(names(exp_raw), "^company_industry_\\d+$")]
ctry_cols <- names(exp_raw)[str_detect(names(exp_raw), "^job_location_country_\\d+$")]

previous_title_cols <- setdiff(title_cols, "title_1")
previous_desc_cols <- setdiff(desc_cols, "job_description_1")
previous_ind_cols <- setdiff(ind_cols, "company_industry_1")
previous_ctry_cols <- setdiff(ctry_cols, "job_location_country_1")

cat("Title columns found:", length(title_cols), "\n")
cat("Description columns found:", length(desc_cols), "\n")
cat("Company columns found:", length(comp_cols), "\n")
cat("Industry columns found:", length(ind_cols), "\n")
cat("Country columns found:", length(ctry_cols), "\n")
```

# Prepare Profile-Level Experience Data

```{r prepare-experience-data}
exp_prepared <- exp_raw |>
  mutate(
    current_job_title = title_1,
    current_job_description = job_description_1,
    current_company_name = company_name_1,
    current_job_industry = company_industry_1,
    current_job_country = job_location_country_1,
    previous_job_title = paste_columns(cur_data(), previous_title_cols),
    previous_job_description = paste_columns(cur_data(), previous_desc_cols),
    previous_job_country = paste_columns(cur_data(), previous_ctry_cols),
    previous_job_industry = paste_columns(cur_data(), previous_ind_cols),
    all_titles = paste_columns(cur_data(), title_cols),
    all_descriptions = paste_columns(cur_data(), desc_cols),
    all_companies = paste_columns(cur_data(), comp_cols),
    all_industries = paste_columns(cur_data(), ind_cols),
    total_experience_count = rowSums(
      !is.na(select(cur_data(), all_of(title_cols))) &
        select(cur_data(), all_of(title_cols)) != ""
    ),
    has_current_job_description = if_else(
      !is.na(current_job_description) & str_squish(current_job_description) != "",
      1L,
      0L
    ),
    has_previous_job_description = if_else(
      !is.na(previous_job_description) & str_squish(previous_job_description) != "",
      1L,
      0L
    ),
    profile_text = pmap_chr(
      list(all_titles, all_descriptions, all_companies, all_industries),
      paste_nonempty
    )
  ) |>
  transmute(
    linkedin_url,
    current_job_title,
    current_job_description,
    current_company_name,
    current_job_industry,
    current_job_country,
    previous_job_title,
    previous_job_description,
    previous_job_country,
    previous_job_industry,
    total_experience_count,
    has_current_job_description,
    has_previous_job_description,
    profile_text
  )

cat("Prepared experience rows and columns:", dim(exp_prepared), "\n")
glimpse(exp_prepared)
```

# Prepare Skills Ground Truth

```{r prepare-skills-data}
skills_prepared <- skills_raw |>
  mutate(
    linkedin_url = str_squish(linkedin_url),
    skill = str_squish(str_to_lower(skill))
  ) |>
  filter(!is.na(linkedin_url), linkedin_url != "", !is.na(skill), skill != "") |>
  distinct(linkedin_url, skill) |>
  group_by(linkedin_url) |>
  summarise(
    skills_list = list(sort(unique(skill))),
    skills = paste(sort(unique(skill)), collapse = ", "),
    skill_count = n_distinct(skill),
    .groups = "drop"
  )

cat("Prepared skills rows and columns:", dim(skills_prepared), "\n")
glimpse(skills_prepared)
```

# Merge Experience and Skills Data

```{r merge-profile-and-skills-data}
model_ready <- exp_prepared |>
  left_join(skills_prepared, by = "linkedin_url") |>
  mutate(
    skills_list = map(skills_list, ~ if (is.null(.x) || length(.x) == 0) character(0) else .x),
    skills = replace_na(skills, ""),
    skill_count = replace_na(skill_count, 0L),
    has_any_skill = if_else(skill_count > 0, 1L, 0L)
  ) |>
  filter(!is.na(profile_text), str_squish(profile_text) != "") |>
  distinct(linkedin_url, .keep_all = TRUE)

cat("Model-ready rows and columns:", dim(model_ready), "\n")
glimpse(model_ready)
```

# Quality Checks

```{r quality-checks}
model_ready |>
  summarise(
    total_profiles = n(),
    profiles_with_skills = sum(has_any_skill),
    avg_skill_count = mean(skill_count),
    median_skill_count = median(skill_count),
    avg_experience_count = mean(total_experience_count)
  )

model_ready |>
  count(skill_count, sort = TRUE) |>
  head(10)
```

# Save Prepared Dataset

```{r save-prepared-data}
write_rds(model_ready, prepared_rds_path)
write_csv(model_ready |> select(-skills_list), prepared_csv_path)

write_rds(model_ready, model_ready_rds_path)
write_csv(model_ready |> select(-skills_list), model_ready_csv_path)

cat("Prepared CSV saved to:", prepared_csv_path, "\n")
cat("Prepared RDS saved to:", prepared_rds_path, "\n")
cat("Model-ready CSV saved to:", model_ready_csv_path, "\n")
cat("Model-ready RDS saved to:", model_ready_rds_path, "\n")
```
