---
title: "5k Model: Baseline Frequency Prior"
author: "Team"
output: html_document
---

## File Overview

This notebook trains the simplest 5k model: a frequency-prior baseline. It predicts each skill using that skill's training-set frequency, tunes decision thresholds on validation data, evaluates the baseline, and saves the results.

This notebook trains and evaluates the baseline frequency-prior model for the
5k LinkedIn skills extraction dataset. The model predicts each skill using its
training-set prevalence.

The setup chunk controls output formatting and loads the packages used for table manipulation and metric reporting.

**Summary:** This chunk sets global R Markdown options so rendered reports show code while suppressing warnings and messages. It also loads any packages needed immediately by this notebook.

```{r setup, include=FALSE}
# Set notebook output options and load required packages.
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(dplyr)
library(tibble)
```

# Load Encoded 5k Inputs

This block loads the saved train, validation, and test split objects from the encoding step. It extracts TF-IDF feature matrices, target label matrices, and retained skill names, then prints dimensions for a quick shape check.

**Summary:** This chunk loads the encoded train, validation, and test objects produced by feature encoding. It extracts input matrices, target matrices, and skill labels for the model notebook.

```{r load-data}
# Load encoded feature matrices and multi-label targets.
project_root <- normalizePath(file.path("..", "..", "..", ".."))
split_input_path <- file.path(project_root, "data", "processed", "train_validation_test_split.rds")

split_objects <- readRDS(split_input_path)

X_train <- split_objects$X_train
X_validation <- split_objects$X_validation
X_test <- split_objects$X_test

y_train <- as.matrix(split_objects$y_train)
y_validation <- as.matrix(split_objects$y_validation)
y_test <- as.matrix(split_objects$y_test)

retained_skills <- split_objects$retained_skills

cat("X_train dimensions:", dim(X_train), "\n")
cat("X_validation dimensions:", dim(X_validation), "\n")
cat("X_test dimensions:", dim(X_test), "\n")
cat("y_train dimensions:", dim(y_train), "\n")
cat("y_validation dimensions:", dim(y_validation), "\n")
cat("y_test dimensions:", dim(y_test), "\n")
```

# Modeling Scope

This block limits the model to the 100 most common skill labels. The baseline is simple, but using the same label scope as the other model notebooks keeps results comparable.

**Summary:** This chunk limits modeling to the most frequent skill labels. The cap keeps training computationally manageable and ensures all compared models use the same target set.

```{r modeling-scope}
# Select the most frequent skill labels for this model run.
max_model_labels <- 100

label_prevalence <- colMeans(y_train)
label_order <- order(label_prevalence, decreasing = TRUE)
selected_label_indices <- label_order[seq_len(min(max_model_labels, length(label_order)))]

y_train <- y_train[, selected_label_indices, drop = FALSE]
y_validation <- y_validation[, selected_label_indices, drop = FALSE]
y_test <- y_test[, selected_label_indices, drop = FALSE]
retained_skills <- colnames(y_train)

cat("Modeled labels:", ncol(y_train), "\n")
print(retained_skills)
```

# Helper Functions

This block defines reusable helpers for multi-label evaluation, probability thresholding, and validation-set threshold tuning. These functions convert probability scores into binary skill predictions and calculate micro/macro precision, recall, and F1.

**Summary:** This chunk defines reusable helper functions for multi-label evaluation, probability thresholding, class checks, and validation-set threshold tuning. These helpers keep scoring consistent across skill labels.

```{r helper-functions}
# Define metric, prediction, and threshold tuning helpers.
safe_divide <- function(num, den) {
  ifelse(den == 0, 0, num / den)
}

evaluate_multilabel <- function(y_true, y_pred_binary) {
  y_true <- as.matrix(y_true)
  y_pred_binary <- as.matrix(y_pred_binary)

  tp_micro <- sum((y_true == 1) & (y_pred_binary == 1))
  fp_micro <- sum((y_true == 0) & (y_pred_binary == 1))
  fn_micro <- sum((y_true == 1) & (y_pred_binary == 0))

  precision_micro <- safe_divide(tp_micro, tp_micro + fp_micro)
  recall_micro <- safe_divide(tp_micro, tp_micro + fn_micro)
  f1_micro <- safe_divide(2 * precision_micro * recall_micro, precision_micro + recall_micro)

  precision_by_label <- numeric(ncol(y_true))
  recall_by_label <- numeric(ncol(y_true))
  f1_by_label <- numeric(ncol(y_true))

  for (j in seq_len(ncol(y_true))) {
    tp <- sum((y_true[, j] == 1) & (y_pred_binary[, j] == 1))
    fp <- sum((y_true[, j] == 0) & (y_pred_binary[, j] == 1))
    fn <- sum((y_true[, j] == 1) & (y_pred_binary[, j] == 0))

    p <- safe_divide(tp, tp + fp)
    r <- safe_divide(tp, tp + fn)
    f <- safe_divide(2 * p * r, p + r)

    precision_by_label[j] <- p
    recall_by_label[j] <- r
    f1_by_label[j] <- f
  }

  tibble(
    precision_micro = precision_micro,
    recall_micro = recall_micro,
    f1_micro = f1_micro,
    precision_macro = mean(precision_by_label, na.rm = TRUE),
    recall_macro = mean(recall_by_label, na.rm = TRUE),
    f1_macro = mean(f1_by_label, na.rm = TRUE)
  )
}

predict_binary_from_probs <- function(prob_matrix, threshold = 0.50) {
  prob_matrix <- as.matrix(prob_matrix)

  if (length(threshold) == 1) {
    return(ifelse(prob_matrix >= threshold, 1L, 0L))
  }

  threshold_matrix <- matrix(
    rep(threshold, each = nrow(prob_matrix)),
    nrow = nrow(prob_matrix),
    byrow = FALSE
  )

  ifelse(prob_matrix >= threshold_matrix, 1L, 0L)
}

tune_label_thresholds <- function(y_true, prob_matrix, grid = seq(0.05, 0.95, by = 0.05)) {
  y_true <- as.matrix(y_true)
  prob_matrix <- as.matrix(prob_matrix)
  thresholds <- rep(0.50, ncol(y_true))
  names(thresholds) <- colnames(y_true)

  for (j in seq_len(ncol(y_true))) {
    if (sum(y_true[, j] == 1) == 0) {
      next
    }

    scores <- sapply(grid, function(threshold) {
      pred <- ifelse(prob_matrix[, j] >= threshold, 1L, 0L)
      tp <- sum((y_true[, j] == 1) & (pred == 1))
      fp <- sum((y_true[, j] == 0) & (pred == 1))
      fn <- sum((y_true[, j] == 1) & (pred == 0))
      precision <- safe_divide(tp, tp + fp)
      recall <- safe_divide(tp, tp + fn)
      safe_divide(2 * precision * recall, precision + recall)
    })

    thresholds[j] <- grid[which.max(scores)]
  }

  thresholds
}
```

# Train Baseline

This block builds the baseline predictions. It uses each skill's training-set prevalence as the predicted probability for every validation and test profile, tunes one threshold per skill on validation data, and evaluates the resulting binary predictions.

**Summary:** This chunk trains the frequency-prior baseline. It predicts each skill using training-set prevalence, tunes thresholds on validation data, and evaluates validation and test performance.

```{r train-baseline}
# Predict each skill with its training-set frequency and evaluate the baseline.
skill_prevalence <- colMeans(y_train)

baseline_probs_validation <- matrix(
  rep(skill_prevalence, each = nrow(y_validation)),
  nrow = nrow(y_validation),
  byrow = FALSE
)

baseline_probs_test <- matrix(
  rep(skill_prevalence, each = nrow(y_test)),
  nrow = nrow(y_test),
  byrow = FALSE
)

colnames(baseline_probs_validation) <- colnames(y_train)
colnames(baseline_probs_test) <- colnames(y_train)

baseline_thresholds <- tune_label_thresholds(y_validation, baseline_probs_validation)

baseline_pred_validation <- predict_binary_from_probs(
  baseline_probs_validation,
  threshold = baseline_thresholds
)

baseline_pred_test <- predict_binary_from_probs(
  baseline_probs_test,
  threshold = baseline_thresholds
)

baseline_validation_metrics <- evaluate_multilabel(y_validation, baseline_pred_validation)
baseline_test_metrics <- evaluate_multilabel(y_test, baseline_pred_test)

baseline_validation_metrics
baseline_test_metrics
```

# Save Outputs

This block saves the baseline metrics, probabilities, predictions, thresholds, and retained skill labels so they can be reused in the model comparison summary.

**Summary:** This chunk saves this model's outputs, including metrics, thresholds, predictions, probabilities, retained labels, and fitted artifacts where applicable. Saving these files supports later comparison and reporting.

```{r save-outputs}
# Write baseline artifacts and metric CSV files to the model output folder.
model_output_dir <- file.path(project_root, "data", "processed", "5k_model_outputs", "baseline_frequency_prior")
dir.create(model_output_dir, recursive = TRUE, showWarnings = FALSE)

baseline_results <- list(
  model = "Baseline Frequency Prior",
  validation_metrics = baseline_validation_metrics,
  test_metrics = baseline_test_metrics,
  thresholds = baseline_thresholds,
  validation_probabilities = baseline_probs_validation,
  test_probabilities = baseline_probs_test,
  validation_predictions = baseline_pred_validation,
  test_predictions = baseline_pred_test,
  retained_skills = retained_skills
)

saveRDS(baseline_results, file.path(model_output_dir, "baseline_frequency_prior_5k_results.rds"))
write.csv(baseline_validation_metrics, file.path(model_output_dir, "validation_metrics.csv"), row.names = FALSE)
write.csv(baseline_test_metrics, file.path(model_output_dir, "test_metrics.csv"), row.names = FALSE)

cat("Baseline outputs saved to:", model_output_dir, "\n")
```
