---
title: "5k Model: One-vs-Rest KNN"
author: "Team"
output: html_document
---

## File Overview

This notebook trains a one-vs-rest KNN model for the 5k skill prediction task. It reduces and scales the TF-IDF feature space for KNN, predicts skill labels based on nearby profiles, tunes thresholds, evaluates performance, and saves outputs.

This notebook trains and evaluates the one-vs-rest KNN model for the 5k
LinkedIn skills extraction dataset.

The setup chunk controls output formatting and loads the packages used for table manipulation, sparse matrices, and KNN classification.

**Summary:** This chunk sets global R Markdown options so rendered reports show code while suppressing warnings and messages. It also loads any packages needed immediately by this notebook.

```{r setup, include=FALSE}
# Set notebook output options and load required packages.
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(dplyr)
library(tibble)
library(Matrix)
library(class)
```

# Load Encoded 5k Inputs

This block loads the encoded train, validation, and test objects from the previous step. It extracts TF-IDF feature matrices, multi-label target matrices, and retained skill labels, then prints dimensions to verify the split sizes.

**Summary:** This chunk loads the encoded train, validation, and test objects produced by feature encoding. It extracts input matrices, target matrices, and skill labels for the model notebook.

```{r load-data}
# Load encoded feature matrices and multi-label targets.
project_root <- normalizePath(file.path("..", "..", "..", ".."))
split_input_path <- file.path(project_root, "data", "processed", "train_validation_test_split.rds")

split_objects <- readRDS(split_input_path)

X_train <- split_objects$X_train
X_validation <- split_objects$X_validation
X_test <- split_objects$X_test

y_train <- as.matrix(split_objects$y_train)
y_validation <- as.matrix(split_objects$y_validation)
y_test <- as.matrix(split_objects$y_test)

retained_skills <- split_objects$retained_skills

cat("X_train dimensions:", dim(X_train), "\n")
cat("X_validation dimensions:", dim(X_validation), "\n")
cat("X_test dimensions:", dim(X_test), "\n")
cat("y_train dimensions:", dim(y_train), "\n")
cat("y_validation dimensions:", dim(y_validation), "\n")
cat("y_test dimensions:", dim(y_test), "\n")
```

# Modeling Scope

This block limits the model to the 100 most common skills. KNN becomes expensive on large high-dimensional label problems, so this scope keeps the run manageable and comparable with the other standalone model files.

**Summary:** This chunk limits modeling to the most frequent skill labels. The cap keeps training computationally manageable and ensures all compared models use the same target set.

```{r modeling-scope}
# Select the most frequent skill labels for this model run.
max_model_labels <- 100

label_prevalence <- colMeans(y_train)
label_order <- order(label_prevalence, decreasing = TRUE)
selected_label_indices <- label_order[seq_len(min(max_model_labels, length(label_order)))]

y_train <- y_train[, selected_label_indices, drop = FALSE]
y_validation <- y_validation[, selected_label_indices, drop = FALSE]
y_test <- y_test[, selected_label_indices, drop = FALSE]
retained_skills <- colnames(y_train)

cat("Modeled labels:", ncol(y_train), "\n")
print(retained_skills)
```

# Helper Functions

This block defines reusable helpers for multi-label evaluation, probability thresholding, checking whether a label has both classes, and tuning skill-specific thresholds on validation data.

**Summary:** This chunk defines reusable helper functions for multi-label evaluation, probability thresholding, class checks, and validation-set threshold tuning. These helpers keep scoring consistent across skill labels.

```{r helper-functions}
# Define metric, prediction, class-check, and threshold tuning helpers.
safe_divide <- function(num, den) {
  ifelse(den == 0, 0, num / den)
}

evaluate_multilabel <- function(y_true, y_pred_binary) {
  y_true <- as.matrix(y_true)
  y_pred_binary <- as.matrix(y_pred_binary)

  tp_micro <- sum((y_true == 1) & (y_pred_binary == 1))
  fp_micro <- sum((y_true == 0) & (y_pred_binary == 1))
  fn_micro <- sum((y_true == 1) & (y_pred_binary == 0))

  precision_micro <- safe_divide(tp_micro, tp_micro + fp_micro)
  recall_micro <- safe_divide(tp_micro, tp_micro + fn_micro)
  f1_micro <- safe_divide(2 * precision_micro * recall_micro, precision_micro + recall_micro)

  precision_by_label <- numeric(ncol(y_true))
  recall_by_label <- numeric(ncol(y_true))
  f1_by_label <- numeric(ncol(y_true))

  for (j in seq_len(ncol(y_true))) {
    tp <- sum((y_true[, j] == 1) & (y_pred_binary[, j] == 1))
    fp <- sum((y_true[, j] == 0) & (y_pred_binary[, j] == 1))
    fn <- sum((y_true[, j] == 1) & (y_pred_binary[, j] == 0))

    p <- safe_divide(tp, tp + fp)
    r <- safe_divide(tp, tp + fn)
    f <- safe_divide(2 * p * r, p + r)

    precision_by_label[j] <- p
    recall_by_label[j] <- r
    f1_by_label[j] <- f
  }

  tibble(
    precision_micro = precision_micro,
    recall_micro = recall_micro,
    f1_micro = f1_micro,
    precision_macro = mean(precision_by_label, na.rm = TRUE),
    recall_macro = mean(recall_by_label, na.rm = TRUE),
    f1_macro = mean(f1_by_label, na.rm = TRUE)
  )
}

predict_binary_from_probs <- function(prob_matrix, threshold = 0.50) {
  prob_matrix <- as.matrix(prob_matrix)

  if (length(threshold) == 1) {
    return(ifelse(prob_matrix >= threshold, 1L, 0L))
  }

  threshold_matrix <- matrix(
    rep(threshold, each = nrow(prob_matrix)),
    nrow = nrow(prob_matrix),
    byrow = FALSE
  )

  ifelse(prob_matrix >= threshold_matrix, 1L, 0L)
}

has_two_classes <- function(y) {
  length(unique(y)) == 2
}

tune_label_thresholds <- function(y_true, prob_matrix, grid = seq(0.05, 0.95, by = 0.05)) {
  y_true <- as.matrix(y_true)
  prob_matrix <- as.matrix(prob_matrix)
  thresholds <- rep(0.50, ncol(y_true))
  names(thresholds) <- colnames(y_true)

  for (j in seq_len(ncol(y_true))) {
    if (sum(y_true[, j] == 1) == 0) {
      next
    }

    scores <- sapply(grid, function(threshold) {
      pred <- ifelse(prob_matrix[, j] >= threshold, 1L, 0L)
      tp <- sum((y_true[, j] == 1) & (pred == 1))
      fp <- sum((y_true[, j] == 0) & (pred == 1))
      fn <- sum((y_true[, j] == 1) & (pred == 0))
      precision <- safe_divide(tp, tp + fp)
      recall <- safe_divide(tp, tp + fn)
      safe_divide(2 * precision * recall, precision + recall)
    })

    thresholds[j] <- grid[which.max(scores)]
  }

  thresholds
}
```

# Train KNN

This block trains the KNN-style prediction workflow. It selects the most active TF-IDF features, converts them to dense matrices, scales validation and test data using training statistics, predicts each skill with `k = 5`, tunes thresholds, and evaluates validation/test performance.

**Summary:** This chunk trains the KNN model using a reduced and scaled TF-IDF feature matrix. The reduction is needed because KNN is memory-intensive on high-dimensional sparse text features.

```{r train-knn}
# Build dense reduced features, run KNN per skill, and evaluate predictions.
set.seed(621)

knn_feature_count <- min(500, ncol(X_train))
feature_variance <- Matrix::colMeans(X_train != 0)
top_feature_indices <- order(feature_variance, decreasing = TRUE)[seq_len(knn_feature_count)]

X_train_dense_raw <- as.matrix(X_train[, top_feature_indices])
X_validation_dense_raw <- as.matrix(X_validation[, top_feature_indices])
X_test_dense_raw <- as.matrix(X_test[, top_feature_indices])

knn_center <- colMeans(X_train_dense_raw)
knn_scale <- apply(X_train_dense_raw, 2, sd)
knn_scale[is.na(knn_scale) | knn_scale == 0] <- 1

X_train_dense <- scale(X_train_dense_raw, center = knn_center, scale = knn_scale)
X_validation_dense <- scale(X_validation_dense_raw, center = knn_center, scale = knn_scale)
X_test_dense <- scale(X_test_dense_raw, center = knn_center, scale = knn_scale)

k_value <- 5

knn_validation_probs <- matrix(0, nrow = nrow(X_validation_dense), ncol = ncol(y_train))
knn_test_probs <- matrix(0, nrow = nrow(X_test_dense), ncol = ncol(y_train))

for (j in seq_len(ncol(y_train))) {
  y_j <- y_train[, j]

  if (has_two_classes(y_j)) {
    train_labels_factor <- factor(y_j, levels = c(0, 1))

    pred_validation <- knn(
      train = X_train_dense,
      test = X_validation_dense,
      cl = train_labels_factor,
      k = k_value,
      prob = TRUE
    )

    pred_validation_char <- as.character(pred_validation)
    pred_validation_prob <- ifelse(
      pred_validation_char == "1",
      attr(pred_validation, "prob"),
      1 - attr(pred_validation, "prob")
    )

    knn_validation_probs[, j] <- pred_validation_prob

    pred_test <- knn(
      train = X_train_dense,
      test = X_test_dense,
      cl = train_labels_factor,
      k = k_value,
      prob = TRUE
    )

    pred_test_char <- as.character(pred_test)
    pred_test_prob <- ifelse(
      pred_test_char == "1",
      attr(pred_test, "prob"),
      1 - attr(pred_test, "prob")
    )

    knn_test_probs[, j] <- pred_test_prob
  } else {
    knn_validation_probs[, j] <- mean(y_j)
    knn_test_probs[, j] <- mean(y_j)
  }
}

colnames(knn_validation_probs) <- colnames(y_train)
colnames(knn_test_probs) <- colnames(y_train)

knn_thresholds <- tune_label_thresholds(y_validation, knn_validation_probs)

knn_pred_validation <- predict_binary_from_probs(
  knn_validation_probs,
  threshold = knn_thresholds
)

knn_pred_test <- predict_binary_from_probs(
  knn_test_probs,
  threshold = knn_thresholds
)

knn_validation_metrics <- evaluate_multilabel(y_validation, knn_pred_validation)
knn_test_metrics <- evaluate_multilabel(y_test, knn_pred_test)

knn_validation_metrics
knn_test_metrics
```

# Save Outputs

This block saves the KNN settings, scaling values, selected feature indexes, probabilities, predictions, thresholds, metrics, and retained skill labels for downstream reporting.

**Summary:** This chunk saves this model's outputs, including metrics, thresholds, predictions, probabilities, retained labels, and fitted artifacts where applicable. Saving these files supports later comparison and reporting.

```{r save-outputs}
# Write KNN artifacts and metric CSV files to the output folder.
model_output_dir <- file.path(project_root, "data", "processed", "5k_model_outputs", "knn")
dir.create(model_output_dir, recursive = TRUE, showWarnings = FALSE)

knn_results <- list(
  model = "KNN",
  validation_metrics = knn_validation_metrics,
  test_metrics = knn_test_metrics,
  thresholds = knn_thresholds,
  k_value = k_value,
  top_feature_indices = top_feature_indices,
  center = knn_center,
  scale = knn_scale,
  validation_probabilities = knn_validation_probs,
  test_probabilities = knn_test_probs,
  validation_predictions = knn_pred_validation,
  test_predictions = knn_pred_test,
  retained_skills = retained_skills
)

saveRDS(knn_results, file.path(model_output_dir, "knn_5k_results.rds"))
write.csv(knn_validation_metrics, file.path(model_output_dir, "validation_metrics.csv"), row.names = FALSE)
write.csv(knn_test_metrics, file.path(model_output_dir, "test_metrics.csv"), row.names = FALSE)

cat("KNN outputs saved to:", model_output_dir, "\n")
```
