---
title: "500 Model: One-vs-Rest Logistic Regression"
author: "Team"
output: html_document
---

This notebook trains and evaluates a one-vs-rest regularized logistic regression
model for the 500-profile LinkedIn skills extraction dataset.

# Inputs and Skill Outputs

Input variables used by this model:

- `X_train`, `X_valid`, `X_test`: TF-IDF feature matrices created from `profile_text_nostop`.
- `y_train`, `y_valid`, `y_test`: binary skill-label matrices.
- `top_skills`: the 30 most frequent skills selected from the training set.

The model output is predicted skills for each profile. Each skill in `top_skills`
is predicted as:

- `1`: the model predicts the profile has the skill
- `0`: the model predicts the profile does not have the skill

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(tidyverse)
library(glmnet)
library(Matrix)
```

# Load Encoded 500 Inputs

```{r load-encoded-inputs}
project_root <- normalizePath(file.path("..", "..", "..", ".."))
processed_dir <- file.path(project_root, "data", "processed")
model_output_dir <- file.path(processed_dir, "500_model_outputs", "logistic_regression")

dir.create(model_output_dir, recursive = TRUE, showWarnings = FALSE)

X_train <- read_rds(file.path(processed_dir, "X_train.rds"))
X_valid <- read_rds(file.path(processed_dir, "X_valid.rds"))
X_test <- read_rds(file.path(processed_dir, "X_test.rds"))

y_train <- read_rds(file.path(processed_dir, "y_train.rds"))
y_valid <- read_rds(file.path(processed_dir, "y_valid.rds"))
y_test <- read_rds(file.path(processed_dir, "y_test.rds"))

top_skills <- read_rds(file.path(processed_dir, "top_skills.rds"))

dim(X_train)
dim(X_valid)
dim(X_test)

dim(y_train)
dim(y_valid)
dim(y_test)
```

# Helper Functions

```{r helper-functions}
safe_divide <- function(num, den) {
  ifelse(den == 0, 0, num / den)
}

has_two_classes <- function(y) {
  length(unique(y)) == 2
}

predict_binary_from_probs <- function(prob_matrix, threshold = 0.50) {
  ifelse(as.matrix(prob_matrix) >= threshold, 1L, 0L)
}

evaluate_multilabel <- function(y_true, y_pred, model_name, split_name) {
  y_true_vec <- as.vector(y_true)
  y_pred_vec <- as.vector(y_pred)

  tp <- sum(y_true_vec == 1 & y_pred_vec == 1)
  fp <- sum(y_true_vec == 0 & y_pred_vec == 1)
  fn <- sum(y_true_vec == 1 & y_pred_vec == 0)

  precision_micro <- safe_divide(tp, tp + fp)
  recall_micro <- safe_divide(tp, tp + fn)
  f1_micro <- safe_divide(2 * precision_micro * recall_micro, precision_micro + recall_micro)

  f1_by_skill <- map_dbl(seq_len(ncol(y_true)), function(j) {
    yt <- y_true[, j]
    yp <- y_pred[, j]

    tp_j <- sum(yt == 1 & yp == 1)
    fp_j <- sum(yt == 0 & yp == 1)
    fn_j <- sum(yt == 1 & yp == 0)

    precision_j <- safe_divide(tp_j, tp_j + fp_j)
    recall_j <- safe_divide(tp_j, tp_j + fn_j)
    safe_divide(2 * precision_j * recall_j, precision_j + recall_j)
  })

  tibble(
    model = model_name,
    split = split_name,
    precision_micro = precision_micro,
    recall_micro = recall_micro,
    f1_micro = f1_micro,
    macro_f1 = mean(f1_by_skill),
    subset_accuracy = mean(apply(y_true == y_pred, 1, all))
  )
}
```

# Train Logistic Regression

```{r train-logistic-regression}
set.seed(621)

logit_models <- vector("list", ncol(y_train))
names(logit_models) <- top_skills

logit_valid_probs <- matrix(0, nrow = nrow(X_valid), ncol = ncol(y_train))
logit_test_probs <- matrix(0, nrow = nrow(X_test), ncol = ncol(y_train))
colnames(logit_valid_probs) <- top_skills
colnames(logit_test_probs) <- top_skills

for (j in seq_along(top_skills)) {
  y_j <- y_train[, j]

  if (has_two_classes(y_j)) {
    model_j <- cv.glmnet(
      x = X_train,
      y = y_j,
      family = "binomial",
      alpha = 0,
      type.measure = "class",
      nfolds = 5
    )

    logit_models[[j]] <- model_j
    logit_valid_probs[, j] <- as.numeric(
      predict(model_j, newx = X_valid, s = "lambda.min", type = "response")
    )
    logit_test_probs[, j] <- as.numeric(
      predict(model_j, newx = X_test, s = "lambda.min", type = "response")
    )
  } else {
    logit_valid_probs[, j] <- mean(y_j)
    logit_test_probs[, j] <- mean(y_j)
  }
}

logit_valid_pred <- predict_binary_from_probs(logit_valid_probs, threshold = 0.50)
logit_test_pred <- predict_binary_from_probs(logit_test_probs, threshold = 0.50)

validation_metrics <- evaluate_multilabel(y_valid, logit_valid_pred, "Logistic Regression", "Validation")
test_metrics <- evaluate_multilabel(y_test, logit_test_pred, "Logistic Regression", "Test")

validation_metrics
test_metrics
```

# Save Outputs

```{r save-outputs}
logit_results <- list(
  model = "Logistic Regression",
  validation_metrics = validation_metrics,
  test_metrics = test_metrics,
  models = logit_models,
  validation_probabilities = logit_valid_probs,
  test_probabilities = logit_test_probs,
  validation_predictions = logit_valid_pred,
  test_predictions = logit_test_pred,
  top_skills = top_skills
)

saveRDS(logit_results, file.path(model_output_dir, "logistic_regression_500_results.rds"))
write_csv(validation_metrics, file.path(model_output_dir, "validation_metrics.csv"))
write_csv(test_metrics, file.path(model_output_dir, "test_metrics.csv"))

cat("Logistic regression outputs saved to:", model_output_dir, "\n")
```
