---
title: "500 Model: One-vs-Rest KNN"
author: "Team"
output: html_document
---

This notebook trains and evaluates a one-vs-rest KNN model for the 500-profile
LinkedIn skills extraction dataset.

# Inputs and Skill Outputs

Input variables used by this model:

- `X_train`, `X_valid`, `X_test`: TF-IDF feature matrices created from `profile_text_nostop`.
- `y_train`, `y_valid`, `y_test`: binary skill-label matrices.
- `top_skills`: the 30 most frequent skills selected from the training set.

The model output is predicted skills for each profile. Each skill in `top_skills`
is predicted as:

- `1`: the model predicts the profile has the skill
- `0`: the model predicts the profile does not have the skill

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(tidyverse)
library(Matrix)
library(class)
```

# Load Encoded 500 Inputs

```{r load-encoded-inputs}
project_root <- normalizePath(file.path("..", "..", "..", ".."))
processed_dir <- file.path(project_root, "data", "processed")
model_output_dir <- file.path(processed_dir, "500_model_outputs", "knn")

dir.create(model_output_dir, recursive = TRUE, showWarnings = FALSE)

X_train <- read_rds(file.path(processed_dir, "X_train.rds"))
X_valid <- read_rds(file.path(processed_dir, "X_valid.rds"))
X_test <- read_rds(file.path(processed_dir, "X_test.rds"))

y_train <- read_rds(file.path(processed_dir, "y_train.rds"))
y_valid <- read_rds(file.path(processed_dir, "y_valid.rds"))
y_test <- read_rds(file.path(processed_dir, "y_test.rds"))

top_skills <- read_rds(file.path(processed_dir, "top_skills.rds"))

dim(X_train)
dim(X_valid)
dim(X_test)

dim(y_train)
dim(y_valid)
dim(y_test)
```

# Helper Functions

```{r helper-functions}
safe_divide <- function(num, den) {
  ifelse(den == 0, 0, num / den)
}

has_two_classes <- function(y) {
  length(unique(y)) == 2
}

predict_binary_from_probs <- function(prob_matrix, threshold = 0.50) {
  ifelse(as.matrix(prob_matrix) >= threshold, 1L, 0L)
}

evaluate_multilabel <- function(y_true, y_pred, model_name, split_name) {
  y_true_vec <- as.vector(y_true)
  y_pred_vec <- as.vector(y_pred)

  tp <- sum(y_true_vec == 1 & y_pred_vec == 1)
  fp <- sum(y_true_vec == 0 & y_pred_vec == 1)
  fn <- sum(y_true_vec == 1 & y_pred_vec == 0)

  precision_micro <- safe_divide(tp, tp + fp)
  recall_micro <- safe_divide(tp, tp + fn)
  f1_micro <- safe_divide(2 * precision_micro * recall_micro, precision_micro + recall_micro)

  f1_by_skill <- map_dbl(seq_len(ncol(y_true)), function(j) {
    yt <- y_true[, j]
    yp <- y_pred[, j]

    tp_j <- sum(yt == 1 & yp == 1)
    fp_j <- sum(yt == 0 & yp == 1)
    fn_j <- sum(yt == 1 & yp == 0)

    precision_j <- safe_divide(tp_j, tp_j + fp_j)
    recall_j <- safe_divide(tp_j, tp_j + fn_j)
    safe_divide(2 * precision_j * recall_j, precision_j + recall_j)
  })

  tibble(
    model = model_name,
    split = split_name,
    precision_micro = precision_micro,
    recall_micro = recall_micro,
    f1_micro = f1_micro,
    macro_f1 = mean(f1_by_skill),
    subset_accuracy = mean(apply(y_true == y_pred, 1, all))
  )
}
```

# Train KNN

```{r train-knn}
set.seed(621)

knn_feature_count <- min(500, ncol(X_train))
feature_variance <- Matrix::colMeans(X_train != 0)
top_feature_indices <- order(feature_variance, decreasing = TRUE)[seq_len(knn_feature_count)]

X_train_dense_raw <- as.matrix(X_train[, top_feature_indices])
X_valid_dense_raw <- as.matrix(X_valid[, top_feature_indices])
X_test_dense_raw <- as.matrix(X_test[, top_feature_indices])

knn_center <- colMeans(X_train_dense_raw)
knn_scale <- apply(X_train_dense_raw, 2, sd)
knn_scale[is.na(knn_scale) | knn_scale == 0] <- 1

X_train_dense <- scale(X_train_dense_raw, center = knn_center, scale = knn_scale)
X_valid_dense <- scale(X_valid_dense_raw, center = knn_center, scale = knn_scale)
X_test_dense <- scale(X_test_dense_raw, center = knn_center, scale = knn_scale)

k_value <- 5

knn_valid_probs <- matrix(0, nrow = nrow(X_valid_dense), ncol = ncol(y_train))
knn_test_probs <- matrix(0, nrow = nrow(X_test_dense), ncol = ncol(y_train))
colnames(knn_valid_probs) <- top_skills
colnames(knn_test_probs) <- top_skills

for (j in seq_along(top_skills)) {
  y_j <- y_train[, j]

  if (has_two_classes(y_j)) {
    train_labels_factor <- factor(y_j, levels = c(0, 1))

    pred_valid <- knn(
      train = X_train_dense,
      test = X_valid_dense,
      cl = train_labels_factor,
      k = k_value,
      prob = TRUE
    )

    pred_valid_char <- as.character(pred_valid)
    knn_valid_probs[, j] <- ifelse(
      pred_valid_char == "1",
      attr(pred_valid, "prob"),
      1 - attr(pred_valid, "prob")
    )

    pred_test <- knn(
      train = X_train_dense,
      test = X_test_dense,
      cl = train_labels_factor,
      k = k_value,
      prob = TRUE
    )

    pred_test_char <- as.character(pred_test)
    knn_test_probs[, j] <- ifelse(
      pred_test_char == "1",
      attr(pred_test, "prob"),
      1 - attr(pred_test, "prob")
    )
  } else {
    knn_valid_probs[, j] <- mean(y_j)
    knn_test_probs[, j] <- mean(y_j)
  }
}

knn_valid_pred <- predict_binary_from_probs(knn_valid_probs, threshold = 0.50)
knn_test_pred <- predict_binary_from_probs(knn_test_probs, threshold = 0.50)

validation_metrics <- evaluate_multilabel(y_valid, knn_valid_pred, "KNN", "Validation")
test_metrics <- evaluate_multilabel(y_test, knn_test_pred, "KNN", "Test")

validation_metrics
test_metrics
```

# Save Outputs

```{r save-outputs}
knn_results <- list(
  model = "KNN",
  validation_metrics = validation_metrics,
  test_metrics = test_metrics,
  k_value = k_value,
  top_feature_indices = top_feature_indices,
  center = knn_center,
  scale = knn_scale,
  validation_probabilities = knn_valid_probs,
  test_probabilities = knn_test_probs,
  validation_predictions = knn_valid_pred,
  test_predictions = knn_test_pred,
  top_skills = top_skills
)

saveRDS(knn_results, file.path(model_output_dir, "knn_500_results.rds"))
write_csv(validation_metrics, file.path(model_output_dir, "validation_metrics.csv"))
write_csv(test_metrics, file.path(model_output_dir, "test_metrics.csv"))

cat("KNN outputs saved to:", model_output_dir, "\n")
```
